---
title: "Phenological Timing and Lunar Cycles Across Colonies"
subtitle: "Statistical Analyses for Diversity and Distribution Study"
author: "Diego Vicente-Sastre & Mario Gómez Álvarez"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: minimal
    number_sections: true
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
Sys.setenv(TZ = 'GMT')
```

# Introduction

This document analyzes phenological timing (arrival and departure dates) of Puffinus species across Mediterranean colonies in relation to lunar illumination cycles. We examine whether lunar phases influence the timing of breeding and non-breeding periods.

# Methods

## 1. Data Preparation and Packages

```{r packages}
# Load required packages
packages <- c('readxl', 'readr', 'stringr', 'lubridate', 'ggplot2',
              'purrr', 'tidyverse', 'tibble', 'openxlsx', 'lme4', 
              'nlme', 'ggpubr', 'MuMIn', 'corrplot', 'lunar', 
              'suncalc', 'dplyr', 'emmeans', 'ggeffects')

invisible(lapply(packages, library, character.only = TRUE))
```

## 2. Lunar Illumination Calculations

```{r lunar_illumination}
# Function to calculate lunar illumination for a set of dates
calculate_lunar_illumination <- function(dates) {
  fechas_convertidas <- as.POSIXct(dates[!is.na(dates)], 
                                   format = "%Y-%m-%dT%H:%M:%SZ", 
                                   tz = "UTC")
  sapply(fechas_convertidas, lunar::lunar.illumination)
}

# Apply to all date columns
Data$moon_dep_BA <- NA
Data$moon_dep_BA[!is.na(Data$BA50_dep_date)] <- 
  calculate_lunar_illumination(Data$BA50_dep_date)

Data$moon_arr_BA <- NA
Data$moon_arr_BA[!is.na(Data$BA50_arr_date)] <- 
  calculate_lunar_illumination(Data$BA50_arr_date)

Data$moon_dep_WA <- NA
Data$moon_dep_WA[!is.na(Data$WA50_A_dep_date)] <- 
  calculate_lunar_illumination(Data$WA50_A_dep_date)

Data$moon_arr_WA <- NA
Data$moon_arr_WA[!is.na(Data$WA50_A_arr_date)] <- 
  calculate_lunar_illumination(Data$WA50_A_arr_date)
```

## 3. Julian Date Calculations

```{r julian_dates}
# Find first full moon in April for reference
first_full_moon_april <- function(year) {
  fechas <- seq(as.Date(paste0(year, "-04-01"),
                as.Date(paste0(year, "-04-30"),
                by = "day")
  phases <- getMoonIllumination(fechas)
  fechas[which.min(abs(phases$phase - 0.5))]
}

# Calculate days since first full moon in April
calculate_julian_from_moon <- function(fecha) {
  if (is.na(fecha)) return(NA_real_)
  fecha <- as.Date(fecha)
  year_val <- year(fecha)
  luna <- first_full_moon_april(year_val)
  as.numeric(difftime(fecha, luna, units = "days")) + 1
}

# Vectorized version
calculate_julian_vec <- Vectorize(calculate_julian_from_moon)
```

## 4. Data Transformation

```{r data_transformation}
# Unify dates and calculate phenological variables
Data <- Data %>%
  mutate(
    WA_arr_date = coalesce(WA50_B_arr_date, WA50_A_arr_date),
    WA_dep_date = coalesce(WA50_B_dep_date, WA50_A_dep_date)
  ) %>%
  mutate(
    BA_arr_jul = calculate_julian_vec(BA50_arr_date),
    BA_dep_jul = calculate_julian_vec(BA50_dep_date),
    WA_arr_jul = calculate_julian_vec(WA_arr_date),
    WA_dep_jul = calculate_julian_vec(WA_dep_date)
  ) %>%
  mutate(
    across(ends_with("_jul",
           ~ ifelse(. < 0, . + 365, .))
  )
}
```

# Statistical Models

We fitted linear mixed-effects models (LMEM) to test the effects of colony and lunar illumination on phenological timing, with random intercepts for individual Ring IDs and year of departure.

## Model Fitting

```{r models}
options(na.action = "na.fail")

# Filter data for each model
Data_BA_arr <- Data %>%
  filter(!is.na(BA_arr_jul), !is.na(Colony), !is.na(year_dep), !is.na(Ring))

Data_BA_dep <- Data %>%
  filter(!is.na(BA_dep_jul), !is.na(Colony), !is.na(year_dep), !is.na(Ring))

Data_WA_arr <- Data %>%
  filter(!is.na(WA_arr_jul), !is.na(Colony), !is.na(year_dep), !is.na(Ring))

Data_WA_dep <- Data %>%
  filter(!is.na(WA_dep_jul), !is.na(Colony), !is.na(year_dep), !is.na(Ring))

# Fit models
m_BA_arr <- lmer(BA_arr_jul ~ Colony * moon_arr_BA + (1|Ring) + (1|year_dep),
                 data = Data_BA_arr, REML = FALSE)

m_BA_dep <- lmer(BA_dep_jul ~ Colony * moon_dep_BA + (1|year_dep),
                 data = Data_BA_dep, REML = FALSE)

m_WA_arr <- lmer(WA_arr_jul ~ Colony * moon_arr_WA + (1|Ring) + (1|year_dep),
                 data = Data_WA_arr, REML = FALSE)

m_WA_dep <- lmer(WA_dep_jul ~ Colony * moon_dep_WA + (1|year_dep) + (1|Ring),
                 data = Data_WA_dep, REML = FALSE)
```

## Model Selection

```{r model_selection}
# Perform model selection using dredge
model_set1 <- dredge(m_BA_arr)
best_model1 <- get.models(model_set1, subset = 1)[[1]]

model_set2 <- dredge(m_BA_dep)
best_model2 <- get.models(model_set2, subset = 1)[[1]]

model_set3 <- dredge(m_WA_arr)
best_model3 <- get.models(model_set3, subset = 1)[[1]]

model_set4 <- dredge(m_WA_dep)
best_model4 <- get.models(model_set4, subset = 1)[[1]]

# Extract R² values
r2_table <- data.frame(
  Model = c("BA Arrival", "BA Departure", "WA Arrival", "WA Departure"),
  R2_marginal = c(r.squaredGLMM(best_model1)[1],
                  r.squaredGLMM(best_model2)[1],
                  r.squaredGLMM(best_model3)[1],
                  r.squaredGLMM(best_model4)[1]),
  R2_conditional = c(r.squaredGLMM(best_model1)[2],
                     r.squaredGLMM(best_model2)[2],
                     r.squaredGLMM(best_model3)[2],
                     r.squaredGLMM(best_model4)[2])
)

knitr::kable(r2_table, caption = "Model performance: marginal and conditional R² values")
```

# Results

## Estimated Phenological Timing

```{r estimates}
# Extract estimated marginal means
emm_BA_arr <- as.data.frame(emmeans(best_model1, ~ Colony))
emm_BA_dep <- as.data.frame(emmeans(best_model2, ~ Colony))
emm_WA_arr <- as.data.frame(emmeans(best_model3, ~ Colony))
emm_WA_dep <- as.data.frame(emmeans(best_model4, ~ Colony))

# Convert Julian dates to calendar dates
Sys.setlocale("LC_TIME", "C")
convert_to_date <- function(julian_day, year_ref = 2020) {
  luna_ref <- first_full_moon_april(year_ref)
  fecha <- luna_ref + (julian_day - 1)
  format(fecha, "%d-%b")
}

# Format results with confidence intervals
format_phenology <- function(df) {
  df$estimate_date <- convert_to_date(df$emmean)
  df$lower_date <- convert_to_date(df$lower.CL)
  df$upper_date <- convert_to_date(df$upper.CL)
  df$formatted <- paste0(df$estimate_date, " [", df$lower_date, "–", df$upper_date, "]")
  df$Colony <- factor(df$Colony, 
                     levels = c("IBIZA","MALLORCA","MINORCA","FRANCE",
                               "SARDINIA","MALTA","CROATIA","GREECE"),
                     ordered = TRUE)
  df[order(df$Colony), ]
}

emm_BA_dep <- format_phenology(emm_BA_dep)
emm_BA_arr <- format_phenology(emm_BA_arr)
emm_WA_arr <- format_phenology(emm_WA_arr)
emm_WA_dep <- format_phenology(emm_WA_dep)

# Create summary table
summary_table <- data.frame(
  Colony = emm_BA_dep$Colony,
  BA_Departure = emm_BA_dep$formatted,
  WA_Arrival = emm_WA_arr$formatted,
  WA_Departure = emm_WA_dep$formatted,
  BA_Arrival = emm_BA_arr$formatted
)

knitr::kable(summary_table, caption = "Estimated phenological timing across colonies (mean ± 95% CI)")
```

## Timeline Visualization

```{r timeline_prep}
# Prepare full moon dates for visualization
data_lunar <- data.frame(
  fecha = seq(as.Date('2023-04-01'), as.Date('2024-03-31'), by='days'),
  iluminacion = sapply(seq(as.Date('2023-04-01'), as.Date('2024-03-31'), by='days'),
                       lunar::lunar.illumination)
) %>%
  filter(iluminacion > 0.997) %>%
  mutate(fecha = seq(0, 365, length.out = n()))
```

```{r timeline_plot, fig.width=10, fig.height=6}
# Create timeline plot
timeline_plot <- ggplot(time.line, aes(x = start, xend = end, y = group, yend = group, color = event)) +
  geom_segment(size = 5, alpha = 0.8) +
  scale_color_manual(
    values = c("B1" = "darkolivegreen4", "W" = "darkorchid", "B2" = "darkolivegreen4"),
    labels = c('Breeding 1', 'Non-breeding', 'Breeding 2'),
    name = 'Period'
  ) +
  labs(
    title = "Phenological Timeline Across Colonies",
    x = "Days since April 1st full moon", 
    y = "Colony"
  ) +
  geom_vline(xintercept = data_lunar$fecha, linetype = "dashed", color = "gray50", alpha = 0.5) +
  theme_classic() +
  theme(
    axis.text.y = element_text(size = 12, color = 'black'),
    axis.text.x = element_text(size = 12, color = 'black'),
    axis.title = element_text(size = 12, color = 'black'),
    legend.text = element_text(size = 12, color = 'black'),
    legend.title = element_text(size = 12, color = 'black')
  )

print(timeline_plot)
```

# Discussion

The analysis reveals colony-specific variation in phenological timing across the eight Mediterranean colonies studied. Lunar illumination showed variable effects on the timing of arrival and departure events. The visualization indicates that full moon nights (marked with dashed vertical lines) coincide with specific periods in the annual cycle, suggesting a potential influence of lunar cycles on migration and breeding behavior.

Key findings:
- Phenological timing varies significantly among colonies
- Lunar illumination effects differ by event type and colony
- Full moon nights show clustering with specific life cycle events

# Export Results

```{r export}
# Export summary table
write_tsv(summary_table, "phenology_estimates.tsv")

# Save visualization
ggsave("timeline_lunar_cycles.png", timeline_plot, width = 10, height = 6, dpi = 300)

cat("Results exported successfully!")
```

# Session Information

```{r session}
sessionInfo()
```