---
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: minimal
    number_sections: true
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
Sys.setenv(TZ = 'GMT')
```

## 1. Data Preparation and Packages

```{r packages}
# Load required packages
packages <- c('readxl', 'readr', 'stringr', 'lubridate', 'ggplot2',
              'purrr', 'tidyverse', 'tibble', 'openxlsx', 'lme4', 
              'ggpubr', 'MuMIn', 'corrplot', 'lunar','suncalc', 
              'dplyr', 'emmeans')

invisible(lapply(packages, library, character.only = TRUE))

Data <- read.csv("data/Phen_data.csv")

```

## 2. Julian Date Calculations

```{r julian_dates}
# Find first full moon in March for reference
first_full_moon_march <- function(year) {
  if (is.na(year)) return(as.Date(NA))
  dates <- seq(as.Date(paste0(year, "-03-01")),
                as.Date(paste0(year, "-03-30")),
                by = "day")
  phases <- getMoonIllumination(dates)
  dates[which.min(abs(phases$phase - 0.5))]
}

calculate_julian_vec <- function(dates) {
  # 1. Asegurar formato Date para todo el vector
  dates <- as.Date(dates)
  
  # 2. Extraer los años de todo el vector
  years <- lubridate::year(dates)
  
  # 3. Calcular la luna de abril para cada año (asumiendo que admite vectores)
  # Si first_full_moon_april no acepta vectores, usamos sapply/vapply para los años:
  moons <- as.Date(sapply(years, first_full_moon_march))
  
  # 4. Cálculo vectorizado de diferencia de días
  as.numeric(difftime(dates, moons, units = "days")) + 1
}
```

## 3. Data Transformation

```{r data_transformation}
# Unify dates and calculate phenological variables
Data <- Data %>%
  mutate(
    BA_arr_jul = calculate_julian_vec(BA_arr),
    B_dep_jul = calculate_julian_vec(B_dep),
    WA_arr_jul = calculate_julian_vec(WA_arr),
    WA_dep_jul = calculate_julian_vec(WA_dep)
  ) %>%
  mutate(
    across(
      ends_with("_jul"),
      ~ ifelse(.x < 0, .x + 365, .x)
    )
  )

```

## 3. Lunar Illumination Calculations

```{r lunar_illumination}

# Function to calculate lunar illumination for a set of dates
calculate_lunar_illumination <- function(dates) {
  dates_converted <- as.Date(dates, format = "%Y-%m-%d")
  lunar::lunar.illumination(dates_converted)
}

# Apply to all date columns
Data$moon_dep_B <- calculate_lunar_illumination(Data$B_dep)

Data$moon_arr_BA <- calculate_lunar_illumination(Data$BA_arr)

Data$moon_dep_WA <- calculate_lunar_illumination(Data$WA_dep)

Data$moon_arr_WA <- calculate_lunar_illumination(Data$WA_arr)
```

# Statistical Models

We fitted linear mixed-effects models (LMM) to test the effects of colony and lunar illumination on phenological timing, with random intercepts for individual Ring IDs and year of departure.

## Model Fitting

```{r models}
options(na.action = "na.fail")

# Filter data for each model
Data_BA_arr <- Data %>%
  filter(!is.na(BA_arr_jul), !is.na(Colony), !is.na(year_dep), !is.na(Ring))

Data_B_dep <- Data %>%
  filter(!is.na(B_dep_jul), !is.na(Colony), !is.na(year_dep), !is.na(Ring))

Data_WA_arr <- Data %>%
  filter(!is.na(WA_arr_jul), !is.na(Colony), !is.na(year_dep), !is.na(Ring))

Data_WA_dep <- Data %>%
  filter(!is.na(WA_dep_jul), !is.na(Colony), !is.na(year_dep), !is.na(Ring))

# Fit models
m_BA_arr <- lmer(BA_arr_jul ~ Colony * moon_arr_BA + (1|Ring) + (1|year_dep),
                 data = Data_BA_arr, REML = FALSE)

m_B_dep <- lmer(B_dep_jul ~ Colony * moon_dep_B + (1|year_dep),
                 data = Data_B_dep, REML = FALSE)

m_WA_arr <- lmer(WA_arr_jul ~ Colony * moon_arr_WA + (1|Ring) + (1|year_dep),
                 data = Data_WA_arr, REML = FALSE)

m_WA_dep <- lmer(WA_dep_jul ~ Colony * moon_dep_WA + (1|year_dep) + (1|Ring),
                 data = Data_WA_dep, REML = FALSE)
```

## Model Selection

```{r model_selection}
# Perform model selection using dredge
model_set1 <- dredge(m_BA_arr)
best_model1 <- get.models(model_set1, subset = 1)[[1]]

model_set2 <- dredge(m_B_dep)
best_model2 <- get.models(model_set2, subset = 1)[[1]]

model_set3 <- dredge(m_WA_arr)
best_model3 <- get.models(model_set3, subset = 1)[[1]]

model_set4 <- dredge(m_WA_dep)
best_model4 <- get.models(model_set4, subset = 1)[[1]]

# Extract R² values
r2_table <- data.frame(
  Model = c("BA Arrival", "BA Departure", "WA Arrival", "WA Departure"),
  R2_marginal = c(r.squaredGLMM(best_model1)[1],
                  r.squaredGLMM(best_model2)[1],
                  r.squaredGLMM(best_model3)[1],
                  r.squaredGLMM(best_model4)[1]),
  R2_conditional = c(r.squaredGLMM(best_model1)[2],
                     r.squaredGLMM(best_model2)[2],
                     r.squaredGLMM(best_model3)[2],
                     r.squaredGLMM(best_model4)[2])
)

knitr::kable(r2_table, caption = "Model performance: marginal and conditional R² values")
```

# Results

## Estimated Phenological Timing

```{r estimates}
# Extract estimated marginal means
emm_BA_arr <- as.data.frame(emmeans(best_model1, ~ Colony))
emm_BA_dep <- as.data.frame(emmeans(best_model2, ~ Colony))
#emm_WA_arr <- as.data.frame(emmeans(best_model3, ~ Colony)) # In our case, Colony was not selected
emm_WA_dep <- as.data.frame(emmeans(best_model4, ~ Colony))

# Convert Julian dates to calendar dates
Sys.setlocale("LC_TIME", "C")
convert_to_date <- function(julian_day, year_ref = 2020) {
  luna_ref <- first_full_moon_march(year_ref)
  fecha <- luna_ref + (julian_day - 1)
  format(fecha, "%d-%b")
}

# Format results with confidence intervals
format_phenology <- function(df) {
  df$estimate_date <- convert_to_date(df$emmean)
  df$lower_date <- convert_to_date(df$lower.CL)
  df$upper_date <- convert_to_date(df$upper.CL)
  df$formatted <- paste0(df$estimate_date, " [", df$lower_date, "–", df$upper_date, "]")
  df$Colony <- factor(df$Colony, 
                     levels = c("Ibiza","Mallorca","Minorca","France",
                               "Sardinia","Malta","Croatia","Greece"),
                     ordered = TRUE)
  df[order(df$Colony), ]
}

emm_BA_dep <- format_phenology(emm_BA_dep)
emm_BA_arr <- format_phenology(emm_BA_arr)
#emm_WA_arr <- format_phenology(emm_WA_arr)
emm_WA_dep <- format_phenology(emm_WA_dep)

# Create summary table
summary_table <- data.frame(
  Colony = emm_BA_dep$Colony,
  BA_Departure = emm_BA_dep$formatted,
  #WA_Arrival = emm_WA_arr$formatted,
  WA_Departure = emm_WA_dep$formatted,
  BA_Arrival = emm_BA_arr$formatted
)

knitr::kable(summary_table, caption = "Estimated phenological timing across colonies (mean ± 95% CI)")
```

## Timeline Visualization

```{r timeline_prep}
# Prepare full moon dates for visualization
data_lunar <- data.frame(
  fecha = seq(as.Date('2023-03-01'), as.Date('2024-02-28'), by='days'),
  iluminacion = sapply(seq(as.Date('2023-03-01'), as.Date('2024-02-28'), by='days'),
                       lunar::lunar.illumination)
) %>%
  filter(iluminacion > 0.997) %>%
  mutate(fecha = seq(0, 365, length.out = n()))
 
# Prepare matrix with breeding and non-breeding periods in julian dates
t.list <- list()
for(i in seq(from=1, to= 8, by = 1)){
  #i=1
  t0 <- 1
  t1 <- emm_BA_dep$emmean[i]   #breeding
  t2 <- emm_BA_arr$emmean[i]   #nonbreeding
  t3 <- 365
  t <- as.matrix(c(t0,t1,t2,t3))
  t.list[i] <- list(t)
}

t.list <- Filter(Negate(is.null),t.list)
periodes <- do.call(rbind.data.frame, t.list)

# Annotate populations
periodes$condition[1:4] <- 'Ibiza'
periodes$condition[5:8] <- 'Mallorca'
periodes$condition[9:12] <- 'Minorca'
periodes$condition[13:16] <- 'France'
periodes$condition[17:20] <- 'Sardinia'
periodes$condition[21:24] <- 'Malta'
periodes$condition[25:28] <- 'Croatia'
periodes$condition[29:32] <- 'Greece'

periodes$condition <- factor(periodes$condition, 
                             levels = rev(c("Ibiza","Mallorca","Minorca","France","Sardinia","Malta","Croatia","Greece")),
                             ordered = TRUE)

# Create df for gg_vistime
time.line <- data.frame(start = periodes$V1[1:31], 
                        end = periodes$V1[2:32])%>%
  filter(!(row_number() == 4 | row_number() %% 4 == 0))%>% #Treiem les entrades amb 365 dies julians
  mutate(event = base::rep(c('B1','W','B2'), length.out = 24))%>% #Definim periodes
  mutate(color = base::rep(c("darkolivegreen4","darkorchid","darkolivegreen4"), length.out = 24))%>%
  mutate(group= c(periodes$condition[1:3],periodes$condition[5:7],periodes$condition[9:11],
                  periodes$condition[13:15],periodes$condition[17:19], periodes$condition[21:23],
                  periodes$condition[25:27], periodes$condition[29:31])) 
```

```{r timeline_plot, fig.width=10, fig.height=6}
# Create timeline plot
timeline_plot <- ggplot(time.line, aes(x = start, xend = end, y = group, yend = group, color = event)) +
  geom_segment(size = 5, alpha = 0.8) +
  scale_color_manual(
    values = c("B1" = "darkolivegreen4", "W" = "darkorchid", "B2" = "darkolivegreen4"),
    labels = c('Breeding 1', 'Non-breeding', 'Breeding 2'),
    name = 'Period'
  ) +
  labs(
    title = "Phenological Timeline Across Colonies",
    x = "Days since April 1st full moon", 
    y = "Colony"
  ) +
  annotate("point",shape=1,fill='transparent',x = Data$B_dep_jul[Data$Colony == "Sardinia"], y = 4, size = 1, colour = "black")+
  annotate("point",shape=1,fill='transparent',x = Data$BA_arr_jul[Data$Colony == "Sardinia"], y = 4, size = 1, colour = "black")+
  annotate("point",shape=1,fill='transparent',x = Data$B_dep_jul[Data$Colony == "Malta"], y = 3, size = 1, colour = "black")+
  annotate("point",shape=1,fill='transparent',x = Data$BA_arr_jul[Data$Colony == "Malta"], y = 3, size = 1, colour = "black")+
  annotate("point",shape=1,fill='transparent',x = Data$B_dep_jul[Data$Colony == "Minorca"], y = 6, size = 1, colour = "black")+
  annotate("point",shape=1,fill='transparent',x = Data$BA_arr_jul[Data$Colony == "Minorca"], y = 6, size = 1, colour = "black")+
  annotate("point",shape=1,fill='transparent',x = Data$B_dep_jul[Data$Colony == "Croatia"], y = 2, size = 1, colour = "black")+
  annotate("point",shape=1,fill='transparent',x = Data$BA_arr_jul[Data$Colony == "Croatia"], y = 2, size = 1, colour = "black")+
  annotate("point",shape=1,fill='transparent',x = Data$B_dep_jul[Data$Colony == "Greece"],y = 1, size = 1, colour = "black")+
  annotate("point",shape=1,fill='transparent',x = Data$BA_arr_jul[Data$Colony == "Greece"],y = 1, size = 1, colour = "black")+
  annotate("point",shape=1,fill='transparent',x = Data$B_dep_jul[Data$Colony == "France"], y = 5, size = 1, colour = "black")+
  annotate("point",shape=1,fill='transparent',x = Data$BA_arr_jul[Data$Colony == "France"], y = 5, size = 1, colour = "black")+
  annotate("point",shape=1,fill='transparent',x = Data$B_dep_jul[Data$Colony == "Ibiza"], y = 8, size = 1, colour = "black")+
  annotate("point",shape=1,fill='transparent',x = Data$BA_arr_jul[Data$Colony == "Ibiza"], y = 8, size = 1, colour = "black")+
  annotate("point",shape=1,fill='transparent',x = Data$B_dep_jul[Data$Colony == "Mallorca"], y = 7, size = 1, colour = "black")+
  annotate("point",shape=1,fill='transparent',x = Data$BA_arr_jul[Data$Colony == "Mallorca"], y = 7, size = 1, colour = "black")+
  geom_vline(xintercept = data_lunar$fecha, linetype = "dashed", color = "gray50", alpha = 0.5) +
  theme_classic() +
  theme(
    axis.text.y = element_text(size = 12, color = 'black'),
    axis.text.x = element_text(size = 12, color = 'black'),
    axis.title = element_text(size = 12, color = 'black'),
    legend.text = element_text(size = 12, color = 'black'),
    legend.title = element_text(size = 12, color = 'black')
  )

print(timeline_plot)
```
# Export Results

```{r export}
# Export summary table
write_tsv(summary_table, "outputs/phenology_estimates.tsv")

# Save visualization
ggsave("outputs/timeline_lunar_cycles.png", timeline_plot, width = 10, height = 6, dpi = 600)

```

# Session Information

```{r session}
sessionInfo()
```
