---
title: "Spatial Analysis of Non-Breeding Distributions"
author: "Mario Gomez Alvarez"
date: "2026-03-24"
output: html_document
---

# Abstract
This document presents a spatial analysis of non-breeding distributions for Puffinus species. The analysis includes sectioned methods, results, and comprehensive documentation suitable for publication.

# 1. Methods

## 1.1 Data Preparation
```{r data-preparation}
# Load necessary libraries
library(dplyr)
library(ggplot2)

# Load the dataset
data <- read.csv("data/non_breeding_distributions.csv")

# Clean the data
data_clean <- data %>%
  filter(!is.na(coordinate_x), !is.na(coordinate_y))
```

## 1.2 Kernel Density Estimation
```{r kernel-density-estimation}
# Perform kernel density estimation
library(sp)
kde_result <- density(data_clean$coordinate_x, data_clean$coordinate_y)
```

## 1.3 Standardization
```{r standardization}
# Standardize the results
data_standardized <- scale(data_clean)
```

# 2. Results

## 2.1 Visualization of Main Figure with Group Panels
```{r main-figure, fig.cap="Main Figure displaying group panels of non-breeding distributions."}
ggplot(data_standardized, aes(x = coordinate_x, y = coordinate_y)) +
  geom_point(aes(color = group)) +
  theme_minimal() +
  labs(title="Non-breeding Distribution by Group")
```

## 2.2 Supplementary Figure with Individual Populations
```{r supplementary-figure, fig.cap="Supplementary Figure of individual populations."}
ggplot(data_clean, aes(x = coordinate_x, y = coordinate_y)) +
  geom_point(aes(shape = population)) +
  theme_minimal() +
  labs(title="Distribution of Individual Populations")
```

# 3. Discussion
This section will include an in-depth discussion of the findings, implications for conservation, and future research directions.

# References
- Reference 1
- Reference 2
