# Spatial Analysis of Non-Breeding Distributions

## Data Preparation

```{r data-preparation}
# Load necessary libraries
library(dplyr)
library(ggplot2)

# Import data
data <- read.csv('non_breeding_data.csv')

# Clean and prepare data
prepared_data <- data %>%
  filter(!is.na(variable_of_interest)) %>%  # Remove NA values
  mutate(new_variable = log(variable_of_interest)) # Example transformation
```

## KDE Calculations

```{r kde-calculations}
# Load necessary library for KDE
library(KernSmooth)

# Compute KDE
kde_result <- bkde(prepared_data$new_variable)
```

## Standardization

```{r standardization}
# Standardizing the data
standardized_data <- scale(prepared_data$new_variable)
```

## Main Figure with Group Panels

```{r main-figure}
# Main figure with group panels
ggplot(prepared_data, aes(x = variable_of_interest, y = response_variable)) +
  geom_point() +
  facet_wrap(~ grouping_variable) +
  theme_minimal() +
  ggtitle('Main Figure: Non-Breeding Distributions')
```

## Supplementary Figure with Individual Populations

```{r supplementary-figure}
# Supplementary figure for individual populations
ggplot(prepared_data, aes(x = variable_of_interest, y = response_variable, color = population)) +
  geom_line() +
  theme_minimal() +
  ggtitle('Supplementary Figure: Individual Populations')
```

## Methods

This analysis employs a combination of KDE and statistical modeling to interpret non-breeding distribution data. 

## Results

The results indicate significant patterns in non-breeding distributions across different populations.

## Discussion

The findings contribute to our understanding of spatial ecology and non-breeding behaviors.

## Publication-Ready Formatting

This document is formatted according to the publication standards.