---
title: "Non-Breeding Distributions"
author: "Mario Gómez Álvarez"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: hide
    fig_width: 12
    fig_height: 10
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = 'center',
  dpi = 300
)

# Set timezone (important for correct temporal calculations)
Sys.setenv(TZ = 'GMT')
```

## Introduction

This analysis examines non-breeding spatial distributions of Mediterranean shearwaters (Balearic and Yelkouan) across eight biogeographical populations (WesternMost, Central-West, Central-East, EasternMost) and across the eight breeding populations. We employ kernel density estimation (KDE) to identify core (UD50%) and home range (UD90%) areas during the core non-breeding season (August-September).

### Data Standardization

Two-step standardization procedure to prevent sampling bias:

1. **Individual-level standardization**: Minimum 30 locations per individual track
2. **Population-level standardization**: Equal number of individuals per population (n = minimum across populations)

This ensures large populations do not dominate group-level kernels.

---

# Load Libraries & Set Parameters

```{r libraries}
library(tidyverse)
library(sf)
library(adehabitatHR)
library(readxl)
library(rnaturalearth)
library(rworldmap)

# Analysis parameters
BANDWIDTH <- 1.86
GRID_RES <- 200
MIN_LOCS_PER_IND <- 30
MONTHS_FILTER <- c(8, 9)
CRS_PROJ <- 4326
```

---

# 1. Load & Filter Data

```{r data-load}
# Load trip data
trips <- read.csv("trips.csv")

# Define biogeographic groups
trips_filtered <- trips %>%
  mutate(Group = case_when(
    Population %in% c('IBIZA', 'MALLORCA') ~ 'WesternMost',
    Population %in% c('MINORCA', 'FRANCE', 'SARDINIA') ~ 'Central-West',
    Population %in% c('MALTA', 'CROATIA') ~ 'Central-East',
    Population == 'GREECE' ~ 'EasternMost',
    TRUE ~ NA_character_
  )) %>%
  filter(month %in% MONTHS_FILTER, !is.na(Group))

```

---

# 2. Standardize Effort by Individual

```{r standardize-individual}
# Filter individuals with sufficient locations
set.seed(123)
trips_balanced_ind <- trips_filtered %>%
  group_by(geo_trip) %>%
  filter(n() >= MIN_LOCS_PER_IND) %>%
  slice_sample(n = MIN_LOCS_PER_IND) %>%
  ungroup()

cat("Individuals meeting minimum locations:", n_distinct(trips_balanced_ind$geo_trip), "\n")
cat("Total records after standardization:", nrow(trips_balanced_ind), "\n")
```

---

# 3. Standardize Effort by Population

```{r standardize-population}
# Identify minimum number of individuals across populations
min_inds <- trips_balanced_ind %>%
  group_by(Population) %>%
  summarise(n_ind = n_distinct(geo_trip), .groups = 'drop') %>%
  pull(n_ind) %>%
  min()

cat("Minimum individuals per population:", min_inds, "\n\n")

# Sample equal number of individuals per population
selected_ids <- trips_balanced_ind %>%
  group_by(Population) %>%
  distinct(geo_trip) %>%
  slice_sample(n = min_inds) %>%
  pull(geo_trip)

trips_final <- trips_balanced_ind %>% filter(geo_trip %in% selected_ids)

```

---

# 4. Calculate Group-Level KDE

```{r kde-groups}
# Convert to SpatialPointsDataFrame
coords <- trips_final[, c("Lon", "Lat")]
sp_df <- SpatialPointsDataFrame(
  coords,
  data = data.frame(Group = trips_final$Group),
  proj4string = CRS("+proj=longlat +datum=WGS84")
)

# Calculate KDE by group
kde_groups <- kernelUD(sp_df[, "Group"], h = BANDWIDTH, grid = GRID_RES, same4all = TRUE)

# Extract contours
ver_90 <- st_as_sf(getverticeshr(kde_groups, percent = 90))
ver_50 <- st_as_sf(getverticeshr(kde_groups, percent = 50))

# Rename ID column for consistency
ver_90$Group <- ver_90$id
ver_50$Group <- ver_50$id

cat("Group-level KDE calculated successfully\n")
```

---

# 5. Load Base Map & Colonies

```{r basedata}
# Base map
wrld <- rworldmap::getMap(resolution = "high")
sf_wrld <- st_as_sf(wrld) %>% st_transform(CRS_PROJ)

# Colonies coordinates
colonies <- read_excel('colonies_coords.csv')

# Convert to sf object
origin_sf <- st_as_sf(colonies, coords = c("Lon", "Lat"), crs = CRS_PROJ)

cat("Colonies data loaded:", nrow(origin_sf), "colonies\n")
```

---

# 6. Calculate Individual Activity Centers

```{r individual-centroids}
# Define factor levels for consistent ordering
pop_order <- c('IBIZA', 'MALLORCA', 'MINORCA', 'FRANCE', 'SARDINIA', 'MALTA', 'CROATIA', 'GREECE')
pop_order_lower <- tolower(pop_order)

# Calculate 5% UD contour for each individual
individual_k5 <- trips_balanced_ind %>%
  group_split(geo_trip) %>%
  map_dfr(function(df) {
    coords <- df[, c("Lon", "Lat")]
    sp_pts <- SpatialPoints(coords, proj4string = CRS("+proj=longlat +datum=WGS84"))

    kde_ind <- kernelUD(sp_pts, h = BANDWIDTH, grid = 100)
    ver_5 <- st_as_sf(getverticeshr(kde_ind, percent = 5))

    centroid <- st_centroid(ver_5) %>%
      mutate(
        geo_trip = unique(df$geo_trip),
        Pop = unique(df$Population),
        Group = unique(df$Group)
      ) %>%
      select(geo_trip, Pop, Group)

    return(centroid)
  })

cat("Individual activity centers calculated:", nrow(individual_k5), "individuals\n")
```

---

# 7. Set Factor Levels & Order Data

```{r factor-ordering}
# Define levels
group_order <- c('WesternMost', 'Central-West', 'Central-East', 'EasternMost')

# Apply to all objects
origin_sf$Pop <- factor(origin_sf$Pop, levels = pop_order_lower)
origin_sf$Group <- factor(origin_sf$Group, levels = group_order)

individual_k5$Pop <- factor(individual_k5$Pop, levels = pop_order)
individual_k5$Group <- factor(individual_k5$Group, levels = group_order)

ver_90$Group <- factor(ver_90$Group, levels = group_order)
ver_50$Group <- factor(ver_50$Group, levels = group_order)
```

---

# 8. Main Figure: Group-Level Distributions (Figure 2)

```{r main-figure, fig.width=12, fig.height=10}
# Panel labels
panel_labels <- c(
  'WesternMost' = "A",
  'Central-West' = "B",
  'Central-East' = "C",
  'EasternMost' = "D"
)

p <- ggplot() +
  # Layer 1: Group-level kernels (background)
  geom_sf(data = ver_90, fill = "#355C7D", color = NA, alpha = 0.15) +
  geom_sf(data = ver_50, fill = "#355C7D", color = NA, alpha = 0.3) +

  # Layer 2: Base map
  geom_sf(data = sf_wrld, fill = "gray60", color = "black", linewidth = 0.1) +

  # Layer 3: Individual activity centers (colored by population)
  geom_sf(data = individual_k5, aes(color = Pop), size = 1.2, alpha = 0.7, show.legend = FALSE) +

  # Layer 4: Origin colonies (triangle markers)
  geom_sf(data = origin_sf, aes(fill = Pop), size = 3, shape = 24, color = "black", stroke = 0.5) +

  # Aesthetics
  facet_wrap(~Group, ncol = 2, labeller = as_labeller(panel_labels)) +
  scale_fill_viridis_d(option = "turbo", name = "Population", drop = FALSE) +
  scale_color_viridis_d(option = "turbo", drop = FALSE) +
  coord_sf(xlim = c(-15, 42), ylim = c(28, 50)) +
  
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold", size = 12, hjust = 0),
    legend.position = "top",
    panel.grid = element_blank()
  )

print(p)
```

```{r save-main-figure}
ggsave(
  filename = "outputs/Figure_2_NonBreeding_Panels.png",
  plot = p,
  device = "png",
  width = 210,
  height = 150,
  units = "mm",
  dpi = 300,
  bg = "white"
)

cat("Main figure saved: Figure_2_NonBreeding_Panels.png\n")
```

---

# 9. Calculate Population-Level KDE

```{r kde-populations}
# Ensure proper factor ordering for populations
trips_balanced_ind$Population <- factor(trips_balanced_ind$Population, levels = pop_order)

# Convert to SpatialPointsDataFrame for population-level KDE
coords_pop <- trips_balanced_ind[, c("Lon", "Lat")]
sp_pop <- SpatialPointsDataFrame(
  coords_pop,
  data = data.frame(Pop = trips_balanced_ind$Population),
  proj4string = CRS("+proj=longlat +datum=WGS84")
)

# Calculate KDE by population
kde_pops <- kernelUD(sp_pop[, "Pop"], h = BANDWIDTH, grid = GRID_RES, same4all = TRUE)

# Extract contours
ver_90_pop <- st_as_sf(getverticeshr(kde_pops, percent = 90))
ver_50_pop <- st_as_sf(getverticeshr(kde_pops, percent = 50))

# Rename ID column
names(ver_90_pop)[1] <- "Pop"
names(ver_50_pop)[1] <- "Pop"

# Apply consistent factor levels (uppercase)
ver_90_pop$Pop <- factor(toupper(ver_90_pop$Pop), levels = pop_order)
ver_50_pop$Pop <- factor(toupper(ver_50_pop$Pop), levels = pop_order)
individual_k5$Pop <- factor(toupper(individual_k5$Pop), levels = pop_order)
origin_sf$Pop <- factor(toupper(origin_sf$Pop), levels = pop_order_lower)

cat("Population-level KDE calculated successfully\n")
```

---

# 10. Supplementary Figure: Population-Level Distributions

```{r supplementary-figure, fig.width=12, fig.height=16}
# Prepare annotation labels with sample sizes
ann_labels <- data.frame(
  Pop = factor(pop_order, levels = pop_order),
  label = c(
    'Ibiza (Sa Conillera, n = 22)',
    'Mallorca (Sa Cella, n = 70)',
    'Minorca (La Mola, n = 12)',
    'France (Port-Cros & Porquerolles, n = 40)',
    'Sardinia (Colata & Contrabbandieri, n = 19)',
    'Malta (Majjistral & Rdum, n = 12)',
    'Croatia (Zaklopatica & Veli Maslovnjak, n = 9)',
    'EasternMost (Gyaros & Kappari, n = 8)'
  ),
  x = 40,
  y = 29
)

figura_sup <- ggplot() +
  # Layer 1: Population-level kernels
  geom_sf(data = ver_90_pop, fill = "#355C7D", color = NA, alpha = 0.2) +
  geom_sf(data = ver_50_pop, fill = "#355C7D", color = NA, alpha = 0.5) +

  # Layer 2: Base map
  geom_sf(data = sf_wrld, fill = "gray60", color = "black", linewidth = 0.1) +

  # Layer 3: Individual activity centers
  geom_sf(data = individual_k5, color = "black", size = 1, alpha = 0.7) +

  # Layer 4: Origin colonies
  geom_sf(data = origin_sf, fill = "yellow", size = 3, shape = 24, color = "black", stroke = 0.3) +

  # Layer 5: Population labels with sample sizes
  geom_text(
    data = ann_labels, aes(x = x, y = y, label = label),
    hjust = 1, vjust = 0, size = 3, fontface = "bold.italic", color = "black"
  ) +

  # Aesthetics
  coord_sf(xlim = c(-15, 42), ylim = c(28, 50), expand = FALSE) +
  facet_wrap(~Pop, ncol = 2) +
  
  theme_minimal() +
  theme(
    strip.text = element_blank(),
    axis.text = element_blank(),
    axis.title = element_blank(),
    legend.position = "none"
  )

print(figura_sup)
```

```{r save-supplementary-figure}
ggsave(
  filename = "outputs/Supplementary_Figure_KDE_8Populations.png",
  plot = figura_sup,
  device = "png",
  width = 210,
  height = 270,
  units = "mm",
  dpi = 600,
  bg = "white"
)

cat("Supplementary figure saved: Supplementary_Figure_KDE_8Populations.png\n")

```

## Session Information

```{r session-info}
sessionInfo()
```
