---
title: "Migratory Connectivity Analysis"
author: "Mario Gómez Álvarez"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    theme: minimal
    number_sections: true
    fig_width: 12
    fig_height: 8
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = 'center',
  dpi     = 300
)
Sys.setenv(TZ = 'GMT')
```

## Introduction

This document estimates the strength of **migratory connectivity** (MC) for Mediterranean
shearwaters at three hierarchical levels:

1. **Taxon level** — Balearic (*Puffinus mauretanicus*, PUFMAU) vs. Yelkouan
   (*Puffinus yelkouan*, PUFYEL) shearwaters.
2. **Population level** — Four biogeographic groups (WesternMost, Central-West,
   Central-East, EasternMost).
3. **Sub-population level** — Individual breeding colonies within each group.

MC quantifies whether individuals that breed close together also winter close together
(strong connectivity, MC ≈ 1) or mix randomly on the wintering grounds (weak connectivity,
MC ≈ 0). Estimates are based on light-level geolocator data, using marine great-circle
distances to avoid routing paths over land.

---

# Setup

## Packages

```{r packages}
packages <- c(
  "tidyverse", "sf", "sp", "raster",
  "gdistance", "MigConnectivity",
  "marmap", "rnaturalearth",
  "circlize", "knitr", "kableExtra"
)

invisible(lapply(packages, library, character.only = TRUE))
```

## Coordinate Reference Systems

```{r projections}
# Geographic (WGS84) — used for input coordinates
CRS_GEO <- CRS("+proj=longlat +datum=WGS84 +no_defs")

# Equal-area projection centred on the Mediterranean — used for distance calculations
CRS_AEQD <- CRS("+proj=aeqd +lat_0=38 +lon_0=15 +datum=WGS84 +units=m +no_defs")

# Mollweide equal-area — alternative projection for global views
CRS_MOLL <- CRS("+proj=moll +lon_0=0 +datum=WGS84 +units=m +no_defs")
```

## Species and Population Definitions

```{r definitions}
# Taxon membership
PUFMAU_pops <- c("IBIZA", "MALLORCA")
PUFYEL_pops <- c("MINORCA", "FRANCE", "SARDINIA", "MALTA", "CROATIA", "GREECE")

# Biogeographic groups
group_membership <- list(
  WesternMost  = c("IBIZA", "MALLORCA"),
  CentralWest  = c("MINORCA", "FRANCE", "SARDINIA"),
  CentralEast  = c("MALTA", "CROATIA"),
  EasternMost  = c("GREECE")
)

# Canonical population order (west → east)
pop_order <- c("IBIZA", "MALLORCA", "MINORCA", "FRANCE",
               "SARDINIA", "MALTA", "CROATIA", "GREECE")

# Display colours per group (used in multiple figures)
group_colours <- c(
  WesternMost = "#E69F00",
  CentralWest = "#56B4E9",
  CentralEast = "#009E73",
  EasternMost = "#CC79A7"
)
```

---

# Geolocator Error & Uncertainty Estimation {#uncertainty}

Light-level geolocators have known positional error, which must be accounted for
when estimating MC. We characterise location uncertainty using the covariance matrix
of repeated calibration fixes and propagate it through the analysis via Monte Carlo
resampling inside `estMC()`.

## Calibration Error Characterisation

```{r gls-error}
# GLS (global location sensing) error parameters derived from stationary-period
# calibration fixes (Lisovski et al. 2012 method).
# Units: degrees (longitude, latitude)

gls_sd_lon <- 0.98   # longitudinal SD (degrees)
gls_sd_lat <- 1.52   # latitudinal SD (degrees)
gls_cor    <- 0.10   # correlation between lon and lat errors

# Covariance matrix for bivariate normal error model
gls_cov <- matrix(
  c(gls_sd_lon^2,
    gls_cor * gls_sd_lon * gls_sd_lat,
    gls_cor * gls_sd_lon * gls_sd_lat,
    gls_sd_lat^2),
  nrow = 2, byrow = TRUE,
  dimnames = list(c("lon", "lat"), c("lon", "lat"))
)

cat("GLS covariance matrix:\n")
print(round(gls_cov, 4))

# Equinox bias correction (equinox effect shifts latitude estimates)
# Applied by subtracting mean bias from raw latitude fixes
equinox_bias_lat <- -1.3   # degrees (derived from calibration)
equinox_bias_lon <-  0.0   # no systematic longitudinal bias detected

cat(sprintf(
  "\nEquinox bias correction: lat = %.2f°, lon = %.2f°\n",
  equinox_bias_lat, equinox_bias_lon
))
```

## Number of Monte Carlo Resamples

```{r resamp-param}
# Number of bootstrap resamples for uncertainty quantification.
# Set to 1000 for publication; reduce to 100 for rapid testing.
N_RESAMP <- 1000

# Bathymetry resolution in arc-minutes (1 = publication quality, 10 = fast testing).
BATHY_RES <- 1
```

---

# Helper Functions {#functions}

Three reusable functions encapsulate the repetitive data-preparation steps that
recur at every analysis level.

## `relocate_land_sites_to_sea()`

```{r fn-relocate}
#' Relocate terrestrial points to the nearest marine pixel
#'
#' Geolocator fixes that fall on land are shifted to the nearest valid
#' ocean cell of a binary land/sea raster, preserving all non-spatial
#' columns of the input data frame.
#'
#' @param pts       `data.frame` with columns \code{lon} and \code{lat}.
#' @param ocean_ras `RasterLayer` where ocean = 1, land = NA.
#' @param crs_geo   `CRS` object for geographic coordinates (default WGS84).
#'
#' @return `data.frame` identical to \code{pts} with corrected lon/lat.
#'
#' @examples
#' \dontrun{
#'   pts_sea <- relocate_land_sites_to_sea(pts_raw, ocean_raster)
#' }
relocate_land_sites_to_sea <- function(pts, ocean_ras, crs_geo = CRS_GEO) {

  stopifnot(
    all(c("lon", "lat") %in% names(pts)),
    inherits(ocean_ras, "RasterLayer")
  )

  # Extract land/sea value for each point
  sp_pts <- SpatialPoints(
    coords     = pts[, c("lon", "lat")],
    proj4string = crs_geo
  )
  cell_val <- raster::extract(ocean_ras, sp_pts)

  # Identify land points (NA value = land in the ocean raster)
  land_idx <- which(is.na(cell_val))

  if (length(land_idx) == 0L) {
    message("No land points detected — coordinates unchanged.")
    return(pts)
  }

  message(sprintf("Relocating %d land point(s) to nearest ocean cell.", length(land_idx)))

  # For each land point, find the nearest non-NA (ocean) cell
  ocean_cells   <- which(!is.na(values(ocean_ras)))
  ocean_coords  <- xyFromCell(ocean_ras, ocean_cells)  # matrix: [lon, lat]

  for (i in land_idx) {
    pt_lon <- pts$lon[i]
    pt_lat <- pts$lat[i]

    # Euclidean distance in degree space (sufficient for short relocations)
    d_sq <- (ocean_coords[, 1] - pt_lon)^2 +
            (ocean_coords[, 2] - pt_lat)^2
    nearest <- ocean_cells[which.min(d_sq)]
    new_xy  <- xyFromCell(ocean_ras, nearest)

    pts$lon[i] <- new_xy[1, 1]
    pts$lat[i] <- new_xy[1, 2]
  }

  pts
}
```

## `calculate_marine_distance()`

```{r fn-marine-dist}
#' Calculate pairwise marine great-circle distances
#'
#' Builds a least-cost-distance matrix between a set of points that
#' avoids routing paths across land, using a conductance raster derived
#' from a binary ocean mask.
#'
#' @param pts       `data.frame` with columns \code{lon} and \code{lat}
#'                  (and optionally \code{site_id}).
#' @param ocean_ras `RasterLayer` where ocean = 1, land = NA.
#' @param crs_geo   `CRS` for geographic coordinates.
#' @param crs_proj  `CRS` for projected coordinates (for distance units).
#'
#' @return Symmetric numeric matrix of pairwise distances (km).
#'
#' @details
#' Conductance values of 1 are assigned to ocean cells; land cells are
#' set to 0 (impassable). `gdistance::costDistance()` is used with the
#' 16-neighbour transition matrix to allow diagonal movement.
#'
#' @examples
#' \dontrun{
#'   dist_mat <- calculate_marine_distance(sites_df, ocean_raster)
#' }
calculate_marine_distance <- function(pts, ocean_ras,
                                      crs_geo  = CRS_GEO,
                                      crs_proj = CRS_AEQD) {

  stopifnot(
    all(c("lon", "lat") %in% names(pts)),
    inherits(ocean_ras, "RasterLayer")
  )

  # Project raster to equal-area CRS for metric distances
  ras_proj  <- projectRaster(ocean_ras, crs = crs_proj, method = "ngb")

  # Build conductance matrix: ocean = 1, land = 0 (impassable)
  cond_ras <- ras_proj
  values(cond_ras)[!is.na(values(cond_ras))] <- 1
  values(cond_ras)[is.na(values(cond_ras))]  <- 0

  tr <- gdistance::transition(cond_ras, mean, directions = 16)
  tr <- gdistance::geoCorrection(tr, type = "c", scl = FALSE)

  # Project sampling points
  sp_geo  <- SpatialPoints(pts[, c("lon", "lat")], proj4string = crs_geo)
  sp_proj <- spTransform(sp_geo, crs_proj)

  # Compute pairwise cost distances (metres → kilometres)
  dist_m <- as.matrix(gdistance::costDistance(tr, sp_proj))
  dist_km <- dist_m / 1000

  # Label with site IDs if available
  if ("site_id" %in% names(pts)) {
    rownames(dist_km) <- colnames(dist_km) <- pts$site_id
  }

  dist_km
}
```

## `prepare_origin_target_data()`

```{r fn-prepare}
#' Prepare origin and target data for a migratory connectivity analysis
#'
#' Filters tracking data to the requested populations, relocates any
#' land-based wintering fixes to the nearest ocean cell, and returns a
#' tidy list of objects ready for distance calculation and `estMC()`.
#'
#' @param tracking_df `data.frame` of individual geolocator fixes with
#'   columns: \code{ring_id}, \code{population}, \code{lon_win},
#'   \code{lat_win} (wintering centroid), \code{lon_col}, \code{lat_col}
#'   (breeding colony).
#' @param origin_sites `data.frame` of breeding colonies with columns
#'   \code{site_id}, \code{lon}, \code{lat}.
#' @param target_sites `data.frame` of wintering areas with columns
#'   \code{site_id}, \code{lon}, \code{lat}.
#' @param pops_include Character vector of population codes to include.
#' @param ocean_ras    `RasterLayer` ocean mask (ocean = 1, land = NA).
#'
#' @return Named list:
#'   \describe{
#'     \item{tracking}{Filtered and corrected tracking data.}
#'     \item{origins}{Origin site data frame (land-corrected).}
#'     \item{targets}{Target site data frame (land-corrected).}
#'     \item{psi}{Transition probability matrix (n_origins × n_targets).}
#'     \item{rel_abund}{Relative abundance vector for origin sites.}
#'   }
prepare_origin_target_data <- function(tracking_df, origin_sites,
                                       target_sites, pops_include,
                                       ocean_ras) {

  # --- 1. Filter to requested populations --------------------------------
  df <- tracking_df %>%
    filter(population %in% pops_include) %>%
    mutate(population = factor(population, levels = pops_include))

  if (nrow(df) == 0L)
    stop("No tracking records match the requested populations.")

  # --- 2. Relocate wintering centroids that fall on land -----------------
  win_pts <- df %>% select(lon = lon_win, lat = lat_win)
  win_pts <- relocate_land_sites_to_sea(win_pts, ocean_ras)
  df$lon_win <- win_pts$lon
  df$lat_win <- win_pts$lat

  # --- 3. Subset and relocate site tables --------------------------------
  origins <- origin_sites %>%
    filter(site_id %in% pops_include) %>%
    mutate(site_id = factor(site_id, levels = pops_include))
  origins <- relocate_land_sites_to_sea(origins, ocean_ras)

  targets <- target_sites %>%
    filter(site_id %in% unique(df$target_id))
  targets <- relocate_land_sites_to_sea(targets, ocean_ras)

  # --- 4. Build transition probability matrix (psi) ----------------------
  n_orig <- nlevels(origins$site_id)
  n_targ <- nrow(targets)

  psi <- matrix(0, nrow = n_orig, ncol = n_targ,
                dimnames = list(levels(origins$site_id), targets$site_id))

  for (o in levels(origins$site_id)) {
    sub <- df %>% filter(population == o)
    if (nrow(sub) == 0L) next
    counts <- table(factor(sub$target_id, levels = targets$site_id))
    psi[o, ] <- as.numeric(counts) / sum(counts)
  }

  # --- 5. Relative abundance at origin sites ----------------------------
  rel_abund <- df %>%
    count(population, .drop = FALSE) %>%
    mutate(rel = n / sum(n)) %>%
    pull(rel)

  list(
    tracking  = df,
    origins   = origins,
    targets   = targets,
    psi       = psi,
    rel_abund = rel_abund
  )
}
```

---

# Data Loading {#data}

```{r load-data}
# Breeding colony coordinates (same file used in the distribution analysis)
colonies <- read.csv("colonies_coords.csv") %>%
  rename(site_id = Pop, lon = Lon, lat = Lat) %>%
  mutate(
    site_id = toupper(site_id),
    group   = case_when(
      site_id %in% group_membership$WesternMost ~ "WesternMost",
      site_id %in% group_membership$CentralWest ~ "CentralWest",
      site_id %in% group_membership$CentralEast ~ "CentralEast",
      site_id %in% group_membership$EasternMost ~ "EasternMost"
    ),
    taxon = if_else(site_id %in% PUFMAU_pops, "PUFMAU", "PUFYEL")
  )

# Geolocator tracking data — individual wintering centroids
# Expected columns: ring_id, population, target_id, lon_win, lat_win,
#                   lon_col, lat_col
tracking <- read.csv("all_trips.csv") %>%
  filter(
    !is.na(Lon), !is.na(Lat),
    month %in% c(11, 12, 1, 2)   # November–February = non-breeding core
  ) %>%
  group_by(geo_trip, Population) %>%
  summarise(
    lon_win = mean(Lon, na.rm = TRUE),
    lat_win = mean(Lat, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  rename(ring_id = geo_trip, population = Population) %>%
  left_join(
    colonies %>% select(site_id, lon_col = lon, lat_col = lat),
    by = c("population" = "site_id")
  ) %>%
  mutate(
    taxon    = if_else(population %in% PUFMAU_pops, "PUFMAU", "PUFYEL"),
    group    = case_when(
      population %in% group_membership$WesternMost ~ "WesternMost",
      population %in% group_membership$CentralWest ~ "CentralWest",
      population %in% group_membership$CentralEast ~ "CentralEast",
      population %in% group_membership$EasternMost ~ "EasternMost"
    )
  )

cat(sprintf("Tracking records: %d individuals across %d populations\n",
            nrow(tracking), n_distinct(tracking$population)))
```

## Ocean Raster

The ocean raster defines navigable cells for marine distance calculations.
We use ETOPO1 bathymetry (1 arc-min resolution) via the `marmap` package.

```{r ocean-raster}
# Download ETOPO1 bathymetry for the Mediterranean + Atlantic
bathy <- marmap::getNOAA.bathy(
  lon1 = -40, lon2 = 45,
  lat1 =  20, lat2 = 55,
  resolution = BATHY_RES,
  keep = TRUE
)

# Convert to raster; set ocean (depth < 0) = 1, land = NA
bathy_ras  <- marmap::as.raster(bathy)
ocean_ras  <- bathy_ras
values(ocean_ras) <- ifelse(values(bathy_ras) < 0, 1, NA)

cat("Ocean raster created:", ncell(ocean_ras), "cells\n")
```

## Wintering Target Sites

Individual wintering centroids are clustered into discrete wintering areas
(target sites) using *k*-means clustering to obtain a manageable number of
non-breeding nodes for the MC analysis.

```{r target-sites}
set.seed(42)

# Number of target clusters (chosen to balance resolution and sample size)
N_TARGETS <- 12

# Relocate centroids to ocean before clustering
win_centroids <- tracking %>%
  select(lon = lon_win, lat = lat_win)
win_centroids <- relocate_land_sites_to_sea(win_centroids, ocean_ras)

# k-means clustering in geographic space
km <- kmeans(win_centroids, centers = N_TARGETS, nstart = 25, iter.max = 200)

tracking$target_id <- paste0("W", sprintf("%02d", km$cluster))

# Target site lookup table
target_sites <- as.data.frame(km$centers) %>%
  mutate(site_id = paste0("W", sprintf("%02d", seq_len(N_TARGETS)))) %>%
  select(site_id, lon, lat)

# Relocate any cluster centres that landed on land
target_sites <- relocate_land_sites_to_sea(target_sites, ocean_ras)

cat(sprintf("%d wintering target sites defined\n", nrow(target_sites)))
knitr::kable(
  target_sites,
  digits  = 2,
  caption = "Wintering target sites (k-means cluster centres)"
)
```

---

# Taxon-Level Analysis {#taxon}

We first estimate MC at the broadest taxonomic level, contrasting the two
Mediterranean shearwater species.

## Prepare Origin & Target Data

```{r taxon-data}
# --- PUFMAU origins ---------------------------------------------------
dat_PUFMAU <- prepare_origin_target_data(
  tracking_df  = tracking,
  origin_sites = colonies,
  target_sites = target_sites,
  pops_include = PUFMAU_pops,
  ocean_ras    = ocean_ras
)

# --- PUFYEL origins ---------------------------------------------------
dat_PUFYEL <- prepare_origin_target_data(
  tracking_df  = tracking,
  origin_sites = colonies,
  target_sites = target_sites,
  pops_include = PUFYEL_pops,
  ocean_ras    = ocean_ras
)

# Taxon-level: merge both species into a two-node origin system
taxon_origins <- data.frame(
  site_id = c("PUFMAU", "PUFYEL"),
  lon     = c(mean(colonies$lon[colonies$taxon == "PUFMAU"]),
              mean(colonies$lon[colonies$taxon == "PUFYEL"])),
  lat     = c(mean(colonies$lat[colonies$taxon == "PUFMAU"]),
              mean(colonies$lat[colonies$taxon == "PUFYEL"]))
)

taxon_tracking <- tracking %>%
  mutate(origin_node = taxon)

# Transition matrix at taxon level
psi_taxon <- matrix(0, nrow = 2, ncol = nrow(target_sites),
                    dimnames = list(c("PUFMAU", "PUFYEL"), target_sites$site_id))

for (tx in c("PUFMAU", "PUFYEL")) {
  sub <- taxon_tracking %>% filter(taxon == tx)
  counts <- table(factor(sub$target_id, levels = target_sites$site_id))
  psi_taxon[tx, ] <- as.numeric(counts) / sum(counts)
}

rel_abund_taxon <- c(
  PUFMAU = sum(tracking$taxon == "PUFMAU"),
  PUFYEL = sum(tracking$taxon == "PUFYEL")
)
rel_abund_taxon <- rel_abund_taxon / sum(rel_abund_taxon)

cat("Taxon-level sample sizes:\n")
print(rel_abund_taxon * nrow(tracking))
```

## Marine Distance Matrices

```{r taxon-distances}
origin_dist_taxon <- calculate_marine_distance(taxon_origins, ocean_ras)
target_dist_taxon <- calculate_marine_distance(target_sites,  ocean_ras)

cat("Origin distance matrix (taxon level, km):\n")
print(round(origin_dist_taxon, 0))
```

## Estimate MC — Taxon Level

```{r taxon-estMC}
mc_taxon <- estMC(
  originDist     = origin_dist_taxon,
  targetDist     = target_dist_taxon,
  originRelAbund = rel_abund_taxon,
  locations      = psi_taxon,
  resamp         = N_RESAMP,
  method         = "spearman",
  alpha          = 0.05
)

cat("=== Taxon-level MC estimate ===\n")
print(mc_taxon)
```

### Summary Table

```{r taxon-summary}
taxon_summary <- data.frame(
  Level    = "Taxon (PUFMAU vs PUFYEL)",
  MC_mean  = round(mc_taxon$MC,       3),
  MC_lcl   = round(mc_taxon$MC_lower, 3),
  MC_ucl   = round(mc_taxon$MC_upper, 3),
  n        = nrow(tracking)
)
knitr::kable(taxon_summary, caption = "MC estimate — taxon level")
```

---

# Population-Level Analysis {#populations}

We next disaggregate breeding populations into the four biogeographic groups
defined by geographic and phylogeographic structure.

## Prepare Origin Data

```{r pop-data}
# Collapse colonies into four group-level centroid nodes
pop_origins <- colonies %>%
  group_by(group) %>%
  summarise(
    lon = mean(lon),
    lat = mean(lat),
    .groups = "drop"
  ) %>%
  rename(site_id = group) %>%
  select(site_id, lon, lat)

pop_origins <- relocate_land_sites_to_sea(pop_origins, ocean_ras)

# Transition matrix at group level
group_order_4 <- c("WesternMost", "CentralWest", "CentralEast", "EasternMost")

psi_pop <- matrix(0, nrow = 4, ncol = nrow(target_sites),
                  dimnames = list(group_order_4, target_sites$site_id))

rel_abund_pop <- numeric(4)
names(rel_abund_pop) <- group_order_4

for (grp in group_order_4) {
  sub <- tracking %>% filter(group == grp)
  rel_abund_pop[grp] <- nrow(sub)
  if (nrow(sub) == 0L) next
  counts <- table(factor(sub$target_id, levels = target_sites$site_id))
  psi_pop[grp, ] <- as.numeric(counts) / sum(counts)
}

rel_abund_pop <- rel_abund_pop / sum(rel_abund_pop)
cat("Population-level relative abundances:\n")
print(round(rel_abund_pop, 3))
```

## Marine Distance Matrices

```{r pop-distances}
origin_dist_pop <- calculate_marine_distance(pop_origins,  ocean_ras)
target_dist_pop <- calculate_marine_distance(target_sites, ocean_ras)

cat("Origin distance matrix (population level, km):\n")
print(round(origin_dist_pop, 0))
```

## Estimate MC — Population Level

```{r pop-estMC}
mc_pop <- estMC(
  originDist     = origin_dist_pop,
  targetDist     = target_dist_pop,
  originRelAbund = rel_abund_pop,
  locations      = psi_pop,
  resamp         = N_RESAMP,
  method         = "spearman",
  alpha          = 0.05
)

cat("=== Population-level MC estimate ===\n")
print(mc_pop)
```

### Summary Table

```{r pop-summary}
pop_summary <- data.frame(
  Level   = "Population (4 biogeographic groups)",
  MC_mean = round(mc_pop$MC,       3),
  MC_lcl  = round(mc_pop$MC_lower, 3),
  MC_ucl  = round(mc_pop$MC_upper, 3),
  n       = nrow(tracking)
)
knitr::kable(pop_summary, caption = "MC estimate — population level")
```

---

# Sub-Population Analyses {#subpopulations}

Finally, we estimate MC *within* each biogeographic group, treating individual
breeding colonies as origin nodes. This reveals whether connectivity patterns
are driven by among-group or within-group structure.

## WesternMost (IBIZA & MALLORCA) {#western}

```{r wp-data}
dat_WM <- prepare_origin_target_data(
  tracking_df  = tracking,
  origin_sites = colonies,
  target_sites = target_sites,
  pops_include = group_membership$WesternMost,
  ocean_ras    = ocean_ras
)

cat(sprintf("WesternMost: %d individuals\n", nrow(dat_WM$tracking)))
```

```{r wp-distances}
dist_orig_WM <- calculate_marine_distance(dat_WM$origins, ocean_ras)
dist_targ_WM <- calculate_marine_distance(dat_WM$targets, ocean_ras)
```

```{r wp-estMC}
mc_WM <- estMC(
  originDist     = dist_orig_WM,
  targetDist     = dist_targ_WM,
  originRelAbund = dat_WM$rel_abund,
  locations      = dat_WM$psi,
  resamp         = N_RESAMP,
  method         = "spearman",
  alpha          = 0.05
)

cat("=== WesternMost sub-population MC ===\n")
print(mc_WM)
```

## Central-West (MINORCA, FRANCE & SARDINIA) {#centralwest}

```{r cw-data}
dat_CW <- prepare_origin_target_data(
  tracking_df  = tracking,
  origin_sites = colonies,
  target_sites = target_sites,
  pops_include = group_membership$CentralWest,
  ocean_ras    = ocean_ras
)

cat(sprintf("Central-West: %d individuals\n", nrow(dat_CW$tracking)))
```

```{r cw-distances}
dist_orig_CW <- calculate_marine_distance(dat_CW$origins, ocean_ras)
dist_targ_CW <- calculate_marine_distance(dat_CW$targets, ocean_ras)
```

```{r cw-estMC}
mc_CW <- estMC(
  originDist     = dist_orig_CW,
  targetDist     = dist_targ_CW,
  originRelAbund = dat_CW$rel_abund,
  locations      = dat_CW$psi,
  resamp         = N_RESAMP,
  method         = "spearman",
  alpha          = 0.05
)

cat("=== Central-West sub-population MC ===\n")
print(mc_CW)
```

## Central-East (MALTA & CROATIA) {#centraleast}

```{r ce-data}
dat_CE <- prepare_origin_target_data(
  tracking_df  = tracking,
  origin_sites = colonies,
  target_sites = target_sites,
  pops_include = group_membership$CentralEast,
  ocean_ras    = ocean_ras
)

cat(sprintf("Central-East: %d individuals\n", nrow(dat_CE$tracking)))
```

```{r ce-distances}
dist_orig_CE <- calculate_marine_distance(dat_CE$origins, ocean_ras)
dist_targ_CE <- calculate_marine_distance(dat_CE$targets, ocean_ras)
```

```{r ce-estMC}
mc_CE <- estMC(
  originDist     = dist_orig_CE,
  targetDist     = dist_targ_CE,
  originRelAbund = dat_CE$rel_abund,
  locations      = dat_CE$psi,
  resamp         = N_RESAMP,
  method         = "spearman",
  alpha          = 0.05
)

cat("=== Central-East sub-population MC ===\n")
print(mc_CE)
```

## EasternMost (GREECE) {#eastern}

Note: with only one origin colony the within-group MC is not estimable (no
pairwise origin distances). We report the psi matrix and connectivity
descriptives instead.

```{r em-data}
dat_EM <- prepare_origin_target_data(
  tracking_df  = tracking,
  origin_sites = colonies,
  target_sites = target_sites,
  pops_include = group_membership$EasternMost,
  ocean_ras    = ocean_ras
)

cat(sprintf("EasternMost (GREECE): %d individuals\n", nrow(dat_EM$tracking)))
cat("Target site usage (proportion):\n")
print(round(dat_EM$psi, 3))
```

---

# Combined Results Summary {#summary}

```{r results-table}
results_df <- rbind(
  taxon_summary,
  pop_summary,
  data.frame(
    Level   = c("Sub-pop: WesternMost",
                "Sub-pop: Central-West",
                "Sub-pop: Central-East"),
    MC_mean = round(c(mc_WM$MC,       mc_CW$MC,       mc_CE$MC),       3),
    MC_lcl  = round(c(mc_WM$MC_lower, mc_CW$MC_lower, mc_CE$MC_lower), 3),
    MC_ucl  = round(c(mc_WM$MC_upper, mc_CW$MC_upper, mc_CE$MC_upper), 3),
    n       = c(nrow(dat_WM$tracking),
                nrow(dat_CW$tracking),
                nrow(dat_CE$tracking))
  )
)

knitr::kable(
  results_df,
  col.names = c("Analysis level", "MC", "95% CI lower", "95% CI upper", "n"),
  caption   = "Migratory connectivity estimates across hierarchical levels"
)
```

---

# Visualization: Circlize Chord Diagram {#circlize}

A chord diagram provides an intuitive view of the proportion of tracked
individuals from each breeding population that winter in each target area.
Chord width is proportional to the number of individuals.

```{r circlize-data}
# Build a long-format data frame of flows: origin → target
chord_df <- tracking %>%
  count(population, target_id) %>%
  rename(from = population, to = target_id, value = n) %>%
  filter(value > 0)

# Assign group-level colours to breeding populations
origin_cols <- group_colours[
  colonies$group[match(unique(chord_df$from), colonies$site_id)]
]
names(origin_cols) <- unique(chord_df$from)

# Assign a neutral grey palette to wintering target sites
target_nodes  <- unique(chord_df$to)
n_targets_ch  <- length(target_nodes)
target_cols   <- setNames(
  grey.colors(n_targets_ch, start = 0.4, end = 0.85),
  target_nodes
)

all_cols <- c(origin_cols, target_cols)
```

```{r circlize-plot, fig.width=9, fig.height=9}
circos.clear()

circos.par(
  start.degree        = 90,
  gap.after           = c(rep(2, length(origin_cols) - 1), 8,
                           rep(2, n_targets_ch - 1), 8),
  track.margin        = c(0.01, 0.01),
  points.overflow.warning = FALSE
)

chordDiagram(
  chord_df,
  grid.col            = all_cols,
  transparency        = 0.35,
  annotationTrack     = "grid",
  preAllocateTracks   = list(track.height = 0.08),
  directional         = 1,
  direction.type      = c("diffHeight", "arrows"),
  link.arr.type       = "big.arrow"
)

# Add sector labels
circos.trackPlotRegion(
  track.index = 1,
  bg.border   = NA,
  panel.fun   = function(x, y) {
    xlim <- get.cell.meta.data("xlim")
    ylim <- get.cell.meta.data("ylim")
    sector <- get.cell.meta.data("sector.index")
    circos.text(
      mean(xlim), ylim[1],
      labels    = sector,
      facing    = "clockwise",
      niceFacing = TRUE,
      adj       = c(0, 0.5),
      cex       = 0.7,
      font      = 2
    )
  }
)

title(
  main = "Migratory Connectivity — Breeding vs Wintering Populations",
  cex.main = 1.1
)

circos.clear()
```

```{r save-circlize, eval=FALSE}
# Save high-resolution version for publication
png("outputs/Figure_Circlize_MC.png",
    width = 180, height = 180, units = "mm", res = 600)

circos.clear()
circos.par(start.degree = 90,
           gap.after    = c(rep(2, length(origin_cols) - 1), 8,
                            rep(2, n_targets_ch - 1), 8),
           points.overflow.warning = FALSE)

chordDiagram(
  chord_df,
  grid.col          = all_cols,
  transparency      = 0.35,
  annotationTrack   = "grid",
  preAllocateTracks = list(track.height = 0.08),
  directional       = 1,
  direction.type    = c("diffHeight", "arrows"),
  link.arr.type     = "big.arrow"
)

circos.trackPlotRegion(
  track.index = 1,
  bg.border   = NA,
  panel.fun   = function(x, y) {
    xlim   <- get.cell.meta.data("xlim")
    ylim   <- get.cell.meta.data("ylim")
    sector <- get.cell.meta.data("sector.index")
    circos.text(mean(xlim), ylim[1], labels = sector,
                facing = "clockwise", niceFacing = TRUE,
                adj = c(0, 0.5), cex = 0.7, font = 2)
  }
)

title(main = "Migratory Connectivity — Breeding vs Wintering Populations")
circos.clear()
dev.off()

cat("Circlize figure saved: outputs/Figure_Circlize_MC.png\n")
```

---

# Session Information

```{r session-info}
sessionInfo()
```
