---
title: "Migratory Connectivity Analysis"
author: "Mario Gómez Álvarez"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    theme: minimal
    number_sections: true
    fig_width: 12
    fig_height: 8
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  warning   = FALSE,
  message   = FALSE,
  fig.align = 'center',
  dpi       = 300
)
Sys.setenv(TZ = 'GMT')
```

## Introduction

This document estimates the strength of **migratory connectivity** (MC) for Mediterranean
shearwaters at three hierarchical levels:

1. **Taxon level** — Balearic (*Puffinus mauretanicus*, PUFMAU) vs. Yelkouan
   (*Puffinus yelkouan*, PUFYEL) shearwaters.
2. **Population level** — Four biogeographic groups (WesternMost, Central-West,
   Central-East, EasternMost).
3. **Sub-population level** — Individual breeding colonies within each group.

MC quantifies whether individuals that breed close together also winter close together
(strong connectivity, MC ≈ 1) or mix randomly on the wintering grounds (weak connectivity,
MC ≈ 0). Estimates use light-level geolocator data. Individual wintering locations and
breeding colony polygons are passed directly to `estMC()`, which handles transition
probability estimation and distance calculations internally.

---

# Setup

## Packages

```{r packages}
packages <- c(
  "tidyverse", "sf",
  "MigConnectivity",
  "rnaturalearth", "rnaturalearthdata",
  "circlize", "knitr", "kableExtra"
)

invisible(lapply(packages, library, character.only = TRUE))
```

## Coordinate Reference Systems

```{r projections}
# Geographic WGS84 — used for input coordinates and site polygons
CRS_GEO <- 4326

# Equal-area projection centred on the Mediterranean — used for buffering colonies
CRS_AEQD <- "+proj=aeqd +lat_0=38 +lon_0=15 +datum=WGS84 +units=m +no_defs"

# Mollweide equal-area — alternative projection for global views
CRS_MOLL <- "+proj=moll +lon_0=0 +datum=WGS84 +units=m +no_defs"
```

## Species and Population Definitions

```{r definitions}
# Taxon membership
PUFMAU_pops <- c("IBIZA", "MALLORCA")
PUFYEL_pops <- c("MINORCA", "FRANCE", "SARDINIA", "MALTA", "CROATIA", "GREECE")

# Biogeographic groups
group_membership <- list(
  WesternMost = c("IBIZA", "MALLORCA"),
  CentralWest = c("MINORCA", "FRANCE", "SARDINIA"),
  CentralEast = c("MALTA", "CROATIA"),
  EasternMost = c("GREECE")
)

# Canonical population order (west → east)
pop_order <- c("IBIZA", "MALLORCA", "MINORCA", "FRANCE",
               "SARDINIA", "MALTA", "CROATIA", "GREECE")

# Display colours per group (used in multiple figures)
group_colours <- c(
  WesternMost = "#E69F00",
  CentralWest = "#56B4E9",
  CentralEast = "#009E73",
  EasternMost = "#CC79A7"
)
```

---

# Geolocator Error & Uncertainty Estimation {#uncertainty}

Light-level geolocators have known positional error. We characterise this uncertainty
using the covariance matrix of repeated calibration fixes and propagate it through
the analysis via Monte Carlo resampling inside `estMC()`.

## Calibration Error Parameters

GLS positional error is estimated empirically by comparing device-estimated
positions (from `all_trips`) against the known colony coordinates during the
stationary period when each migrant was at the breeding site — i.e. **before**
departure (`BA_dep_date`) and, where available, **after** return (`BA_arr_date`).
Only individuals classified as migrants in `phen_data` are used.

```{r gls-error}
# Load phenological data; retain migrants with known departure dates
phen_migrants <- read.csv("Phen_data.csv") %>%
  filter(Migration == "Migrant", !is.na(BA_dep_date)) %>%
  mutate(
    BA_dep_dt = as.POSIXct(BA_dep_date, format = "%Y-%m-%dT%H:%M:%SZ", tz = "UTC"),
    BA_arr_dt = as.POSIXct(BA_arr_date, format = "%Y-%m-%dT%H:%M:%SZ", tz = "UTC")
  ) %>%
  select(geo_trip, BA_dep_dt, BA_arr_dt)

# Load raw tracking positions and parse datetime
all_trips_raw <- read.csv("all_trips.csv") %>%
  mutate(Date_Time = as.POSIXct(Date_Time, tz = "UTC"))

# True colony coordinates
colony_coords <- read.csv("colonies_coords.csv") %>%
  select(Colony, colony_lon = Lon, colony_lat = Lat)

# Keep only calibration-period positions:
#   - BEFORE departure from the breeding area (bird still at colony), OR
#   - AFTER return to the breeding area (bird back at colony).
#   Positions recorded during the actual migration trip are excluded because
#   the bird is no longer at a known location.
calib_pts <- all_trips_raw %>%
  inner_join(phen_migrants, by = "geo_trip") %>%
  filter(
    Date_Time < BA_dep_dt |
    (!is.na(BA_arr_dt) & Date_Time > BA_arr_dt)
  ) %>%
  left_join(colony_coords, by = "Colony") %>%
  filter(!is.na(colony_lon)) %>%
  mutate(
    res_lon = Lon - colony_lon,
    res_lat = Lat - colony_lat
  )

cat(sprintf("Calibration positions: %d from %d individuals\n",
            nrow(calib_pts), n_distinct(calib_pts$geo_trip)))

# GLS bias: mean residual per axis
gls_bias <- c(
  lon = mean(calib_pts$res_lon, na.rm = TRUE),
  lat = mean(calib_pts$res_lat, na.rm = TRUE)
)

# 2×2 covariance matrix of the bivariate GLS error
gls_cov <- cov(calib_pts[, c("res_lon", "res_lat")])
dimnames(gls_cov) <- list(c("lon", "lat"), c("lon", "lat"))

gls_sd_lon <- sqrt(gls_cov["lon", "lon"])
gls_sd_lat <- sqrt(gls_cov["lat", "lat"])
gls_cor    <- gls_cov["lon", "lat"] / (gls_sd_lon * gls_sd_lat)

cat("GLS bias (lon, lat):", round(gls_bias, 3), "\n")
cat(sprintf("GLS SD  lon: %.3f°   lat: %.3f°   correlation: %.3f\n",
            gls_sd_lon, gls_sd_lat, gls_cor))
cat("GLS covariance matrix:\n")
print(round(gls_cov, 4))
```

## Number of Monte Carlo Resamples

```{r resamp-param}
# Number of bootstrap resamples for uncertainty quantification.
# 1000 is recommended for publication (runtime ~2–5 min per estMC() call
# depending on sample size). Use 100 for rapid exploration (~10–30 sec).
N_RESAMP <- 1000
```

---

# Helper Functions {#functions}

Three reusable functions encapsulate the data-preparation steps that recur at
every analysis level.

## `relocate_land_sites_to_sea()`

```{r fn-relocate}
#' Relocate terrestrial points to the nearest coastline
#'
#' Individual wintering centroids that fall on land are snapped to the
#' nearest point on the coast. Detection uses a world land polygon from
#' `rnaturalearth`; no raster or bathymetry download is required.
#'
#' @param pts      `data.frame` with columns \code{lon} and \code{lat}.
#' @param land_sf  `sf` POLYGON/MULTIPOLYGON of land areas. If `NULL`,
#'   loads medium-resolution countries from \pkg{rnaturalearth}.
#' @param crs_geo  Integer EPSG code for geographic CRS (default 4326).
#'
#' @return `data.frame` identical to \code{pts} with corrected lon/lat.
#'
#' @examples
#' \dontrun{
#'   pts_sea <- relocate_land_sites_to_sea(pts_raw)
#' }
relocate_land_sites_to_sea <- function(pts, land_sf = NULL, crs_geo = CRS_GEO) {

  stopifnot(all(c("lon", "lat") %in% names(pts)))

  # Load world land polygon if not supplied
  if (is.null(land_sf)) {
    land_sf <- ne_countries(scale = "medium", returnclass = "sf") %>%
      st_union()
  }

  pts_sf  <- st_as_sf(pts, coords = c("lon", "lat"), crs = crs_geo, remove = FALSE)
  on_land <- as.logical(st_within(pts_sf, land_sf, sparse = FALSE))

  if (!any(on_land)) {
    message("No land points detected — coordinates unchanged.")
    return(pts)
  }

  message(sprintf("Snapping %d land point(s) to nearest coast.", sum(on_land)))

  coast <- st_cast(st_boundary(land_sf), "MULTILINESTRING")

  for (i in which(on_land)) {
    seg    <- st_nearest_points(pts_sf[i, ], coast)
    new_pt <- st_cast(seg, "POINT")
    coords <- st_coordinates(new_pt)
    # st_nearest_points returns a LINESTRING: row 1 = source, row 2 = target
    pts$lon[i] <- coords[2, 1]
    pts$lat[i] <- coords[2, 2]
  }

  pts
}
```

## `prepare_analysis_level()`

```{r fn-prepare}
#' Prepare individual-level origin and target data for `estMC()`
#'
#' Filters tracking records to the requested populations, optionally
#' corrects any wintering centroids that fall on land, and returns `sf`
#' POINT objects for `originPoints` and `targetPoints` ready for direct
#' use in \code{\link[MigConnectivity]{estMC}}.
#'
#' `estMC()` assigns each point to the appropriate site polygon and
#' estimates transition probabilities (psi) and distances internally —
#' no psi matrix or distance matrix needs to be pre-computed.
#'
#' @param tracking_df  `data.frame` with columns \code{population},
#'   \code{lon_win}, \code{lat_win}, \code{lon_col}, \code{lat_col}.
#' @param pops_include Character vector of population codes to retain.
#' @param correct_land Logical; relocate wintering centroids on land?
#'   (default `TRUE`).
#' @param land_sf      `sf` land polygon passed to
#'   \code{relocate_land_sites_to_sea()} (optional).
#'
#' @return Named list:
#'   \describe{
#'     \item{originPoints}{`sf` POINT — one per bird at its breeding colony.}
#'     \item{targetPoints}{`sf` POINT — one per bird at its wintering centroid.}
#'     \item{isGL}{Logical vector — TRUE for every geolocator-tracked bird.}
#'     \item{n}{Number of individuals.}
#'     \item{tracking}{Filtered (and corrected) tracking data frame.}
#'   }
prepare_analysis_level <- function(tracking_df, pops_include,
                                   correct_land = TRUE, land_sf = NULL) {

  df <- tracking_df %>%
    filter(population %in% pops_include)

  if (nrow(df) == 0L)
    stop("No tracking records match the requested populations.")

  # Optionally snap wintering centroids that fell on land to the coast
  if (correct_land) {
    win_pts       <- df %>% select(lon = lon_win, lat = lat_win)
    win_pts       <- relocate_land_sites_to_sea(win_pts, land_sf)
    df$lon_win    <- win_pts$lon
    df$lat_win    <- win_pts$lat
  }

  # originPoints: breeding colony location for each individual
  originPoints <- st_as_sf(df, coords = c("lon_col", "lat_col"), crs = CRS_GEO)

  # targetPoints: individual wintering centroid for each individual
  targetPoints <- st_as_sf(df, coords = c("lon_win", "lat_win"), crs = CRS_GEO)

  # All tracks are geolocator-based
  isGL <- rep(TRUE, nrow(df))

  list(
    originPoints = originPoints,
    targetPoints = targetPoints,
    isGL         = isGL,
    n            = nrow(df),
    tracking     = df
  )
}
```

## `assign_to_zones()`

```{r fn-assign}
#' Assign individual wintering points to target zones via spatial join
#'
#' Performs a `st_within()` join of individual `targetPoints` against
#' `targetSites` polygons. Issues a warning if any individuals fall
#' outside all defined zones (which may indicate that zone boundaries
#' need adjustment).
#'
#' @param target_pts `sf` POINT object — individual wintering centroids.
#' @param zones      `sf` POLYGON object — target zone polygons with a
#'   \code{site_id} column.
#' @param pop_vec    Character vector of origin population labels,
#'   parallel to \code{target_pts}.
#'
#' @return `data.frame` with columns \code{population} and \code{site_id}
#'   for matched individuals only (unmatched rows are silently excluded
#'   after the warning).
assign_to_zones <- function(target_pts, zones, pop_vec) {

  result <- st_join(
    target_pts %>% mutate(population = pop_vec),
    zones,
    join = st_within
  ) %>%
    st_drop_geometry()

  n_outside <- sum(is.na(result$site_id))
  if (n_outside > 0L) {
    warning(sprintf(
      paste0("%d individual(s) (%.1f%%) did not fall within any defined target zone ",
             "and were excluded from the chord diagram. ",
             "Consider expanding zone boundaries if this number is large."),
      n_outside,
      100 * n_outside / nrow(result)
    ))
  }

  result %>% filter(!is.na(site_id))
}
```

---

# Data Loading {#data}

```{r load-data}
# Breeding colony coordinates
colonies <- read.csv("colonies_coords.csv") %>%
  rename(site_id = Pop, lon = Lon, lat = Lat) %>%
  mutate(
    site_id = toupper(site_id),
    group = case_when(
      site_id %in% group_membership$WesternMost ~ "WesternMost",
      site_id %in% group_membership$CentralWest ~ "CentralWest",
      site_id %in% group_membership$CentralEast ~ "CentralEast",
      site_id %in% group_membership$EasternMost ~ "EasternMost"
    ),
    taxon = if_else(site_id %in% PUFMAU_pops, "PUFMAU", "PUFYEL")
  )

# Migrant geo_trip IDs from phen_data — connectivity analysis uses migrants only
migrant_ids <- read.csv("Phen_data.csv") %>%
  filter(Migration == "Migrant") %>%
  pull(geo_trip)

# Geolocator tracking data — individual wintering centroids
# Computed as the mean position during the core non-breeding months (Nov–Feb)
# Restricted to migrant individuals identified via phen_data
tracking <- read.csv("all_trips.csv") %>%
  filter(geo_trip %in% migrant_ids,          # migrants only (most selective first)
         !is.na(Lon), !is.na(Lat),
         month %in% c(11, 12, 1, 2)) %>%
  group_by(geo_trip, Population) %>%
  summarise(
    lon_win = mean(Lon, na.rm = TRUE),
    lat_win = mean(Lat, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  rename(ring_id = geo_trip, population = Population) %>%
  left_join(
    colonies %>% select(site_id, lon_col = lon, lat_col = lat),
    by = c("population" = "site_id")
  ) %>%
  mutate(
    taxon = if_else(population %in% PUFMAU_pops, "PUFMAU", "PUFYEL"),
    group = case_when(
      population %in% group_membership$WesternMost ~ "WesternMost",
      population %in% group_membership$CentralWest ~ "CentralWest",
      population %in% group_membership$CentralEast ~ "CentralEast",
      population %in% group_membership$EasternMost ~ "EasternMost"
    )
  )

cat(sprintf("Tracking records: %d migrants across %d populations\n",
            nrow(tracking), n_distinct(tracking$population)))
```

---

# Site Definitions {#sites}

`estMC()` requires geographic polygons that partition the breeding range into
**origin sites** and the non-breeding range into **target sites**. Assignments
of individuals to sites — and the resulting psi matrix — are computed internally
by `estMC()`.

## Origin Sites

Each breeding colony is represented by a circular buffer of radius 50 km,
giving one origin polygon per population. Group- and taxon-level polygons
are formed by dissolving individual colony buffers.

```{r origin-sites}
# Convert colony table to sf and project for buffering
colonies_sf <- st_as_sf(colonies, coords = c("lon", "lat"), crs = CRS_GEO)
colonies_proj <- st_transform(colonies_sf, CRS_AEQD)

# Individual colony buffers (50 km radius)
originSites_pop <- colonies_proj %>%
  mutate(geometry = st_buffer(geometry, dist = 50000)) %>%
  st_transform(CRS_GEO) %>%
  select(site_id, group, taxon)

# Group-level origin polygons (union of colony buffers within each group)
originSites_group <- originSites_pop %>%
  group_by(group) %>%
  summarise(geometry = st_union(geometry), .groups = "drop") %>%
  rename(site_id = group)

# Taxon-level origin polygons (union of colony buffers within each taxon)
originSites_taxon <- originSites_pop %>%
  group_by(taxon) %>%
  summarise(geometry = st_union(geometry), .groups = "drop") %>%
  rename(site_id = taxon)

cat("Origin sites defined:\n")
cat(" Colony level:", nrow(originSites_pop),    "polygons\n")
cat(" Group level: ", nrow(originSites_group), "polygons\n")
cat(" Taxon level: ", nrow(originSites_taxon), "polygons\n")
```

## Target Sites

Non-breeding (wintering) target regions correspond directly to the **WA** (wintering
area) categories recorded in `phen_data` for the original study. The eight named
regions are: Alboran Sea, Western Mediterranean, Central Mediterranean, Adriatic Sea,
Aegean Sea, Black Sea, Saharan Upwelling, and South European Atlantic Shelf.

```{r target-sites}
# Helper to build a rectangular sf polygon from bounding coordinates
make_rect <- function(xmin, xmax, ymin, ymax, crs = CRS_GEO) {
  st_as_sfc(
    st_bbox(c(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
            crs = crs)
  )
}

# Target zones match the WA column in phen_data (original study regions).
# All coordinates are in decimal degrees (WGS84).
targetSites <- tibble::tribble(
  ~site_id,                          ~xmin,  ~xmax, ~ymin, ~ymax,
  # Atlantic shelf off the Iberian Peninsula and Bay of Biscay
  "South European Atlantic Shelf",  -15.0,    0.0,  35.0,  52.0,
  # Upwelling zone off NW Africa (Mauritania – Morocco)
  "Saharan Upwelling",              -20.0,   -8.0,  20.0,  34.0,
  # Narrow sea between Iberia and NW Africa, west of 1.5°E
  "Alboran Sea",                     -5.5,    1.5,  34.0,  37.0,
  # Gulf of Lion, Ligurian and Balearic seas
  "Western Mediterranean",            1.5,   15.0,  36.0,  44.5,
  # Tyrrhenian and northern Ionian seas
  "Central Mediterranean",           10.0,   18.5,  33.0,  39.5,
  # Semi-enclosed sea between the Italian and Balkan peninsulas
  "Adriatic Sea",                    12.5,   20.5,  39.0,  46.5,
  # Sea between mainland Greece and Anatolia
  "Aegean Sea",                      21.5,   28.5,  35.0,  42.0,
  # Enclosed sea north of Anatolia
  "Black Sea",                       27.5,   42.0,  40.5,  47.0
) %>%
  rowwise() %>%
  mutate(geometry = make_rect(xmin, xmax, ymin, ymax)) %>%
  ungroup() %>%
  st_as_sf(crs = CRS_GEO) %>%
  select(site_id, geometry)

cat("Target sites defined:", nrow(targetSites), "wintering zones\n")
print(targetSites$site_id)
```

---

# Taxon-Level Analysis {#taxon}

We first estimate MC at the broadest taxonomic level, contrasting the two
Mediterranean shearwater species. Origin sites are the merged colony polygons
for each taxon; target sites are the wintering zones defined above.

## Prepare Data

```{r taxon-data}
dat_taxon <- prepare_analysis_level(
  tracking_df  = tracking,
  pops_include = pop_order    # all 8 populations
)

cat(sprintf("Taxon-level analysis: %d individuals total\n", dat_taxon$n))
cat("Sample sizes by taxon:\n")
print(table(dat_taxon$tracking$taxon))
```

## Estimate MC — Taxon Level

```{r taxon-estMC}
mc_taxon <- estMC(
  isGL         = dat_taxon$isGL,
  geoBias      = gls_bias,
  geoVCov      = gls_cov,
  originPoints = dat_taxon$originPoints,
  targetPoints = dat_taxon$targetPoints,
  originSites  = originSites_taxon,
  targetSites  = targetSites,
  nSim         = N_RESAMP,
  alpha        = 0.05,
  verbose      = FALSE
)

cat("=== Taxon-level MC estimate ===\n")
print(mc_taxon)
```

### Summary

```{r taxon-summary}
mc_taxon_ci <- quantile(mc_taxon$simulations, c(0.025, 0.975), na.rm = TRUE)

taxon_summary <- data.frame(
  Level   = "Taxon (PUFMAU vs PUFYEL)",
  MC_mean = round(mc_taxon$MC,    3),
  MC_lcl  = round(mc_taxon_ci[1], 3),
  MC_ucl  = round(mc_taxon_ci[2], 3),
  n       = dat_taxon$n
)

knitr::kable(taxon_summary,
             col.names = c("Level", "MC", "2.5%", "97.5%", "n"),
             caption   = "MC estimate — taxon level")
```

---

# Population-Level Analysis {#populations}

We next estimate MC across the four biogeographic groups, using group-level
dissolved origin polygons.

## Prepare Data

```{r pop-data}
dat_pop <- prepare_analysis_level(
  tracking_df  = tracking,
  pops_include = pop_order
)

cat(sprintf("Population-level analysis: %d individuals\n", dat_pop$n))
cat("Sample sizes by group:\n")
print(table(dat_pop$tracking$group))
```

## Estimate MC — Population Level

```{r pop-estMC}
mc_pop <- estMC(
  isGL         = dat_pop$isGL,
  geoBias      = gls_bias,
  geoVCov      = gls_cov,
  originPoints = dat_pop$originPoints,
  targetPoints = dat_pop$targetPoints,
  originSites  = originSites_group,
  targetSites  = targetSites,
  nSim         = N_RESAMP,
  alpha        = 0.05,
  verbose      = FALSE
)

cat("=== Population-level MC estimate ===\n")
print(mc_pop)
```

### Summary

```{r pop-summary}
mc_pop_ci <- quantile(mc_pop$simulations, c(0.025, 0.975), na.rm = TRUE)

pop_summary <- data.frame(
  Level   = "Population (4 biogeographic groups)",
  MC_mean = round(mc_pop$MC,    3),
  MC_lcl  = round(mc_pop_ci[1], 3),
  MC_ucl  = round(mc_pop_ci[2], 3),
  n       = dat_pop$n
)

knitr::kable(pop_summary,
             col.names = c("Level", "MC", "2.5%", "97.5%", "n"),
             caption   = "MC estimate — population level")
```

---

# Sub-Population Analyses {#subpopulations}

Finally, we estimate MC *within* each biogeographic group, treating individual
breeding colonies as origin nodes. This reveals whether connectivity is driven
by among-group or within-group structure.

## WesternMost (IBIZA & MALLORCA) {#western}

```{r wm-prep}
dat_WM <- prepare_analysis_level(
  tracking_df  = tracking,
  pops_include = group_membership$WesternMost
)

originSites_WM <- originSites_pop %>%
  filter(site_id %in% group_membership$WesternMost)

cat(sprintf("WesternMost: %d individuals\n", dat_WM$n))
```

```{r wm-estMC}
mc_WM <- estMC(
  isGL         = dat_WM$isGL,
  geoBias      = gls_bias,
  geoVCov      = gls_cov,
  originPoints = dat_WM$originPoints,
  targetPoints = dat_WM$targetPoints,
  originSites  = originSites_WM,
  targetSites  = targetSites,
  nSim         = N_RESAMP,
  alpha        = 0.05,
  verbose      = FALSE
)

cat("=== WesternMost sub-population MC ===\n")
print(mc_WM)
```

## Central-West (MINORCA, FRANCE & SARDINIA) {#centralwest}

```{r cw-prep}
dat_CW <- prepare_analysis_level(
  tracking_df  = tracking,
  pops_include = group_membership$CentralWest
)

originSites_CW <- originSites_pop %>%
  filter(site_id %in% group_membership$CentralWest)

cat(sprintf("Central-West: %d individuals\n", dat_CW$n))
```

```{r cw-estMC}
mc_CW <- estMC(
  isGL         = dat_CW$isGL,
  geoBias      = gls_bias,
  geoVCov      = gls_cov,
  originPoints = dat_CW$originPoints,
  targetPoints = dat_CW$targetPoints,
  originSites  = originSites_CW,
  targetSites  = targetSites,
  nSim         = N_RESAMP,
  alpha        = 0.05,
  verbose      = FALSE
)

cat("=== Central-West sub-population MC ===\n")
print(mc_CW)
```

## Central-East (MALTA & CROATIA) {#centraleast}

```{r ce-prep}
dat_CE <- prepare_analysis_level(
  tracking_df  = tracking,
  pops_include = group_membership$CentralEast
)

originSites_CE <- originSites_pop %>%
  filter(site_id %in% group_membership$CentralEast)

cat(sprintf("Central-East: %d individuals\n", dat_CE$n))
```

```{r ce-estMC}
mc_CE <- estMC(
  isGL         = dat_CE$isGL,
  geoBias      = gls_bias,
  geoVCov      = gls_cov,
  originPoints = dat_CE$originPoints,
  targetPoints = dat_CE$targetPoints,
  originSites  = originSites_CE,
  targetSites  = targetSites,
  nSim         = N_RESAMP,
  alpha        = 0.05,
  verbose      = FALSE
)

cat("=== Central-East sub-population MC ===\n")
print(mc_CE)
```

## EasternMost (GREECE) {#eastern}

With only one breeding colony, within-group MC is not estimable (pairwise
origin distances are undefined). We report the transition probability vector
to the wintering zones instead.

```{r em-prep}
dat_EM <- prepare_analysis_level(
  tracking_df  = tracking,
  pops_include = group_membership$EasternMost
)

originSites_EM <- originSites_pop %>%
  filter(site_id %in% group_membership$EasternMost)

cat(sprintf("EasternMost (GREECE): %d individuals\n", dat_EM$n))

# Derive target-zone assignment via spatial join
em_assignments <- assign_to_zones(
  dat_EM$targetPoints,
  targetSites,
  pop_vec = dat_EM$tracking$population
)

cat("Wintering zone usage (n individuals):\n")
print(table(em_assignments$site_id))
```

---

# Combined Results Summary {#summary}

```{r results-table}
# CI helper
mc_ci <- function(mc_obj) quantile(mc_obj$simulations, c(0.025, 0.975), na.rm = TRUE)

results_df <- rbind(
  taxon_summary,
  pop_summary,
  data.frame(
    Level   = c("Sub-pop: WesternMost",
                "Sub-pop: Central-West",
                "Sub-pop: Central-East"),
    MC_mean = round(c(mc_WM$MC, mc_CW$MC, mc_CE$MC), 3),
    MC_lcl  = round(c(mc_ci(mc_WM)[1], mc_ci(mc_CW)[1], mc_ci(mc_CE)[1]), 3),
    MC_ucl  = round(c(mc_ci(mc_WM)[2], mc_ci(mc_CW)[2], mc_ci(mc_CE)[2]), 3),
    n       = c(dat_WM$n, dat_CW$n, dat_CE$n)
  )
)

knitr::kable(
  results_df,
  col.names = c("Analysis level", "MC", "2.5% CI", "97.5% CI", "n"),
  caption   = "Migratory connectivity estimates across hierarchical levels"
)
```

---

# Visualization: Circlize Chord Diagram {#circlize}

A chord diagram shows the proportion of tracked individuals from each
breeding population that winter in each target zone. Chord width is
proportional to the number of individuals.

```{r circlize-data}
# Assign each individual's wintering centroid to a target zone
assignments <- assign_to_zones(
  dat_taxon$targetPoints,
  targetSites,
  pop_vec = dat_taxon$tracking$population
)

# Build flow data frame: origin population → wintering zone
chord_df <- assignments %>%
  count(population, site_id) %>%
  rename(from = population, to = site_id, value = n) %>%
  filter(value > 0)

# Assign group-level colours to breeding populations
pop_group_lookup <- setNames(colonies$group, colonies$site_id)
origin_cols <- group_colours[pop_group_lookup[unique(chord_df$from)]]
names(origin_cols) <- unique(chord_df$from)

# Neutral grey palette for wintering zones
target_nodes <- unique(chord_df$to)
n_targets_ch <- length(target_nodes)
target_cols  <- setNames(
  grey.colors(n_targets_ch, start = 0.4, end = 0.85),
  target_nodes
)

all_cols <- c(origin_cols, target_cols)

# Chord diagram gap sizes:
# GAP_SMALL — narrow gap between adjacent sectors within the same category
# GAP_LARGE — wide gap separating the breeding sector block from the wintering sector block
GAP_SMALL <- 2
GAP_LARGE <- 8
```

```{r circlize-plot, fig.width=9, fig.height=9}
circos.clear()

circos.par(
  start.degree            = 90,
  gap.after               = c(rep(GAP_SMALL, length(origin_cols) - 1), GAP_LARGE,
                               rep(GAP_SMALL, n_targets_ch - 1),       GAP_LARGE),
  track.margin            = c(0.01, 0.01),
  points.overflow.warning = FALSE
)

chordDiagram(
  chord_df,
  grid.col          = all_cols,
  transparency      = 0.35,
  annotationTrack   = "grid",
  preAllocateTracks = list(track.height = 0.08),
  directional       = 1,
  direction.type    = c("diffHeight", "arrows"),
  link.arr.type     = "big.arrow"
)

# Add sector labels
circos.trackPlotRegion(
  track.index = 1,
  bg.border   = NA,
  panel.fun   = function(x, y) {
    xlim   <- get.cell.meta.data("xlim")
    ylim   <- get.cell.meta.data("ylim")
    sector <- get.cell.meta.data("sector.index")
    circos.text(
      mean(xlim), ylim[1],
      labels     = sector,
      facing     = "clockwise",
      niceFacing = TRUE,
      adj        = c(0, 0.5),
      cex        = 0.7,
      font       = 2
    )
  }
)

title(
  main     = "Migratory Connectivity — Breeding vs Wintering Zones",
  cex.main = 1.1
)

circos.clear()
```

```{r save-circlize, eval=FALSE}
# eval=FALSE: this chunk is skipped during routine knitting to avoid writing
# output files. Change to eval=TRUE (or run manually) when generating
# publication-ready figures.
png("outputs/Figure_Circlize_MC.png",
    width = 180, height = 180, units = "mm", res = 600)

circos.clear()
circos.par(
  start.degree            = 90,
  gap.after               = c(rep(GAP_SMALL, length(origin_cols) - 1), GAP_LARGE,
                               rep(GAP_SMALL, n_targets_ch - 1),       GAP_LARGE),
  points.overflow.warning = FALSE
)

chordDiagram(
  chord_df,
  grid.col          = all_cols,
  transparency      = 0.35,
  annotationTrack   = "grid",
  preAllocateTracks = list(track.height = 0.08),
  directional       = 1,
  direction.type    = c("diffHeight", "arrows"),
  link.arr.type     = "big.arrow"
)

circos.trackPlotRegion(
  track.index = 1,
  bg.border   = NA,
  panel.fun   = function(x, y) {
    xlim   <- get.cell.meta.data("xlim")
    ylim   <- get.cell.meta.data("ylim")
    sector <- get.cell.meta.data("sector.index")
    circos.text(mean(xlim), ylim[1], labels = sector,
                facing = "clockwise", niceFacing = TRUE,
                adj = c(0, 0.5), cex = 0.7, font = 2)
  }
)

title(main = "Migratory Connectivity — Breeding vs Wintering Zones")
circos.clear()
dev.off()

cat("Circlize figure saved: outputs/Figure_Circlize_MC.png\n")
```

---

# Session Information

```{r session-info}
sessionInfo()
```
