---
title: "Year-Round Distributions Analysis"
author: "Mario Gomez Alvarez"
date: "2026-03-24"
output: html_document
---

## Introduction

This document provides an analysis of year-round distributions focusing on spatial ecology metrics. The following sections detail the methods used and present the results in a structured format suitable for publication.

## Load Required Libraries

```{r load-libraries, echo=TRUE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(sf)
library(sp)
library(raster)
```

## Data Preparation

In this section, we load and preprocess the data necessary for analysis.

```{r data-preparation, echo=TRUE}
# Load the dataset
data <- read.csv('path/to/your/data.csv')

# Data cleaning and preparation
cleaned_data <- data %>% 
  filter(!is.na(variable_of_interest)) %>%  
  mutate(new_variable = transform_function(variable))
```

## Density Grid Computation

We compute density grids based on the cleaned data.

```{r density-computation, echo=TRUE}
density_grid <- compute_density(cleaned_data)
```

## Visualization

The following section illustrates the year-round distributions through density plots.

```{r density-visualization, echo=TRUE}
ggplot(density_grid, aes(x = longitude, y = latitude, fill = density)) + 
  geom_tile() + 
  scale_fill_viridis_c() + 
  theme_minimal() + 
  labs(title = "Year-round Distribution Density", 
       x = "Longitude", 
       y = "Latitude")
```

## Discussion

In the discussion section, we interpret the results of our analysis, addressing their implications for spatial ecology.

## Conclusion

To summarize, our findings provide insight into year-round distribution patterns which are crucial for understanding the ecological dynamics influenced by various factors.