---
title: "Year-Round Distributions of Shearwaters Across Mediterranean Populations"
author: "Mario Gómez Álvarez"
date: "2026-03-24"
output:
  html_document:
    toc: true
    toc_float: true
    theme: minimal
    number_sections: true
    fig_width: 12
    fig_height: 8
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = 'center')
Sys.setenv(TZ = 'GMT')
```

# Introduction

This analysis presents year-round distributional patterns of tracked seabirds across four biogeographical regions of the Mediterranean Sea. We integrate individual-level movement data with population-level estimates to generate spatially-explicit density grids representative of the entire populations. The approach accounts for unequal sampling effort across populations and scales individual tracks to population-level estimates.

---

# Methods

## 1. Data Preparation and Package Loading

We begin by loading all necessary R packages and establishing the working directory. The analysis integrates spatial data processing, population ecology metrics, and advanced visualization techniques.

```{r packages, message=FALSE, warning=FALSE}
# Load required packages
required_packages <- c(
  'raster', 'stars', 'dplyr', 'adehabitatHR',
  'rnaturalearth', 'sf', 'rworldmap', 'ggplot2', 'patchwork', 'scales'
)

invisible(lapply(required_packages, function(pkg) {
  if (!require(pkg, character.only = TRUE)) {
    install.packages(pkg)
    library(pkg, character.only = TRUE)
  }
}))
```

## 2. Load and Inspect Raw Data

We load the primary dataset containing GPS locations from tracked individuals and colony-level information.

```{r load-data}
# Load trip data
trips <- read.csv("Results/all_trips.csv")

# Display data structure
head(trips)
cat("Dataset dimensions:", dim(trips), "\n")
cat("Missing values per column:\n")
print(colSums(is.na(trips)))
```

## 3. Define Population Groups and Temporal Partitioning

Population groups are defined based on geographical proximity and breeding colonies. We also partition the annual cycle into four distinct temporal periods to capture seasonal variation in distribution patterns.

```{r define-groups}
# Define biogeographical population groups and seasonal periods
trips <- trips %>%
  mutate(
    Group = case_when(
      Population %in% c('IBIZA', 'MALLORCA') ~ 'Ibiza-Mallorca',
      Population %in% c('MINORCA', 'FRANCE', 'SARDINIA') ~ 'Minorca-France-Sardinia',
      Population %in% c('MALTA', 'CROATIA') ~ 'Malta-Croatia',
      Population == 'GREECE' ~ 'Greece',
      TRUE ~ NA_character_
    ),
    Month_Set = case_when(
      month %in% c(11:12, 1) ~ "November-January",
      month %in% c(2:4) ~ "February-April",
      month %in% c(5:7) ~ "May-July",
      TRUE ~ "August-October"
    )
  )

# Verify grouping success
cat("Population groups defined:\n")
print(table(trips$Group, useNA = "ifany"))
cat("\nTemporal periods defined:\n")
print(table(trips$Month_Set, useNA = "ifany"))
```

## 4. Colony Population and Size Parameters

We create a lookup table containing colony-specific information including absolute population sizes and tracked individual counts. This table is critical for weighting individual observations to population-level estimates.

```{r colony-parameters}
# Define colony-level parameters
col_sizes <- data.frame(
  Colony = c('Sa Conillera','Sa Cella','La Mola','Porquerolles','Port-Cros',
             'Colata','Majjistral','Rdum','Zaklopatica','Veli Maslovnjak',
             'Gyaros','Kappari'),
  col_size = c(250,200,300,189,93,4500,200,500,207,34,1000,1730) * 2,
  pop_size = c(530,900,300,885,885,4500,1685,1685,400,400,8500,8500) * 2,
  group_size = c(3000, 3000, (300+885+4500),(300+885+4500),(300+885+4500), 
                 (300+885+4500), (1685+400), (1685+400),(1685+400), 
                 (1685+400), 8500, 8500),
  Population = c('IBIZA','MALLORCA','MINORCA','FRANCE','FRANCE',
                 'SARDINIA','MALTA','MALTA','CROATIA','CROATIA',
                 'GREECE','GREECE')
)

# Join colony parameters to trip data
trips <- trips %>%
  left_join(col_sizes, by = c("Colony", "Population"))

# Quality check: ensure no missing population sizes
missing_pop_size <- sum(is.na(trips$pop_size))
cat("Records with missing population size estimates:", missing_pop_size, "\n")
if (missing_pop_size > 0) warning("Some records lack population size data!")
```

## 5. Spatial Reference System Setup with Land Masking

We establish a base raster covering the Mediterranean study area and create binary masks for land and sea to constrain density calculations to marine areas.

```{r spatial-setup}
# Define base raster extent and resolution
r_base <- raster(
  xmn = -15, xmx = 42,
  ymn = 28,  ymx = 50,
  res = 0.5,
  crs = CRS("+proj=longlat +datum=WGS84")
)

cat("Base raster extent: [-15°, 42°] longitude, [28°, 50°] latitude\n")
cat("Base raster resolution: 0.5° cells\n")

# Load and process coastline data
land <- ne_countries(scale = 10, returnclass = "sf") %>%
  st_transform(proj4string(r_base))

# Create land raster (1 = land, NA = sea)
r_land <- rasterize(as_Spatial(land), r_base, field = 1)

# Create sea mask (1 = sea, NA = land)
r_sea <- r_base
values(r_sea) <- 1
r_sea <- mask(r_sea, r_land, inverse = TRUE)

cat("Spatial masks created successfully\n")
```

## 6. Density Grid Calculation Function

We define a custom function that computes population-weighted density grids for any given temporal subset. The function implements a novel weighting scheme that scales individual observations to population-level estimates, accounting for differential sampling effort across populations.

```{r density-grid-function}
generar_grid_meses <- function(df_meses) {
  # Input: dataframe for a specific temporal period
  # Output: sf object with density values per raster cell
  
  # 1. Data cleaning
  df_meses <- df_meses[!is.na(df_meses$pop_size), ]
  
  # Calculate number of tracked individuals per population (this month)
  n_ind_por_pop <- df_meses %>%
    group_by(Population) %>%
    summarise(n_trackeados = n_distinct(geo_trip), .groups = "drop")
  
  # Calculate total population size (each population counted once)
  total_pop_estudio <- df_meses %>%
    distinct(Population, pop_size) %>%
    summarise(total = sum(pop_size)) %>%
    pull(total)
  
  cat("  - Total population (this period):", total_pop_estudio, "individuals\n")
  
  # 2. Initialize output raster
  r_out <- r_base
  values(r_out) <- 0
  
  # Convert to Spatial format for rasterization
  sp_meses <- df_meses
  coordinates(sp_meses) <- ~Lon + Lat
  proj4string(sp_meses) <- CRS("+proj=longlat +datum=WGS84")
  
  # 3. Process each tracked individual
  for (id in unique(df_meses$geo_trip)) {
    
    # Subset data for individual i
    df_i <- df_meses[df_meses$geo_trip == id, ]
    sp_i <- sp_meses[sp_meses$geo_trip == id, ]
    
    # Extract population parameters for individual i
    pop_actual <- unique(df_i$Population)
    tamano_pop <- unique(df_i$pop_size)
    n_en_mi_pop <- n_ind_por_pop$n_trackeados[n_ind_por_pop$Population == pop_actual]
    
    # Compute individual's utilization distribution
    # (proportion of time in each cell)
    r_i <- rasterize(sp_i, r_base, fun = "count", background = 0)
    r_i <- r_i / nrow(df_i)
    
    # Compute scaling weight: represents this individual relative to total population
    # weight = (population_size / n_tracked_from_population) / total_study_population
    w_i <- (tamano_pop / n_en_mi_pop) / total_pop_estudio
    
    # Accumulate weighted utilization distribution
    r_out <- r_out + (r_i * w_i)
  }
  
  # 4. Final processing
  # Scale back to population units (individuals per cell)
  r_out <- r_out * total_pop_estudio
  
  # Mask to marine areas only
  r_out <- mask(r_out, r_sea)
  
  # Convert to sf for ggplot compatibility
  st_as_sf(st_as_stars(r_out))
}
```

## 7. Generate Seasonal Density Grids

We apply the density grid function to each of the four temporal periods, generating population-weighted distributions for the entire annual cycle.

```{r generate-grids}
cat("Generating seasonal density grids...\n")

# November-January (post-breeding dispersal)
grid_nov_jan <- generar_grid_meses(trips[trips$Month_Set == "November-January", ])
grid_nov_jan$Month_Set <- "November-January"
cat("✓ November-January grid computed\n")

# February-April (pre-breeding migration)
grid_feb_apr <- generar_grid_meses(trips[trips$Month_Set == "February-April", ])
grid_feb_apr$Month_Set <- "February-April"
cat("✓ February-April grid computed\n")

# May-July (breeding season)
grid_may_jul <- generar_grid_meses(trips[trips$Month_Set == "May-July", ])
grid_may_jul$Month_Set <- "May-July"
cat("✓ May-July grid computed\n")

# August-October (pre-dispersal staging)
grid_aug_oct <- generar_grid_meses(trips[trips$Month_Set == "August-October", ])
grid_aug_oct$Month_Set <- "August-October"
cat("✓ August-October grid computed\n")

# Combine all grids
grid_total <- rbind(grid_nov_jan, grid_feb_apr, grid_may_jul, grid_aug_oct)

cat("\nAll seasonal grids successfully generated!\n")
```

## 8. Kernel Density Estimation (KDE) for Home Range Boundaries

We compute 50% kernel density estimates for each biogeographical group and temporal period. These serve as minimum convex polygons highlighting core use areas.

```{r kde-function}
kde_period <- function(trips_df, month_set, min_locs = 50, percent = 50) {
  # Input: trip dataframe, temporal period, minimum locations per track
  # Output: sf object with KDE polygons
  
  df <- trips_df %>%
    filter(Month_Set == month_set, !is.na(Group)) %>%
    group_by(geo_trip) %>%
    filter(n() >= min_locs) %>%
    slice_sample(n = min_locs) %>%
    ungroup()
  
  # Identify minimum number of individuals per population
  min_inds <- df %>%
    group_by(Population) %>%
    summarise(n_ind = n_distinct(geo_trip)) %>%
    pull(n_ind) %>% 
    min()
  
  # Randomly subsample to balance populations
  set.seed(123)
  selected_ids <- df %>%
    group_by(Population) %>%
    distinct(geo_trip) %>%
    slice_sample(n = min_inds) %>%
    pull(geo_trip)
  
  df <- df %>% filter(geo_trip %in% selected_ids)
  
  # Convert to SpatialPointsDataFrame
  sp_df <- SpatialPointsDataFrame(
    coords = df[, c("Lon", "Lat")],
    data = data.frame(Group = df$Group),
    proj4string = CRS("+proj=longlat +datum=WGS84")
  )
  
  # Compute KDE using reference bandwidth
  kde <- kernelUD(sp_df[, "Group"], h = 1.86, grid = 200, same4all = TRUE)
  
  # Extract isopleths (percentage home ranges)
  ver <- getverticeshr(kde, percent = percent)
  ver <- st_as_sf(ver)
  ver$Group <- ver$id
  ver
}
```

## 9. Compute KDE for Each Temporal Period

```{r compute-kde}
cat("Computing 50% kernel density estimates...\n")

# Apply KDE to each seasonal period
ver_50_P1 <- kde_period(trips, "November-January")
ver_50_P1$Month_Set <- "November-January"

ver_50_P2 <- kde_period(trips, "February-April")
ver_50_P2$Month_Set <- "February-April"

ver_50_P3 <- kde_period(trips, "May-July")
ver_50_P3$Month_Set <- "May-July"

ver_50_P4 <- kde_period(trips, "August-October")
ver_50_P4$Month_Set <- "August-October"

# Combine all KDE results
ver_50_total <- rbind(ver_50_P1, ver_50_P2, ver_50_P3, ver_50_P4)

cat("KDE computation complete!\n")
```

## 10. Data Processing for Visualization

We prepare the data for publication-quality visualization by ensuring proper factor ordering, handling missing values, and standardizing variable names.

```{r prepare-visualization}
# Load world map for base layer
world <- rnaturalearth::ne_countries(scale = "large", returnclass = "sf")
wrld <- rworldmap::getMap(resolution = "high")
sf_wrld <- st_as_sf(wrld) %>% 
  st_transform(4326)

# Ensure consistent ordering of temporal periods
grid_total$Month_Set <- factor(grid_total$Month_Set, 
  levels = c("November-January", "February-April", "May-July", "August-October"))

grid_total <- grid_total %>%
  dplyr::select(layer.1, geometry, Month_Set) %>%
  rename(layer = layer.1) %>%
  mutate(layer = case_when(
    layer == 0 ~ NA_real_,
    TRUE ~ layer
  ))

# Process KDE data
ver_50_total$Month_Set <- factor(ver_50_total$Month_Set, 
  levels = c("November-January", "February-April", "May-July", "August-October"))

ver_50_total$Group <- factor(ver_50_total$Group, 
  levels = c("Ibiza-Mallorca","Minorca-France-Sardinia","Malta-Croatia","Greece"))

# Create standardized group labels for publication
ver_50_total <- ver_50_total %>%
  mutate(Group = factor(case_when(
    Group == "Ibiza-Mallorca" ~ "Western",
    Group == "Minorca-France-Sardinia" ~ "Central-West",
    Group == "Malta-Croatia" ~ "Central-East",
    Group == "Greece" ~ "Eastern"),
    levels = c("Western","Central-West","Central-East","Eastern")))

cat("Data prepared for visualization\n")
cat("Maximum density value across all periods:", max(grid_total$layer, na.rm = TRUE), "\n")
```

## 11. Marginal Density Calculations for Histograms

We compute marginal distributions along the longitudinal axis to provide supplementary statistical context for the main distributional maps.

```{r marginal-density}
# Calculate marginal density by longitude
offset <- 0.01  # Small offset to prevent log(0)

grid_marginal <- grid_total %>%
  mutate(lon_bin = round(st_coordinates(st_centroid(.))[,1])) %>%
  st_drop_geometry() %>%
  group_by(Month_Set, lon_bin) %>%
  summarise(
    mean_val = sum(layer, na.rm = TRUE) / sum(!is.na(layer)),
    log_val = log10(mean_val + offset),
    .groups = "drop"
  ) %>%
  mutate(display_val = log_val - min(log_val, na.rm = TRUE))

cat("Marginal densities computed for visualization support\n")
```

---

# Results

## Visualization Generation

We create a multi-panel figure displaying year-round distributional patterns across four biogeographical regions. Each panel combines a density map with marginal distribution histogram.

```{r create-panel-function}
# Function to create individual seasonal panel
crear_panel <- function(mes) {
  
  df_grid <- grid_total[grid_total$Month_Set == mes, ]
  df_hist <- grid_marginal[grid_marginal$Month_Set == mes, ]
  df_ver  <- ver_50_total[ver_50_total$Month_Set == mes, ]
  
  # Upper panel: marginal density histogram
  p_top <- ggplot(df_hist, aes(x = lon_bin, y = display_val)) +
    geom_col(fill = "grey30", color = NA, width = 0.8) +
    coord_cartesian(xlim = c(-15, 42), expand = FALSE) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
    theme_void() +
    theme(plot.margin = margin(b = -2, l = 0, r = 0))
  
  # Main panel: density map with KDE overlays
  max_density <- max(grid_total$layer, na.rm = TRUE)
  
  p_bot <- ggplot() +
    geom_sf(data = sf_wrld, fill = "grey60", color = "black", size = 0.1) +
    geom_sf(data = df_grid, aes(fill = layer), color = NA, show.legend = (mes == "November-January")) +
    scale_fill_viridis_c(
      option = "magma", 
      trans = "log10", 
      name = "Estimated Birds\nper cell (log10)",
      na.value = "transparent",
      limits = c(0.1, max_density), 
      breaks = c(1, 10, 100, round(max_density)),
      labels = c("1", "10", "100", as.character(round(max_density))),
      oob = scales::squish 
    ) +
    geom_sf(data = df_ver, aes(color = Group), fill = NA, size = 0.8, 
            show.legend = (mes == "November-January")) +
    scale_color_viridis_d(option = "mako", name = "Biogeographical\nPopulation") +
    coord_sf(xlim = c(-15, 42), ylim = c(28, 50), expand = FALSE) +
    labs(subtitle = mes) +
    theme_minimal() +
    theme(
      panel.grid = element_blank(),
      axis.title = element_blank(),
      plot.subtitle = element_text(hjust = 0.5, face = "bold", margin = margin(b = 5))
    )
  
  p_top / p_bot + plot_layout(heights = c(0.15, 0.85))
}

cat("Panel function defined successfully\n")
```

## Generate All Seasonal Panels

```{r generate-panels, fig.width=14, fig.height=10}
# Create individual panels for each season
p1 <- crear_panel("November-January")
p2 <- crear_panel("February-April")
p3 <- crear_panel("May-July")
p4 <- crear_panel("August-October")

# Combine into final publication-ready figure
p_final <- (p1 | p2) / (p3 | p4) + 
  plot_layout(guides = 'collect') +
  plot_annotation(tag_levels = "A") & 
  theme(
    legend.position = 'bottom', 
    legend.box = "vertical",
    plot.tag = element_text(face = "bold", size = 14)
  )

print(p_final)

cat("✓ Figure generated successfully\n")
```

## Export Publication-Ready Figure

```{r export-figure}
# Save high-resolution figure for publication
ggsave(
  filename = "Fig3_density_grid.png",
  plot = p_final,
  width = 210,   # A4 width in mm
  height = 210,  # Proportional height
  units = "mm",
  dpi = 600,
  bg = "white"
)

cat("✓ Figure exported: Fig3_density_grid.png (600 DPI)\n")

# Also save as PDF for publication submission
ggsave(
  filename = "Fig3_density_grid.pdf",
  plot = p_final,
  width = 210,
  height = 210,
  units = "mm",
  bg = "white"
)

cat("✓ Figure exported: Fig3_density_grid.pdf\n")
```

---

# Discussion

The year-round distributional patterns presented here demonstrate clear seasonal shifts in space use across Mediterranean populations. The integration of individual tracking data with population-level demographic estimates provides a novel framework for understanding how tracked individuals represent broader population-level processes.

The density grids reveal population-specific distributions that vary substantially across the four temporal periods, reflecting the distinct ecological demands of different annual cycle phases. Core use areas identified by kernel density estimation highlight regions of consistent population presence and potentially important habitat requirements.

---

# Session Information

```{r session-info}
sessionInfo()
```
