---
title: "Migratory Connectivity of Mediterranean Shearwaters"
author: "Mario Gómez Álvarez"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    fig_width: 12
    fig_height: 8
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = 'center',
  dpi = 300
)
Sys.setenv(TZ = 'GMT')
```

## Introduction

Migratory connectivity describes the degree to which individuals from the same breeding population remain spatially associated during the non-breeding season. Here we quantify migratory connectivity for Mediterranean shearwaters (*Puffinus mauretanicus* and *P. yelkouan*) using GPS tracking data from eight breeding populations. We estimate the **MC statistic** (Cohen & Marra 2022) at three hierarchical levels: taxon, biogeographic group, and breeding population.

The MC statistic ranges from −1 (complete mixing of breeding populations on non-breeding grounds) to +1 (complete segregation), with 0 indicating random mixing. Cost-distance matrices are computed over the ocean surface — shearwaters cannot fly over land — to ensure biologically realistic estimates of spatial relationships.

---

# 1. Load Libraries and Parameters

```{r libraries}
library(raster)
library(sp)
library(sf)
library(gdistance)
library(MigConnectivity)
library(dplyr)
library(tidyr)
library(purrr)
library(rnaturalearth)
library(rworldmap)
library(adehabitatHR)
library(ggplot2)
library(circlize)

# Set working directory
setwd('C:/Users/mario/OneDrive - Universitat de Barcelona/R_code')

# Analysis parameters
NON_BREEDING_MONTHS <- c(8, 9, 10, 11, 12, 1, 2, 3)  # August–March
RASTER_RES  <- 0.1   # degrees; coarser resolution speeds up cost-distance computation
N_BOOTSTRAP <- 1000  # bootstrap iterations for MC confidence intervals
SEED        <- 123
set.seed(SEED)
```

---

# 2. Load and Prepare Data

```{r load-data}
# Load GPS tracking data
trips <- read.csv("all_trips.csv")

# Load colony coordinates and population metadata
colonies <- read.csv("colonies_coords.csv")

# Assign biogeographic groups to each GPS location
trips <- trips %>%
  mutate(
    Group = case_when(
      Population %in% c('IBIZA', 'MALLORCA')            ~ 'WesternMost',
      Population %in% c('MINORCA', 'FRANCE', 'SARDINIA') ~ 'Central-West',
      Population %in% c('MALTA', 'CROATIA')              ~ 'Central-East',
      Population == 'GREECE'                             ~ 'EasternMost',
      TRUE ~ NA_character_
    ),
    Taxon = case_when(
      Sp == 'PUFMAU'  ~ 'Balearic',
      Sp == 'PUFYEL'  ~ 'Yelkouan',
      Sp == 'HYBRID'  ~ 'Hybrid',
      TRUE ~ NA_character_
    )
  )

# Filter to non-breeding months only
trips_nb <- trips %>%
  filter(month %in% NON_BREEDING_MONTHS, !is.na(Group))

cat("GPS locations in non-breeding season:", nrow(trips_nb), "\n")
cat("Individuals tracked:", n_distinct(trips_nb$geo_trip), "\n")
```

---

# 3. Define Non-Breeding Zones

Non-breeding zones are defined as rectangular geographic regions that cover the main wintering areas of Mediterranean shearwaters: the Atlantic seaboard (from the Bay of Biscay to the Gulf of Guinea) and the entire Mediterranean basin.

```{r define-zones}
# Define non-breeding zone bounding boxes as simple features (sf)
zone_list <- list(
  Bay_of_Biscay   = c(xmin = -10, xmax =  5, ymin = 43, ymax = 50),
  NW_Africa       = c(xmin = -20, xmax =  0, ymin = 20, ymax = 43),
  Gulf_of_Guinea  = c(xmin = -20, xmax =  8, ymin =  0, ymax = 20),
  W_Mediterranean = c(xmin =  0,  xmax = 10, ymin = 36, ymax = 44),
  C_Mediterranean = c(xmin = 10,  xmax = 20, ymin = 30, ymax = 44),
  E_Mediterranean = c(xmin = 20,  xmax = 37, ymin = 30, ymax = 42)
)

# Convert each bounding box to an sf polygon
zones_sf <- map2_dfr(
  zone_list, names(zone_list),
  function(bb, nm) {
    st_as_sf(
      data.frame(zone = nm),
      geometry = st_sfc(
        st_as_sfc(st_bbox(
          c(xmin = bb["xmin"], ymin = bb["ymin"],
            xmax = bb["xmax"], ymax = bb["ymax"]),
          crs = 4326
        ))
      ),
      crs = 4326
    )
  }
)

cat("Non-breeding zones defined:", nrow(zones_sf), "\n")
print(zones_sf$zone)
```

---

# 4. Assign GPS Points to Non-Breeding Zones

```{r assign-zones}
# Convert GPS locations to spatial features
trips_sf <- st_as_sf(trips_nb, coords = c("Lon", "Lat"), crs = 4326)

# Assign each point to a zone (spatial intersection)
trips_sf <- st_join(trips_sf, zones_sf, join = st_within)

# Count points per individual and zone
zone_counts <- trips_sf %>%
  st_drop_geometry() %>%
  filter(!is.na(zone)) %>%
  group_by(geo_trip, Population, Group, Taxon, zone) %>%
  summarise(n_pts = n(), .groups = "drop")

# Keep the dominant zone for each individual (zone with most GPS fixes)
dominant_zone <- zone_counts %>%
  group_by(geo_trip, Population, Group, Taxon) %>%
  slice_max(n_pts, n = 1, with_ties = FALSE) %>%
  ungroup()

cat("Individuals with a dominant zone assigned:", nrow(dominant_zone), "\n")
print(table(dominant_zone$zone))
```

---

# 5. Define Breeding Sites

Breeding sites are represented by the centroid of each individual's tracking data during the breeding season (months outside the non-breeding window). For individuals with insufficient breeding-season fixes, the colony coordinates are used as a fallback.

```{r breeding-sites}
# Identify breeding-season locations (months not in the non-breeding filter)
trips_br <- trips %>%
  filter(!month %in% NON_BREEDING_MONTHS) %>%
  group_by(geo_trip, Population) %>%
  summarise(
    breed_lon = mean(Lon, na.rm = TRUE),
    breed_lat = mean(Lat, na.rm = TRUE),
    .groups = "drop"
  )

# For individuals without breeding-season fixes, use colony coordinates
colony_lookup <- colonies %>%
  rename(Population = Pop, breed_lon = Lon, breed_lat = Lat) %>%
  select(Population, breed_lon, breed_lat)

breeding_sites <- dominant_zone %>%
  left_join(trips_br, by = c("geo_trip", "Population")) %>%
  left_join(colony_lookup, by = "Population", suffix = c("", "_colony")) %>%
  mutate(
    breed_lon = coalesce(breed_lon, breed_lon_colony),
    breed_lat = coalesce(breed_lat, breed_lat_colony)
  ) %>%
  select(geo_trip, Population, Group, Taxon, zone, breed_lon, breed_lat)

cat("Breeding sites resolved for:", nrow(breeding_sites), "individuals\n")
```

---

# 6. Build Ocean Cost-Distance Matrix

Cost-distance matrices constrain movement paths to the ocean surface, preventing unrealistic overland routes. The raster is built at a moderate resolution to balance accuracy and computational speed.

```{r cost-distance-setup}
# Load a high-resolution land mask
world <- rworldmap::getMap(resolution = "high")
world_sf <- st_as_sf(world)

# Create a raster template covering the study extent
r_template <- raster(
  xmn = -25, xmx = 42,
  ymn = -5,  ymx = 55,
  resolution = RASTER_RES,
  crs = CRS("+proj=longlat +datum=WGS84")
)
values(r_template) <- 1  # All cells start as passable (ocean)

# Detect land cells and set their conductance to zero (impassable)
# rasterize() returns NA for water and 1 for land
land_mask <- rasterize(world_sf, r_template, field = 1, background = NA)
ocean_raster <- r_template
ocean_raster[!is.na(land_mask)] <- 0  # Land cells are blocked

cat("Raster dimensions:", dim(ocean_raster), "\n")
cat("Ocean cells:", sum(values(ocean_raster) == 1, na.rm = TRUE), "\n")
cat("Land cells blocked:", sum(values(ocean_raster) == 0, na.rm = TRUE), "\n")
```

```{r cost-distance-transition}
# Optimization: pre-compute the transition matrix once and reuse it for all
# pairwise distance calculations. This is the most time-consuming step, but
# computing it only once avoids redundant processing.
tr <- transition(ocean_raster, transitionFunction = mean, directions = 16)
tr <- geoCorrection(tr, type = "c", multpl = FALSE)

cat("Transition matrix computed successfully\n")
```

```{r cost-distance-matrix}
# Extract unique site coordinates for breeding and non-breeding locations
breed_coords <- breeding_sites %>%
  distinct(Population, breed_lon, breed_lat) %>%
  arrange(Population)

# Wintering site coordinates: centroid of each zone polygon
winter_coords <- zones_sf %>%
  st_centroid() %>%
  mutate(
    winter_lon = st_coordinates(.)[, 1],
    winter_lat = st_coordinates(.)[, 2]
  ) %>%
  st_drop_geometry() %>%
  arrange(zone)

# Convert to SpatialPoints for gdistance
breed_sp  <- SpatialPoints(
  breed_coords[, c("breed_lon", "breed_lat")],
  proj4string = CRS("+proj=longlat +datum=WGS84")
)

winter_sp <- SpatialPoints(
  winter_coords[, c("winter_lon", "winter_lat")],
  proj4string = CRS("+proj=longlat +datum=WGS84")
)

# Compute pairwise cost-distance matrix among breeding sites (square; required by estMC as originDist)
# Optimization: use costDistance() rather than looping over individual pairs
breed_dist_matrix <- costDistance(tr, breed_sp, breed_sp)
rownames(breed_dist_matrix) <- breed_coords$Population
colnames(breed_dist_matrix) <- breed_coords$Population

# Compute pairwise cost-distance matrix among non-breeding zones (square; required by estMC as targetDist)
winter_dist_matrix <- costDistance(tr, winter_sp, winter_sp)
rownames(winter_dist_matrix) <- winter_coords$zone
colnames(winter_dist_matrix) <- winter_coords$zone

cat("Breeding distance matrix dimensions:", dim(breed_dist_matrix), "\n")
cat("Winter zone distance matrix dimensions:", dim(winter_dist_matrix), "\n")
print(round(breed_dist_matrix / 1000, 0))  # Display distances in km
```

---

# 7. Estimate Migratory Connectivity

## 7a. Population-Level MC

```{r mc-population}
# Prepare the origin (breeding) and target (non-breeding) site indices
# Each individual is mapped to their breeding population and dominant winter zone
origin_index  <- match(breeding_sites$Population, breed_coords$Population)
target_index  <- match(breeding_sites$zone,       winter_coords$zone)

# Remove individuals with unresolved zone assignments
valid_idx <- !is.na(origin_index) & !is.na(target_index)
breeding_sites_valid <- breeding_sites[valid_idx, ]
origin_index_valid   <- origin_index[valid_idx]
target_index_valid   <- target_index[valid_idx]

# Sample sizes per breeding population (used as relative weights)
pop_sizes <- breeding_sites_valid %>%
  count(Population) %>%
  arrange(Population) %>%
  pull(n)

# Estimate MC at the population level with bootstrapped confidence intervals
set.seed(SEED)
mc_pop <- estMC(
  originAssignment  = origin_index_valid,
  targetAssignment  = target_index_valid,
  originDist        = breed_dist_matrix,
  targetDist        = winter_dist_matrix,
  originRelAbund    = pop_sizes / sum(pop_sizes),
  nBoot             = N_BOOTSTRAP,
  alpha             = 0.05
)

cat("MC estimate (population level):", round(mc_pop$MC, 3),
    "95% CI: [", round(mc_pop$lower, 3), ",", round(mc_pop$upper, 3), "]\n")
```

## 7b. Group-Level MC

```{r mc-group}
# Remap individuals to their biogeographic group
group_order <- c("WesternMost", "Central-West", "Central-East", "EasternMost")

group_coords <- breeding_sites_valid %>%
  distinct(Group, breed_lon, breed_lat) %>%
  group_by(Group) %>%
  summarise(
    breed_lon = mean(breed_lon),
    breed_lat = mean(breed_lat),
    .groups = "drop"
  ) %>%
  arrange(match(Group, group_order))

group_sp <- SpatialPoints(
  group_coords[, c("breed_lon", "breed_lat")],
  proj4string = CRS("+proj=longlat +datum=WGS84")
)

# Recompute cost-distance matrix for groups (groups × winter zones, not used in estMC)
# Also compute pairwise distances between groups (square; used as originDist in estMC)
cost_dist_group_sq <- costDistance(tr, group_sp, group_sp)
rownames(cost_dist_group_sq) <- group_coords$Group
colnames(cost_dist_group_sq) <- group_coords$Group

# Map each individual to their group index
origin_grp <- match(breeding_sites_valid$Group, group_coords$Group)
grp_sizes  <- breeding_sites_valid %>%
  count(Group) %>%
  arrange(match(Group, group_order)) %>%
  pull(n)

set.seed(SEED)
mc_grp <- estMC(
  originAssignment  = origin_grp,
  targetAssignment  = target_index_valid,
  originDist        = cost_dist_group_sq,
  targetDist        = winter_dist_matrix,
  originRelAbund    = grp_sizes / sum(grp_sizes),
  nBoot             = N_BOOTSTRAP,
  alpha             = 0.05
)

cat("MC estimate (group level):", round(mc_grp$MC, 3),
    "95% CI: [", round(mc_grp$lower, 3), ",", round(mc_grp$upper, 3), "]\n")
```

## 7c. Taxon-Level MC

```{r mc-taxon}
# Retain only pure-species individuals (exclude hybrids)
bs_taxon <- breeding_sites_valid %>%
  filter(Taxon %in% c("Balearic", "Yelkouan"))

taxon_coords <- bs_taxon %>%
  distinct(Taxon, breed_lon, breed_lat) %>%
  group_by(Taxon) %>%
  summarise(
    breed_lon = mean(breed_lon),
    breed_lat = mean(breed_lat),
    .groups = "drop"
  ) %>%
  arrange(Taxon)

taxon_sp <- SpatialPoints(
  taxon_coords[, c("breed_lon", "breed_lat")],
  proj4string = CRS("+proj=longlat +datum=WGS84")
)

# Pairwise distances between taxa (square; used as originDist in estMC)
cost_dist_taxon_sq <- costDistance(tr, taxon_sp, taxon_sp)
rownames(cost_dist_taxon_sq) <- taxon_coords$Taxon
colnames(cost_dist_taxon_sq) <- taxon_coords$Taxon

origin_tax <- match(bs_taxon$Taxon,  taxon_coords$Taxon)
target_tax <- match(bs_taxon$zone,   winter_coords$zone)
tax_sizes  <- bs_taxon %>%
  count(Taxon) %>%
  arrange(Taxon) %>%
  pull(n)

set.seed(SEED)
mc_tax <- estMC(
  originAssignment  = origin_tax,
  targetAssignment  = target_tax,
  originDist        = cost_dist_taxon_sq,
  targetDist        = winter_dist_matrix,
  originRelAbund    = tax_sizes / sum(tax_sizes),
  nBoot             = N_BOOTSTRAP,
  alpha             = 0.05
)

cat("MC estimate (taxon level):", round(mc_tax$MC, 3),
    "95% CI: [", round(mc_tax$lower, 3), ",", round(mc_tax$upper, 3), "]\n")
```

---

# 8. Summarise MC Results

```{r mc-summary}
# Compile all MC estimates into a single table for reporting
mc_summary <- data.frame(
  Level    = c("Taxon", "Group", "Population"),
  MC       = c(mc_tax$MC,  mc_grp$MC,  mc_pop$MC),
  CI_lower = c(mc_tax$lower, mc_grp$lower, mc_pop$lower),
  CI_upper = c(mc_tax$upper, mc_grp$upper, mc_pop$upper),
  n_ind    = c(nrow(bs_taxon), nrow(breeding_sites_valid), nrow(breeding_sites_valid))
)

mc_summary$MC       <- round(mc_summary$MC, 3)
mc_summary$CI_lower <- round(mc_summary$CI_lower, 3)
mc_summary$CI_upper <- round(mc_summary$CI_upper, 3)

print(mc_summary)
```

---

# 9. Visualisation — Chord Diagram

Chord diagrams illustrate the proportional connections between breeding populations (left arc) and non-breeding zones (right arc). Arc width on the breeding side is proportional to the number of tracked individuals; connections are coloured by the origin population.

```{r chord-diagram-groups, fig.width=10, fig.height=10}
# Prepare the connectivity matrix: rows = breeding groups, columns = winter zones
conn_mat <- table(breeding_sites_valid$Group, breeding_sites_valid$zone)
conn_mat <- conn_mat[group_order, , drop = FALSE]  # Enforce group order

# Assign a distinct colour to each breeding group
group_colors <- c(
  "WesternMost"  = "#E63946",
  "Central-West" = "#F4A261",
  "Central-East" = "#2A9D8F",
  "EasternMost"  = "#457B9D"
)

# Zone colours are lighter shades to distinguish them visually
zone_colors <- setNames(
  rep("#CCCCCC", ncol(conn_mat)),
  colnames(conn_mat)
)

all_colors <- c(group_colors, zone_colors)

# Draw the chord diagram
par(mar = c(1, 1, 3, 1))
chordDiagram(
  conn_mat,
  grid.col      = all_colors,
  transparency  = 0.3,
  annotationTrack = c("name", "grid"),
  preAllocateTracks = list(track.height = 0.05)
)
title("Migratory Connectivity — Breeding Groups to Non-Breeding Zones",
      cex.main = 1.2)
circos.clear()
```

```{r chord-diagram-populations, fig.width=10, fig.height=10}
# Repeat chord diagram at the breeding population level
conn_pop <- table(breeding_sites_valid$Population, breeding_sites_valid$zone)

pop_colors <- setNames(
  scales::hue_pal()(nrow(conn_pop)),
  rownames(conn_pop)
)

par(mar = c(1, 1, 3, 1))
chordDiagram(
  conn_pop,
  grid.col      = c(pop_colors, zone_colors),
  transparency  = 0.3,
  annotationTrack = c("name", "grid"),
  preAllocateTracks = list(track.height = 0.05)
)
title("Migratory Connectivity — Breeding Populations to Non-Breeding Zones",
      cex.main = 1.2)
circos.clear()
```

---

# 10. Export Results

```{r export}
# Save the dominant-zone assignments for downstream analyses
write.csv(
  breeding_sites_valid,
  "Results/migratory_connectivity_assignments.csv",
  row.names = FALSE
)

# Save the MC summary table
write.csv(
  mc_summary,
  "Results/MC_estimates_summary.csv",
  row.names = FALSE
)

cat("Results written to Results/\n")
```
