---
title: "Migratory Connectivity Analysis"
author: "Mario Gómez Álvarez"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    theme: minimal
    number_sections: true
    fig_width: 12
    fig_height: 8
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  warning   = FALSE,
  message   = FALSE,
  fig.align = 'center',
  dpi       = 300
)
Sys.setenv(TZ = 'GMT')
```

## Introduction

This document estimates the strength of **migratory connectivity** (MC) for Mediterranean
shearwaters at three hierarchical levels:

1. **Taxon level** — Balearic (PUFMAU) vs. Yelkouan (PUFYEL) shearwaters.
2. **Biogeographical level** — Four biogeographic units (WesternMost, Central-West,
   Central-East, EasternMost).
3. **Population level** — Individual breeding populations within each unit.

---

# Setup

## Packages

```{r packages}
packages <- c(
  "tidyverse", "sf", "gdistance",
  "MigConnectivity", "viridisLite",
  "rnaturalearth", "rnaturalearthdata",
  "circlize", "knitr", "kableExtra","RANN"
)

invisible(lapply(packages, library, character.only = TRUE))
```

## Coordinate Reference Systems

```{r projections}
CRS_GEO <- "EPSG:4326"
CRS_AEQD <- "+proj=aeqd +lat_0=38 +lon_0=15 +datum=WGS84 +units=m +no_defs"
CRS_MOLL <- "+proj=moll +lon_0=0 +datum=WGS84 +units=m +no_defs"
```

# Ocean Raster Creation

This section creates or loads the ocean raster for cost-distance calculations. The raster
defines ocean (value = 1) and land (value = 0) for the Mediterranean region.

```{r ocean-raster-creation}
# Define Mediterranean extent
xmin <- -20
xmax <- 50
ymin <- 20
ymax <- 55
resolution <- 0.05

# Create empty raster template
ocean.r <- raster(
  xmn = xmin, xmx = xmax,
  ymn = ymin, ymx = ymax,
  res = resolution,
  crs = CRS(CRS_GEO)
)

# Get land polygons from rnaturalearth
land_sf <- ne_countries(scale = "medium", returnclass = "sf") %>%
  st_union()

# Convert sf to sp for rasterization 
land_sp <- as_Spatial(land_sf)

# Rasterize land polygons: land areas will have a value, NA elsewhere
ocean.r <- rasterize(land_sp, ocean.r, field = 1)

# Convert: land = 0, ocean = 1
ocean.r[!is.na(ocean.r)] <- 0  # Land cells set to 0
ocean.r[is.na(ocean.r)] <- 1    # Ocean cells set to 1

cat("Ocean raster created:\n")
```

---

# Helper Functions 

Three reusable functions encapsulate the data-preparation steps that recur at
every analysis level.

The dirst function assigns individual tracks to the dominant WA
```{r fn-relocate}
assign_centroid_from_zones <- function(tracking_df, zones_sf, ocean_raster) {

  pts_sf <- st_as_sf(tracking_df,
                     coords = c("Lon", "Lat"),
                     crs = CRS_GEO,
                     remove = FALSE)

  # Keep ONLY ocean points
  ocean_flag <- raster::extract(ocean_raster, st_coordinates(pts_sf)) == 1
  pts_sf <- pts_sf[ocean_flag, ]

  # Spatial join only ocean-valid points
  joined <- st_join(pts_sf, zones_sf, join = st_within)

  # Keep valid site assignments
  joined <- joined %>%
    filter(!is.na(site_id))

  counts <- joined %>%
    st_drop_geometry() %>%
    group_by(geo_trip, site_id) %>%
    summarise(n = n(), .groups = "drop")

  dominant_zone <- counts %>%
    group_by(geo_trip) %>%
    slice_max(n, n = 1, with_ties = FALSE) %>%
    ungroup()

  cat("\n--- Zone assignment diagnostics ---\n")
  print(table(dominant_zone$site_id, useNA = "ifany"))
  cat("----------------------------------\n\n")

  zone_centroids <- st_centroid(zones_sf)

  zone_centroids_df <- zone_centroids %>%
    st_drop_geometry() %>%
    bind_cols(st_coordinates(zone_centroids) %>% as.data.frame())

  result <- dominant_zone %>%
    left_join(zone_centroids_df, by = "site_id") %>%
    rename(Lon = X, Lat = Y)

  return(result)
}
```

## `prepare_analysis_level()`

Prepare individual-level origin and target data for `estMC()`

```{r fn-prepare}
prepare_analysis_level <- function(tracking_df, pops_include = NULL, zones_sf) {

  df <- tracking_df 
  
  if(!is.null(pops_include)){
  df <- df %>% 
    filter(Population %in% pops_include)}

  if (nrow(df) == 0L)
    stop("No tracking records match the requested populations.")

  # Assign points to targetSites
  win_pts <- assign_centroid_from_zones(df, zones_sf, ocean.r)

  df <- df%>%
    rename(Lon_win_og = Lon, Lat_win_og = Lat) %>%
    left_join(win_pts, by = "geo_trip")
  
  # Reasses NAs using the individual centroid
  df_ok  <- df %>% filter(!is.na(Lon), !is.na(Lat))%>%
    dplyr::select(-Lon_win_og,-Lat_win_og)
  
  df_na  <- df %>% filter(is.na(Lon) | is.na(Lat))%>%
    group_by(geo_trip)%>%
    mutate(
      Lon_win_og = mean(Lon_win_og, na.rm = TRUE),
      Lat_win_og = mean(Lat_win_og, na.rm = TRUE)
      )%>%
    ungroup
  
  na_sf <- st_as_sf(df_na, coords = c("Lon_win_og", "Lat_win_og"), crs = CRS_GEO)
  
  # Assign to the nearest zone
  nearest_zone <- st_nearest_feature(na_sf, targetSites)
  df_na$site_id <- targetSites$site_id[nearest_zone]
  df_na <- df_na %>%
    mutate(Lon = Lon_win_og, Lat = Lat_win_og)%>%
    dplyr::select(-Lon_win_og,-Lat_win_og)
  
  df_fixed <- bind_rows(df_ok, df_na)%>%
    distinct()
  
  # originPoints: breeding colony location for each individual
 originPoints <- df_fixed %>%
  group_by(geo_trip, Population, Group, Sp) %>%
  summarise(
    Lon = mean(Lon_col, na.rm = TRUE),
    Lat = mean(Lat_col, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  st_as_sf(coords = c("Lon", "Lat"), crs = CRS_GEO)
  
 originAssignment <- as.integer(factor(originPoints$Population, levels = unique(originPoints$Population)))  
  
  targetPoints <- df_fixed %>%
  group_by(geo_trip, Population, Group, Sp, site_id) %>%
  summarise(
    Lon = mean(Lon, na.rm = TRUE),
    Lat = mean(Lat, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  st_as_sf(coords = c("Lon", "Lat"), crs = CRS_GEO)
  
  targetAssignment <- as.integer(factor(targetPoints$site_id, levels = unique(targetPoints$site_id))) 
  
  list(
    originPoints = originPoints,
    targetPoints = targetPoints,
    originAssignment = originAssignment,
    targetAssignment = targetAssignment,
    n            = nrow(originPoints)
    )
}
```
#' Calculate marine cost-distance matrix from origin or target sites
#'
#' Uses a raster-based cost-distance approach with gdistance package.
#' Requires an ocean raster (land = 0, ocean = 1).

``` {r cal_dist}
calculate_distance_matrix <- function(sites_sf, ocean_raster) {
  
  # Centroids of the polygons
  sites_centroids <- st_centroid(sites_sf)
  sites_centroids <- st_transform(sites_centroids, CRS_GEO)
  
  sites <- as.data.frame(st_coordinates(sites_centroids))
  colnames(sites) <- c("X", "Y")
  
  # Detect those points in land
  sites.to.relocate <- raster::extract(ocean_raster, sites) == 0
  
  if (any(sites.to.relocate)) {
    
    cat(sprintf("\n.....................................................................\n"))
    cat(sprintf("%d site(s) to be relocated from land to ocean.\n", sum(sites.to.relocate)))
    cat(sprintf(".....................................................................\n\n"))
    
    # Obtain points of the ocean raster
    ocean_cells <- rasterToPoints(ocean_raster, fun = function(x) x == 1)
    ocean_coords <- ocean_cells[, 1:2]  # X, Y
    
    land_sites <- sites[sites.to.relocate, ]
    
    # KNN → closest neighbor
    nn <- nn2(data = ocean_coords, query = land_sites, k = 1)
    
    sites[sites.to.relocate, ] <- ocean_coords[nn$nn.idx, ]
  }
  
  # Cost surface
  cost_surface <- ocean_raster
  projection(cost_surface) <- CRS(CRS_GEO)
  
  raster_tr <- transition(cost_surface, mean, directions = 8)
  raster_tr_corrected <- geoCorrection(raster_tr, type = "c", multpl = FALSE)
  
  # Cost distance
  raster_cosDist <- costDistance(
    raster_tr_corrected,
    as.matrix(sites),
    as.matrix(sites)
  )
  
  return(as.matrix(raster_cosDist))
}
```

---

# Data Loading {#data}

```{r load-data}

# True colony coordinates
colonies <- read.csv("data/colonies_coords.csv")

mean_cols <- colonies%>%
  group_by(Pop)%>%
  summarise(Lat_pop = mean(Lat),
            Lon_pop = mean(Lon))%>%
  ungroup()

colonies <- colonies%>%
  left_join(.,mean_cols, by = "Pop")

# Geolocator tracking data 

tracking <- read.csv("outputs/all_trips_classified.csv") %>%
  filter(period =="non-breeding")%>%
  left_join(
    colonies %>% dplyr::select(Colony, Lon_col = Lon, Lat_col = Lat),
    by = c("Colony")
  )%>%dplyr::select(-Date_Time,-month)


cat(sprintf("Tracking records: %d tracks across %d colonies\n",
            n_distinct(tracking$geo_trip), n_distinct(tracking$Colony)))

```
---

# Site Definitions {#sites}

`estMC()` requires geographic polygons that partition the breeding range into
**origin sites** and the non-breeding range into **target sites**. Assignments
of individuals to sites — and the resulting psi matrix — are computed internally
by `estMC()`.

## Origin Sites

Each breeding colony is represented by a circular buffer of radius 50 km,
giving one origin polygon per population. Unit- and taxon-level polygons
are formed by dissolving individual colony buffers.

```{r origin-sites-dist}
# Convert colony table to sf and project for buffering
colonies_sf <- st_as_sf(colonies, coords = c("Lon_pop", "Lat_pop"), crs = CRS_GEO)
colonies_proj <- st_transform(colonies_sf, CRS_AEQD)

# Individual colony buffers (2 km radius)
originSites_col <- colonies_proj %>%
  mutate(geometry = st_buffer(geometry, dist = 2000)) %>%
  st_transform(CRS_GEO) %>%
  dplyr::select(Pop,Group,Taxon)%>%
  distinct()

# Separate by taxa
originSites_PUFMAU <- originSites_col %>%
  filter(Taxon == "PUFMAU")
originSites_PUFYEL <- originSites_col %>%
  filter(Taxon == "PUFYEL")

# Separate by unit
originSites_WM <- originSites_col %>%
  filter(Group == "WesternMost")
originSites_CW <- originSites_col %>%
  filter(Group == "Central-West")
originSites_CE <- originSites_col %>%
  filter(Group == "Central-East")
originSites_EM <- originSites_col %>%
  filter(Group == "EasternMost")

# Unit-level origin polygons (union of colony buffers within each unit)
originSites_unit <- originSites_col %>%
  group_by(Group) %>%
  summarise(geometry = st_union(geometry), .groups = "drop") %>%
  rename(site_id = Group)

# Population-level distance
originDist_col <- calculate_distance_matrix(originSites_col, ocean.r)
colnames(originDist_col) <- rownames(originSites_col) <-  originSites_col$Pop

cat("Unit origin distance matrix:\n")
print(round(originDist_unit, 1))

# Unit-level distance
originDist_unit <- calculate_distance_matrix(originSites_unit, ocean.r)
colnames(originDist_unit) <- rownames(originDist_unit) <-  originSites_unit$site_id

cat("Unit origin distance matrix:\n")
print(round(originDist_unit, 1))

# Taxon-level origin polygons (union of colony buffers within each taxon)
originSites_taxon <- originSites_col %>%
  group_by(Taxon) %>%
  summarise(geometry = st_union(geometry), .groups = "drop") %>%
  rename(site_id = Taxon)

# Taxon-level distance
originDist_taxon <- calculate_distance_matrix(originSites_taxon, ocean.r)
colnames(originDist_taxon) <- rownames(originDist_taxon) <- originSites_taxon$site_id

cat("Taxon origin distance matrix:\n")
print(round(originDist_taxon, 1))

cat("Origin sites defined:\n")
cat(" Population level:", nrow(originSites_col),    "points\n")
cat(" Unit level: ", nrow(originSites_unit), "points\n")
cat(" Taxon level: ", nrow(originSites_taxon), "points\n")

### --- Pairwise comparisons between biogeographical units --- ###

# WesternMost vs Central-West
originSites_WM_CW <- originSites_unit %>%
  filter(site_id %in% c("WesternMost", "Central-West"))



originDist_WM_CW <- originDist_unit[c("WesternMost", "Central-West"),
                                     c("WesternMost", "Central-West"), 
                                     drop = FALSE]
cat("\nDistance matrix: WesternMost vs Central-West\n")
print(round(originDist_WM_CW, 1))

# WesternMost vs Central-East
originSites_WM_CE <- originSites_unit %>%
  filter(site_id %in% c("WesternMost", "Central-East"))

originDist_WM_CE <- originDist_unit[c("WesternMost", "Central-East"),
                                     c("WesternMost", "Central-East"), 
                                     drop = FALSE]

cat("\nDistance matrix: WesternMost vs Central-East\n")
print(round(originDist_WM_CE, 1))

# WesternMost vs EasternMost
originSites_WM_EM <- originSites_unit %>%
  filter(site_id %in% c("WesternMost", "EasternMost"))

originDist_WM_EM <- originDist_unit[c("WesternMost", "EasternMost"),
                                     c("WesternMost", "EasternMost"), 
                                     drop = FALSE]

cat("\nDistance matrix: WesternMost vs EasternMost\n")
print(round(originDist_WM_EM, 1))

# Central-West vs Central-East
originSites_CW_CE <- originSites_unit %>%
  filter(site_id %in% c("Central-West", "Central-East"))

originDist_CW_CE <- originDist_unit[c("Central-West", "Central-East"),
                                     c("Central-West", "Central-East"), 
                                     drop = FALSE]

cat("\nDistance matrix: Central-West vs Central-East\n")
print(round(originDist_CW_CE, 1))

# Central-West vs EasternMost
originSites_CW_EM <- originSites_unit %>%
  filter(site_id %in% c("Central-West", "EasternMost"))

originDist_CW_EM <- originDist_unit[c("Central-West", "EasternMost"),
                                     c("Central-West", "EasternMost"), 
                                     drop = FALSE]

cat("\nDistance matrix: Central-West vs EasternMost\n")
print(round(originDist_CW_EM, 1))

# Central-East vs EasternMost
originSites_CE_EM <- originSites_unit %>%
  filter(site_id %in% c("Central-East", "EasternMost"))

originDist_CE_EM <- originDist_unit[c("Central-East", "EasternMost"),
                                     c("Central-East", "EasternMost"), 
                                     drop = FALSE]

cat("\nDistance matrix: Central-East vs EasternMost\n")
print(round(originDist_CE_EM, 1))

```

## Target Sites

Non-breeding target regions correspond directly to the **WA** (wintering
area) categories recorded in `phen_data`. The eight named
regions are: Alboran Sea, Western Mediterranean, Central Mediterranean, Adriatic Sea,
Aegean Sea, Black Sea, Saharan Upwelling, and South European Atlantic Shelf.

```{r target-sites-dist}

targetSites <- readRDS("data/targetSites_MC.rds")

cat("Target sites defined:", nrow(targetSites), "wintering zones\n")
print(targetSites$site_id)

targetDist <- calculate_distance_matrix(targetSites, ocean.r)
colnames(targetDist) <- rownames(targetDist) <- targetSites$site_id

cat("Target distance matrix:\n")
print(round(targetDist, 1))
```
---
# Relative abundances of the origin Sites 

```{r rel-ab}
# === TAXON LEVEL ===
pop_sizes_taxa <- matrix(
  c(originSites_taxon$site_id,c(3000*2, 27310*2)),
  nrow=2,ncol=2
  )
originRelAbund_taxa <- c(
  as.numeric(pop_sizes_taxa[1,2])/sum(as.numeric(pop_sizes_taxa[,2])),
  as.numeric(pop_sizes_taxa[2,2])/sum(as.numeric(pop_sizes_taxa[,2]))
  )
# === POPULATION: PUFMAU ===
pop_sizes_PUFMAU <- matrix(
  c(originSites_PUFMAU$Pop,c(530*2, 900*2)),
  nrow=2,ncol=2
  )
originRelAbund_PUFMAU<- c(
  as.numeric(pop_sizes_PUFMAU[1,2])/sum(as.numeric(pop_sizes_PUFMAU[,2])),
  as.numeric(pop_sizes_PUFMAU[2,2])/sum(as.numeric(pop_sizes_PUFMAU[,2]))
  )
# === POPULATION: PUFYEL ===
pop_sizes_PUFYEL <- matrix(
  c(originSites_PUFYEL$Pop, 
    c(11050*2,300*2,1685*2,13000*2,425*2,
      585*2)),
  nrow=6,ncol=2
  )
originRelAbund_PUFYEL<- c(
  as.numeric(pop_sizes_PUFYEL[1,2])/sum(as.numeric(pop_sizes_PUFYEL[,2])),
  as.numeric(pop_sizes_PUFYEL[2,2])/sum(as.numeric(pop_sizes_PUFYEL[,2])),
  as.numeric(pop_sizes_PUFYEL[3,2])/sum(as.numeric(pop_sizes_PUFYEL[,2])),
  as.numeric(pop_sizes_PUFYEL[4,2])/sum(as.numeric(pop_sizes_PUFYEL[,2])),
  as.numeric(pop_sizes_PUFYEL[5,2])/sum(as.numeric(pop_sizes_PUFYEL[,2])),
  as.numeric(pop_sizes_PUFYEL[6,2])/sum(as.numeric(pop_sizes_PUFYEL[,2]))
  )
# === Biogeographic unit LEVEL ===
pop_sizes_unit <- matrix(c(
  originSites_unit$site_id,
  c(1685*2+425*2,585*2+11050*2,13000*2,3000*2+300*2)), # CE,CW,EM,WM
  nrow=4,ncol=2)

### Relative abundance ####

originRelAbund_unit <- c(
  as.numeric(pop_sizes_unit[1,2])/sum(as.numeric(pop_sizes_unit[,2])),
  as.numeric(pop_sizes_unit[2,2])/sum(as.numeric(pop_sizes_unit[,2])),
  as.numeric(pop_sizes_unit[3,2])/sum(as.numeric(pop_sizes_unit[,2])),
  as.numeric(pop_sizes_unit[4,2])/sum(as.numeric(pop_sizes_unit[,2]))
  )


# === POPULATION: WesternMost ===
pop_sizes_WM <- matrix(c(
  originSites_WM$Pop,c(300*2, 530*2, 900*2)),#La Mola, Sa Conillera, Sa Cella
  nrow=3,ncol=2)

originRelAbund_WM <- c(
  as.numeric(pop_sizes_WM[1, 2]) / sum(as.numeric(pop_sizes_WM[, 2])),
  as.numeric(pop_sizes_WM[2, 2]) / sum(as.numeric(pop_sizes_WM[, 2])),
  as.numeric(pop_sizes_WM[3, 2]) / sum(as.numeric(pop_sizes_WM[, 2]))
)

# === POPULATION: Central-West ===
pop_sizes_CW <- matrix(c(
  originSites_CW$Pop,c(11050*2,425*2)),#Tavolara,Porquerolles,Port-Cros
  nrow=2,ncol=2)

originRelAbund_CW <- c(
  as.numeric(pop_sizes_CW[1, 2]) / sum(as.numeric(pop_sizes_CW[, 2])),
  as.numeric(pop_sizes_CW[2, 2]) / sum(as.numeric(pop_sizes_CW[, 2]))
  )

# === POPULATION: Central-East ===
pop_sizes_CE <- matrix(c(
  originSites_CE$Pop,c(1685*2,425*2)),#Majjistral,Rdum,Zaklopatica,Veli Maslovnjak
  nrow=2,ncol=2)

originRelAbund_CE<- c(
  as.numeric(pop_sizes_CE[1, 2]) / sum(as.numeric(pop_sizes_CE[, 2])),
  as.numeric(pop_sizes_CE[2, 2]) / sum(as.numeric(pop_sizes_CE[, 2]))
)

# === POPULATION: EasternMost  ===
pop_sizes_EM <- matrix(c(
  originSites_EM$Colony,c(5460*2, 700*2)),#Gyaros, Kappari
  nrow=2,ncol=2)

originRelAbund_EM <- c(
  as.numeric(pop_sizes_EM[1, 2]) / sum(as.numeric(pop_sizes_EM[, 2])),
  as.numeric(pop_sizes_EM[2, 2]) / sum(as.numeric(pop_sizes_EM[, 2]))
)
cat("Relative abundance vectors calculated for all levels.\n")

### --- Pairwise comparisons between biogeographical units --- ###

# WesternMost vs Central-West

pop_sizes_WM_CW <- matrix(c(
  originSites_WM_CW$site_id,c(585*2+11050*2, 3000*2+300*2)),#
  nrow=2,ncol=2)

originRelAbund_WM_CW <- c(
  as.numeric(pop_sizes_WM_CW[1, 2]) / sum(as.numeric(pop_sizes_WM_CW[, 2])),
  as.numeric(pop_sizes_WM_CW[2, 2]) / sum(as.numeric(pop_sizes_WM_CW[, 2]))
)

# WesternMost vs Central-East

pop_sizes_WM_CE <- matrix(c(
  originSites_WM_CE$site_id,c(1685*2+425*2, 3000*2+300*2)),#
  nrow=2,ncol=2)

originRelAbund_WM_CE <- c(
  as.numeric(pop_sizes_WM_CE[1, 2]) / sum(as.numeric(pop_sizes_WM_CE[, 2])),
  as.numeric(pop_sizes_WM_CE[2, 2]) / sum(as.numeric(pop_sizes_WM_CE[, 2]))
)

# WesternMost vs EasternMost

pop_sizes_WM_EM <- matrix(c(
  originSites_WM_EM$site_id,c(13000*2, 3000*2+300*2)),#
  nrow=2,ncol=2)

originRelAbund_WM_EM <- c(
  as.numeric(pop_sizes_WM_EM[1, 2]) / sum(as.numeric(pop_sizes_WM_EM[, 2])),
  as.numeric(pop_sizes_WM_EM[2, 2]) / sum(as.numeric(pop_sizes_WM_EM[, 2]))
)

# Central-West vs Central-East

pop_sizes_CW_CE <- matrix(c(
  originSites_CW_CE$site_id,c(1685*2+425*2, 585*2+11050*2)),#
  nrow=2,ncol=2)

originRelAbund_CW_CE <- c(
  as.numeric(pop_sizes_CW_CE[1, 2]) / sum(as.numeric(pop_sizes_CW_CE[, 2])),
  as.numeric(pop_sizes_CW_CE[2, 2]) / sum(as.numeric(pop_sizes_CW_CE[, 2]))
)


# Central-West vs EasternMost
pop_sizes_CW_EM <- matrix(c(
  originSites_CW_EM$site_id,c(585*2+11050*2, 13000*2)),#
  nrow=2,ncol=2)

originRelAbund_CW_EM <- c(
  as.numeric(pop_sizes_CW_EM[1, 2]) / sum(as.numeric(pop_sizes_CW_EM[, 2])),
  as.numeric(pop_sizes_CW_EM[2, 2]) / sum(as.numeric(pop_sizes_CW_EM[, 2]))
)


# Central-East vs EasternMost
pop_sizes_CE_EM <- matrix(c(
  originSites_CE_EM$site_id,c(585*2+11050*2, 13000*2)),#
  nrow=2,ncol=2)

originRelAbund_CE_EM <- c(
  as.numeric(pop_sizes_CE_EM[1, 2]) / sum(as.numeric(pop_sizes_CE_EM[, 2])),
  as.numeric(pop_sizes_CE_EM[2, 2]) / sum(as.numeric(pop_sizes_CE_EM[, 2]))
)

```
---

# Taxon-Level Analysis

We first estimate MC at the broadest taxonomic level, contrasting the two
Mediterranean shearwater taxa Origin sites are the merged colony polygons
for each taxon; target sites are the wintering zones defined above.

## Prepare Data

```{r taxon-data}
dat_taxon <- prepare_analysis_level(
  tracking_df  = tracking,
  zones_sf = targetSites
)

cat(sprintf("Taxon-level analysis: %d individuals total\n", n_distinct(dat_taxon$targetPoints$geo_trip)))
cat("Sample sizes by taxon:\n")
print(table(dat_taxon$targetPoints$Sp))

```

```{r convert-meters-AEQD}
# Convert spatial objects to AEQD (meters)

# originPoints and targetPoints

originPoints_taxon <- dat_taxon$originPoints%>%
  mutate(geometry = case_when(
    Sp == "PUFYEL" ~ st_centroid(originSites_PUFYEL$geometry[originSites_PUFYEL$Pop == "Sardinia"]),
    Sp == "PUFMAU" ~ st_centroid(originSites_PUFMAU$geometry[originSites_PUFMAU$Pop == "Mallorca"]),
    TRUE ~ NA
  ))

originPoints_taxon <- st_transform(originPoints_taxon, CRS_AEQD)
targetPoints_taxon <- st_transform(dat_taxon$targetPoints, CRS_AEQD)

# originSites and targetSites
originSites_taxon <- st_transform(originSites_taxon, CRS_AEQD)
targetSites_sf <- st_transform(targetSites, CRS_AEQD)
```

## Estimate MC — Taxon Level

```{r taxon-estMC}
mc_taxon <- estMC(
  isGL         = FALSE, # as we use modeled positions there is no need to add model error
  originDist   = originDist_taxon,
  targetDist   = targetDist,
  originPoints = originPoints_taxon,
  targetPoints = targetPoints_taxon ,
  originSites  = originSites_taxon,
  targetSites  = targetSites_sf,
  originNames  = originSites_taxon$site_id,
  targetNames  = targetSites_sf$site_id,
  originRelAbund = originRelAbund_taxa,
  nSim         = 1000,
  nSamples     = 1000,
  verbose      = FALSE
)

cat("=== Taxon-level MC estimate ===\n")
print(mc_taxon)
```
---

# Biogeographical-Level Analysis

We next estimate MC across the four biogeographic units, using unit-level origin polygons.

```{r convert-meters-AEQD}

originPoints_unit <- dat_taxon$originPoints%>%
  mutate(geometry = case_when(
    Group == "Central-West" ~ st_centroid(originSites_col$geometry[originSites_col$Pop == "Sardinia"]),
    Group == "Central-East" ~ st_centroid(originSites_col$geometry[originSites_col$Pop == "Malta"]),
    Group == "WesternMost" ~ st_centroid(originSites_col$geometry[originSites_col$Pop == "Mallorca"]),
    Group == "EasternMost" ~ st_centroid(originSites_col$geometry[originSites_col$Pop == "Greece"]),
    TRUE ~ NA
  ))

originPoints_unit <- st_transform(originPoints_unit, CRS_AEQD)
originSites_unit <- st_transform(originSites_unit, CRS_AEQD)
```

## Estimate MC — Population Level

```{r unit-estMC}
mc_unit <- estMC(
  isGL         = FALSE,
  originDist   = originDist_unit,
  targetDist   = targetDist,
  originPoints = originPoints_unit,
  targetPoints = targetPoints_taxon, # same as mc_taxon
  originSites  = originSites_unit,
  targetSites  = targetSites_sf,
  originNames  = originSites_unit$site_id,
  targetNames  = targetSites_sf$site_id,
  originRelAbund = originRelAbund_unit,
  nSim         = 1000,
  nSamples     = 1000,
  verbose      = FALSE
)

cat("=== Population-level MC estimate ===\n")
print(mc_unit)
```
---

# Population-level Analyses

Finally, we estimate MC *within* each taxa and biogeographic unit, treating individual
breeding colonies as origin nodes. This reveals whether connectivity is driven
by among-unit or within-unit structure.

## PUFMAU (Ibiza & Mallorca)

```{r wm-prep}
dat_PUFMAU <- prepare_analysis_level(
  tracking_df  = tracking,
  pops_include = c("Ibiza","Mallorca"),
  zones_sf = targetSites
)

cat(sprintf("PUFMAU: %d individuals\n", dat_PUFMAU$n))

# Origin distance
originDist_PUFMAU <- calculate_distance_matrix(originSites_PUFMAU, ocean.r)
colnames(originDist_PUFMAU) <- rownames(originDist_PUFMAU) <-  originSites_PUFMAU$Pop

cat("Unit origin distance matrix:\n")
print(round(originDist_PUFMAU, 1))
```

```{r convert-meters-AEQD}
# Convert spatial objects to AEQD (meters)
originPoints_PUFMAU <- dat_PUFMAU$originPoints%>%
  mutate(geometry = case_when(
    Population == "Ibiza" ~ st_centroid(originSites_col$geometry[originSites_col$Pop == "Ibiza"]),
    Population == "Mallorca" ~ st_centroid(originSites_col$geometry[originSites_col$Pop == "Mallorca"]),
    TRUE ~ NA
  ))

originPoints_PUFMAU <- st_transform(originPoints_PUFMAU, CRS_AEQD)
# originPoints and targetPoints
targetPoints_PUFMAU <- st_transform(dat_PUFMAU$targetPoints, CRS_AEQD)

# originSites
originSites_PUFMAU <- st_transform(originSites_PUFMAU, CRS_AEQD)
```

```{r wm-estMC}
mc_PUFMAU <- estMC(
  isGL         = FALSE,
  originDist = originDist_PUFMAU,
  targetDist = targetDist,
  originPoints = originPoints_PUFMAU,
  targetPoints = targetPoints_PUFMAU,
  originSites  = originSites_PUFMAU,
  targetSites  = targetSites_sf,
  originNames = originSites_PUFMAU$Pop,
  targetNames = targetSites_sf$site_id,
  originRelAbund = originRelAbund_PUFMAU,
  nSim        = 1000,
  nSamples    = 1000,
  verbose      = FALSE
)

cat("=== PUFMAU population MC ===\n")
print(mc_PUFMAU)
```
## PUFYEL (France, Sardinia, Malta, Croatia, Greece)

```{r wm-prep}
dat_PUFYEL <- prepare_analysis_level(
  tracking_df  = tracking,
  pops_include = c("Minorca","France", "Sardinia", "Malta", "Croatia", "Greece"),
  zones_sf = targetSites
)

cat(sprintf("PUFYEL: %d individuals\n", dat_PUFYEL$n))

# Origin distance
originDist_PUFYEL <- calculate_distance_matrix(originSites_PUFYEL, ocean.r)
colnames(originDist_PUFYEL) <- rownames(originDist_PUFYEL) <-  originSites_PUFYEL$Pop

cat("Unit origin distance matrix:\n")
print(round(originDist_PUFYEL, 1))
```

```{r convert-meters-AEQD}
# Convert spatial objects to AEQD (meters)
originPoints_PUFYEL <- dat_PUFYEL$originPoints%>%
  mutate(geometry = case_when(
    Population == "Minorca" ~ st_centroid(originSites_col$geometry[originSites_col$Pop == "Minorca"]),
    Population == "France" ~ st_centroid(originSites_col$geometry[originSites_col$Pop == "France"]),
    Population == "Sardinia" ~ st_centroid(originSites_col$geometry[originSites_col$Pop == "Sardinia"]),
    Population == "Malta" ~ st_centroid(originSites_col$geometry[originSites_col$Pop == "Malta"]),
    Population == "Croatia" ~ st_centroid(originSites_col$geometry[originSites_col$Pop == "Croatia"]),
    Population == "Greece" ~ st_centroid(originSites_col$geometry[originSites_col$Pop == "Greece"]),
    TRUE ~ NA
  ))

originPoints_PUFYEL <- st_transform(originPoints_PUFYEL, CRS_AEQD)
# originPoints and targetPoints
targetPoints_PUFYEL <- st_transform(dat_PUFYEL$targetPoints, CRS_AEQD)

# originSites
originSites_PUFYEL <- st_transform(originSites_PUFYEL, CRS_AEQD)
```

```{r wm-estMC}
mc_PUFYEL <- estMC(
  isGL         = FALSE,
  originDist = originDist_PUFYEL,
  targetDist = targetDist,
  originPoints = originPoints_PUFYEL,
  targetPoints = targetPoints_PUFYEL,
  originSites = originSites_PUFYEL,
  targetSites = targetSites_sf,
  originNames = originSites_PUFYEL$Pop,
  targetNames = targetSites_sf$site_id,
  originRelAbund = originRelAbund_PUFYEL,
  nSim        = 1000,
  nSamples    = 1000,
  verbose      = FALSE
)

cat("=== PUFYEL population MC ===\n")
print(mc_PUFYEL)
```

## WesternMost (Sa Conillera, Sa Cella, Mallorca)

```{r wm-prep}
dat_WM <- prepare_analysis_level(
  tracking_df  = tracking,
  pops_include = c("Ibiza","Mallorca","Minorca"),
  zones_sf = targetSites
)

cat(sprintf("WesternMost: %d individuals\n", dat_WM$n))

# Origin distance
originDist_WM <- calculate_distance_matrix(originSites_WM, ocean.r)
colnames(originDist_WM) <- rownames(originDist_WM) <-  originSites_WM$Colony

cat("Unit origin distance matrix:\n")
print(round(originDist_WM, 1))
```

```{r convert-meters-AEQD}
# Convert spatial objects to AEQD (meters)
originPoints_WM <- dat_WM$originPoints%>%
  mutate(geometry = case_when(
    Population == "Ibiza" ~ st_centroid(originSites_col$geometry[originSites_col$Pop == "Ibiza"]),
    Population == "Mallorca" ~ st_centroid(originSites_col$geometry[originSites_col$Pop == "Mallorca"]),
    Population == "Minorca" ~ st_centroid(originSites_col$geometry[originSites_col$Pop == "Minorca"]),
    TRUE ~ NA
  ))

# originPoints and targetPoints
originPoints_WM <- st_transform(originPoints_WM, CRS_AEQD)
targetPoints_WM <- st_transform(dat_WM$targetPoints, CRS_AEQD)

# originSites
originSites_WM<- st_transform(originSites_WM, CRS_AEQD)

```

```{r wm-estMC}
mc_WM <- estMC(
  isGL         = FALSE,
  originDist = originDist_WM,
  targetDist = targetDist,
  originPoints = originPoints_WM,
  targetPoints = targetPoints_WM ,
  originSites  = originSites_WM,
  targetSites  = targetSites_sf,
  originNames = originSites_WM$Pop,
  targetNames = targetSites_sf$site_id,
  originRelAbund = originRelAbund_WM,
  nSim        = 1000,
  nSamples    = 1000,
  verbose      = FALSE
)

cat("=== WesternMost population MC ===\n")
print(mc_WM)
```

## Central-West (FRANCE & SARDINIA)

```{r cw-prep}
dat_CW <- prepare_analysis_level(
  tracking_df  = tracking,
  pops_include = c("France","Sardinia"),
  zones_sf = targetSites
)

# Origin distance
originDist_CW <- calculate_distance_matrix(originSites_CW, ocean.r)
colnames(originDist_CW) <- rownames(originDist_CW) <-  originSites_CW$Colony

cat("Unit origin distance matrix:\n")
print(round(originDist_CW, 1))
```

```{r convert-meters-AEQD}
# Convert spatial objects to AEQD (meters)

# originPoints and targetPoints
originPoints_CW <- dat_CW$originPoints%>%
  mutate(geometry = case_when(
    Population == "France" ~ st_centroid(originSites_col$geometry[originSites_col$Pop == "France"]),
    Population == "Sardinia" ~ st_centroid(originSites_col$geometry[originSites_col$Pop == "Sardinia"]),
    TRUE ~ NA
  ))

# originPoints and targetPoints
originPoints_CW <- st_transform(originPoints_CW, CRS_AEQD)
targetPoints_CW <- st_transform(dat_CW$targetPoints, CRS_AEQD)

# originSites 
originSites_CW <- st_transform(originSites_CW, CRS_AEQD)
```

```{r cw-estMC}
mc_CW <- estMC(
  isGL         = FALSE,
  originDist = originDist_CW,
  targetDist = targetDist,
  originPoints = originPoints_CW,
  targetPoints = targetPoints_CW,
  originSites  = originSites_CW,
  targetSites  = targetSites_sf,
  originNames = originSites_CW$Colony,
  targetNames = targetSites_sf$site_id,
  originRelAbund = originRelAbund_CW,
  nSim         = 1000,
  nSamples     = 1000,
  verbose      = FALSE
)

cat("=== Central-West population MC ===\n")
print(mc_CW)
```

## Central-East (MALTA & CROATIA)

```{r ce-prep}
dat_CE <- prepare_analysis_level(
  tracking_df  = tracking,
  pops_include = c("Malta","Croatia"),
  zones_sf = targetSites
)

cat(sprintf("Central-East: %d individuals\n", dat_CE$n))

# Origin distance
originDist_CE <- calculate_distance_matrix(originSites_CE, ocean.r)
colnames(originDist_CE) <- rownames(originDist_CE) <-  originSites_CE$Colony

cat("Unit origin distance matrix:\n")
print(round(originDist_CE, 1))
```

```{r convert-meters-AEQD}
# Convert spatial objects to AEQD (meters)

# originPoints and targetPoints
originPoints_CE <- dat_CE$originPoints%>%
  mutate(geometry = case_when(
    Population == "Malta" ~ st_centroid(originSites_col$geometry[originSites_col$Pop == "Malta"]),
    Population == "Croatia" ~ st_centroid(originSites_col$geometry[originSites_col$Pop == "Croatia"]),
    TRUE ~ NA
  ))

# originPoints and targetPoints
originPoints_CE <- st_transform(originPoints_CE, CRS_AEQD)
targetPoints_CE <- st_transform(dat_CE$targetPoints, CRS_AEQD)

# originSites 
originSites_CE <- st_transform(originSites_CE, CRS_AEQD)

```

```{r ce-estMC}
mc_CE <- estMC(
  isGL         = FALSE,
  originDist = originDist_CE,
  targetDist = targetDist,
  originPoints = originPoints_CE,
  targetPoints = targetPoints_CE,
  originSites  = originSites_CE,
  targetSites  = targetSites_sf,
  originNames = originSites_CE$Colony,
  targetNames = targetSites_sf$site_id,
  originRelAbund = originRelAbund_CE,
  nSim         = 1000,
  nSamples     = 1000,
  verbose      = FALSE
)

cat("=== Central-East sub-population MC ===\n")
print(mc_CE)
```

## EasternMost (Cyclades and Dodecadense)

```{r em-prep}
dat_EM <- prepare_analysis_level(
  tracking_df  = tracking,
  pops_include = "Greece",
  zones_sf = targetSites
)

cat(sprintf("EasternMost: %d individuals\n", dat_EM$n))

# Origin distance
originDist_EM <- calculate_distance_matrix(originSites_EM, ocean.r)
colnames(originDist_EM) <- rownames(originDist_EM) <-  originSites_EM$Colony

cat("Unit origin distance matrix:\n")
print(round(originDist_EM, 1))

```
```{r convert-meters-AEQD}
# Convert spatial objects to AEQD (meters)

# originPoints and targetPoints
dat_EM$originPoints <- st_transform(dat_EM$originPoints, CRS_AEQD)
dat_EM$targetPoints <- st_transform(dat_EM$targetPoints, CRS_AEQD)

# originSites 
originSites_EM <- st_transform(originSites_EM, CRS_AEQD)

```

```{r em-estMC}
mc_EM <- estMC(
  isGL         = FALSE,
  originDist = originDist_EM,
  targetDist = targetDist,
  originPoints = dat_EM$originPoints,
  targetPoints = dat_EM$targetPoints,
  originSites  = originSites_EM,
  targetSites  = targetSites_sf,
  originNames = originSites_EM$Colony,
  targetNames = targetSites_sf$site_id,
  originRelAbund = originRelAbund_EM,
  resampleProjection = CRS_GEO,
  nSim         = 1000,
  nSamples     = 1000,
  verbose      = FALSE
)

cat("=== EasternMost sub-population MC ===\n")
print(mc_EM)
```
---

# Pairwise comparisons

## WM vs CW

```{r convert-meters-AEQD}
# Convert spatial objects to AEQD (meters)

# originPoints and targetPoints
# Select the main population of each unit
originPoints_WM_CW <- dat_taxon$originPoints%>%
  filter(Group %in% c("Central-West","WesternMost"))%>%
  mutate(geometry = case_when(
    Group == "Central-West" ~ st_centroid(originSites_col$geometry[originSites_col$Pop == "Sardinia"]),
    Group == "WesternMost" ~ st_centroid(originSites_col$geometry[originSites_col$Pop == "Mallorca"]),
    TRUE ~ NA
  ))

targetPoints_WM_CW <- dat_taxon$targetPoints%>%
  filter(Group %in% c("Central-West","WesternMost"))
  
originPoints_WM_CW <- st_transform(originPoints_WM_CW, CRS_AEQD)
originSites_WM_CW <- st_transform(originSites_WM_CW, CRS_AEQD)

targetPoints_WM_CW <- st_transform(targetPoints_WM_CW, CRS_AEQD)
targetSites_sf <- st_transform(targetSites, CRS_AEQD)
```


```{r WM_CW-estMC}
mc_WM_CW <- estMC(
  isGL         = FALSE, # as we use modeled positions there is no need to add model error
  originDist   = originDist_WM_CW,
  targetDist   = targetDist,
  originPoints = originPoints_WM_CW,
  targetPoints = targetPoints_WM_CW ,
  originSites  = originSites_WM_CW,
  targetSites  = targetSites_sf,
  originNames  = originSites_WM_CW$site_id,
  targetNames  = targetSites_sf$site_id,
  originRelAbund = originRelAbund_WM_CW,
  nSim         = 1000,
  nSamples     = 1000,
  verbose      = FALSE
)

cat("=== WM vs CW MC estimate ===\n")
print(mc_WM_CW)
```

## WM vs CE

```{r convert-meters-AEQD}
# Convert spatial objects to AEQD (meters)

# originPoints and targetPoints
# Select the main population of each unit

originPoints_WM_CE <- dat_taxon$originPoints%>%
  filter(Group %in% c("Central-East","WesternMost"))%>%
  mutate(geometry = case_when(
    Group == "Central-East" ~ st_centroid(originSites_col$geometry[originSites_col$Pop == "Malta"]),
    Group == "WesternMost" ~ st_centroid(originSites_col$geometry[originSites_col$Pop == "Mallorca"]),
    TRUE ~ NA
  ))

targetPoints_WM_CE <- dat_taxon$targetPoints%>%
  filter(Group %in% c("Central-East","WesternMost"))
  
originPoints_WM_CE <- st_transform(originPoints_WM_CE, CRS_AEQD)
originSites_WM_CE <- st_transform(originSites_WM_CE, CRS_AEQD)

targetPoints_WM_CE <- st_transform(targetPoints_WM_CE, CRS_AEQD)
targetSites_sf <- st_transform(targetSites, CRS_AEQD)
```


```{r WM_CE-estMC}
mc_WM_CE <- estMC(
  isGL         = FALSE, # as we use modeled positions there is no need to add model error
  originDist   = originDist_WM_CE,
  targetDist   = targetDist,
  originPoints = originPoints_WM_CE,
  targetPoints = targetPoints_WM_CE ,
  originSites  = originSites_WM_CE,
  targetSites  = targetSites_sf,
  originNames  = originSites_WM_CE$site_id,
  targetNames  = targetSites_sf$site_id,
  originRelAbund = originRelAbund_WM_CE,
  nSim         = 1000,
  nSamples     = 1000,
  verbose      = FALSE
)

cat("=== WM vs CE MC estimate ===\n")
print(mc_WM_CE)
```

## WM vs EM

```{r convert-meters-AEQD}
# Convert spatial objects to AEQD (meters)

# originPoints and targetPoints
# Select the main population of each unit

originPoints_WM_EM <- dat_taxon$originPoints%>%
  filter(Group %in% c("EasternMost","WesternMost"))%>%
  mutate(geometry = case_when(
    Group == "EasternMost" ~ st_centroid(originSites_col$geometry[originSites_col$Pop == "Greece"]),
    Group == "WesternMost" ~ st_centroid(originSites_col$geometry[originSites_col$Pop == "Mallorca"]),
    TRUE ~ NA
  ))

targetPoints_WM_EM <- dat_taxon$targetPoints%>%
  filter(Group %in% c("EasternMost","WesternMost"))
  
originPoints_WM_EM <- st_transform(originPoints_WM_EM, CRS_AEQD)
originSites_WM_EM <- st_transform(originSites_WM_EM, CRS_AEQD)

targetPoints_WM_EM <- st_transform(targetPoints_WM_EM, CRS_AEQD)
targetSites_sf <- st_transform(targetSites, CRS_AEQD)
```


```{r WM_EM-estMC}
mc_WM_EM <- estMC(
  isGL         = FALSE, # as we use modeled positions there is no need to add model error
  originDist   = originDist_WM_EM,
  targetDist   = targetDist,
  originPoints = originPoints_WM_EM,
  targetPoints = targetPoints_WM_EM ,
  originSites  = originSites_WM_EM,
  targetSites  = targetSites_sf,
  originNames  = originSites_WM_EM$site_id,
  targetNames  = targetSites_sf$site_id,
  originRelAbund = originRelAbund_WM_EM,
  nSim         = 1000,
  nSamples     = 1000,
  verbose      = FALSE
)

cat("=== WM vs EM MC estimate ===\n")
print(mc_WM_EM)
```
## CW vs CE

```{r convert-meters-AEQD}
# Convert spatial objects to AEQD (meters)

# originPoints and targetPoints
# Select the main population of each unit

originPoints_CW_CE <- dat_taxon$originPoints%>%
  filter(Group %in% c("Central-West","Central-East"))%>%
  mutate(geometry = case_when(
    Group == "Central-West" ~ st_centroid(originSites_col$geometry[originSites_col$Pop == "Sardinia"]),
    Group == "Central-East" ~ st_centroid(originSites_col$geometry[originSites_col$Pop == "Malta"]),
    TRUE ~ NA
  ))

targetPoints_CW_CE <- dat_taxon$targetPoints%>%
  filter(Group %in% c("Central-West","Central-East"))
  
originPoints_CW_CE <- st_transform(originPoints_CW_CE, CRS_AEQD)
originSites_CW_CE <- st_transform(originSites_CW_CE, CRS_AEQD)

targetPoints_CW_CE <- st_transform(targetPoints_CW_CE, CRS_AEQD)
targetSites_sf <- st_transform(targetSites, CRS_AEQD)
```


```{r CW_CE-estMC}
mc_CW_CE <- estMC(
  isGL         = FALSE, # as we use modeled positions there is no need to add model error
  originDist   = originDist_CW_CE,
  targetDist   = targetDist,
  originPoints = originPoints_CW_CE,
  targetPoints = targetPoints_CW_CE ,
  originSites  = originSites_CW_CE,
  targetSites  = targetSites_sf,
  originNames  = originSites_CW_CE$site_id,
  targetNames  = targetSites_sf$site_id,
  originRelAbund = originRelAbund_CW_CE,
  nSim         = 1000,
  nSamples     = 1000,
  verbose      = FALSE
)

cat("=== CW vs CE MC estimate ===\n")
print(mc_CW_CE)
```
## CW vs EM

```{r convert-meters-AEQD}
# Convert spatial objects to AEQD (meters)

# originPoints and targetPoints
# Select the main population of each unit

originPoints_CW_EM <- dat_taxon$originPoints%>%
  filter(Group %in% c("Central-West","EasternMost"))%>%
  mutate(geometry = case_when(
    Group == "Central-West" ~ st_centroid(originSites_col$geometry[originSites_col$Pop == "Sardinia"]),
    Group == "EasternMost" ~ st_centroid(originSites_col$geometry[originSites_col$Pop == "Greece"]),
    TRUE ~ NA
  ))

targetPoints_CW_EM <- dat_taxon$targetPoints%>%
  filter(Group %in% c("Central-West","EasternMost"))
  
originPoints_CW_EM <- st_transform(originPoints_CW_EM, CRS_AEQD)
originSites_CW_EM <- st_transform(originSites_CW_EM, CRS_AEQD)

targetPoints_CW_EM <- st_transform(targetPoints_CW_EM, CRS_AEQD)
targetSites_sf <- st_transform(targetSites, CRS_AEQD)
```


```{r CW_EM-estMC}
mc_CW_EM <- estMC(
  isGL         = FALSE, # as we use modeled positions there is no need to add model error
  originDist   = originDist_CW_EM,
  targetDist   = targetDist,
  originPoints = originPoints_CW_EM,
  targetPoints = targetPoints_CW_EM ,
  originSites  = originSites_CW_EM,
  targetSites  = targetSites_sf,
  originNames  = originSites_CW_EM$site_id,
  targetNames  = targetSites_sf$site_id,
  originRelAbund = originRelAbund_CW_EM,
  nSim         = 1000,
  nSamples     = 1000,
  verbose      = FALSE
)

cat("=== CW vs EM MC estimate ===\n")
print(mc_CW_EM)
```
## CE vs EM

```{r convert-meters-AEQD}
# Convert spatial objects to AEQD (meters)

# originPoints and targetPoints
# Select the main population of each unit

originPoints_CE_EM <- dat_taxon$originPoints%>%
  filter(Group %in% c("Central-East","EasternMost"))%>%
  mutate(geometry = case_when(
    Group == "Central-East" ~ st_centroid(originSites_col$geometry[originSites_col$Pop == "Malta"]),
    Group == "EasternMost" ~ st_centroid(originSites_col$geometry[originSites_col$Pop == "Greece"]),
    TRUE ~ NA
  ))

targetPoints_CE_EM <- dat_taxon$targetPoints%>%
  filter(Group %in% c("Central-East","EasternMost"))
  
originPoints_CE_EM <- st_transform(originPoints_CE_EM, CRS_AEQD)
originSites_CE_EM <- st_transform(originSites_CE_EM, CRS_AEQD)

targetPoints_CE_EM <- st_transform(targetPoints_CE_EM, CRS_AEQD)
targetSites_sf <- st_transform(targetSites, CRS_AEQD)
```


```{r CE_EM-estMC}
mc_CE_EM <- estMC(
  isGL         = FALSE, # as we use modeled positions there is no need to add model error
  originDist   = originDist_CE_EM,
  targetDist   = targetDist,
  originPoints = originPoints_CE_EM,
  targetPoints = targetPoints_CE_EM ,
  originSites  = originSites_CE_EM,
  targetSites  = targetSites_sf,
  originNames  = originSites_CE_EM$site_id,
  targetNames  = targetSites_sf$site_id,
  originRelAbund = originRelAbund_CE_EM,
  nSim         = 1000,
  nSamples     = 1000,
  verbose      = FALSE
)

cat("=== CE vs EM MC estimate ===\n")
print(mc_CE_EM)
```
---

# Visualization: Circlize Chord Diagram {#circlize}

A chord diagram shows the proportion of tracked individuals from each
breeding population that winter in each target zone. Chord width is
proportional to the number of individuals.

```{r circlize-data}

assignments <- dat_taxon$targetPoints

assignments$n <- 1

pop_sizes <- as.data.frame(matrix(
  c(
    c('Ibiza','Mallorca','Minorca','France',
      'Sardinia','Malta','Croatia','Greece'),
    c(530*2,900*2, 300*2, 585*2, 11050*2, 1685*2, 425*2,13000*2)),
  nrow=8,ncol=2))

assignments <- assignments %>%
  left_join(.,pop_sizes,by=c('Population'='V1'))%>%
  mutate(pop_size = as.numeric(V2))%>%
  dplyr::select(-V2)

for_circle <- assignments %>% 
  group_by(Group, Population, site_id, pop_size, n_col = n) %>%
  dplyr::summarise(n_col_win = n()) %>% 
  mutate(ind_weight = pop_size/n_col) %>% 
  ungroup()


for_circle_pop <- for_circle %>% 
  dplyr::select(Group, Population, site_id, n_col_win, ind_weight) %>% 
  mutate(circle_arch_value = n_col_win*ind_weight)

for_plot <- for_circle_pop %>% 
  mutate(Population = str_to_title(Population),
         WA = case_when(
           site_id == "South European Atlantic Shelf" ~ "S Euro-Atl Shelf",
           site_id == "Saharan Upwelling" ~ "Saharan Upw",
           site_id == "Western Mediterranean" ~ "Western Med",
           site_id == "Central Mediterranean" ~ "Central Med",
           TRUE ~ site_id
         ))%>%
  dplyr::select(Population, WA, circle_arch_value)

union(for_plot[[1]], for_plot[[2]])

new_order <- c("Ibiza","Mallorca",'Minorca','France','Sardinia','Malta','Croatia','Greece',
               'Black Sea' , 'Levantine Sea','Aegean Sea' ,'Adriatic Sea','Central Med','Western Med', 
               'Alboran Sea','Saharan Upw', 'S Euro-Atl Shelf')

grid.col <- c(
  "Ibiza"          = mako(12)[1], 
  "Mallorca"       = mako(12)[3], 
  "Minorca"        = mako(12)[4], 
  "France"         = mako(12)[5], 
  "Sardinia"       = mako(12)[6], 
  "Malta"          = mako(12)[7], 
  "Croatia"        = mako(12)[9], 
  "Greece"         = mako(12)[11], 
  
  "Black Sea"       = "#000000", 
  "Levantine Sea"   = "#08020a", 
  "Aegean Sea"      = "#180424", 
  "Adriatic Sea"    = "#34074a", 
  "Central Med"     = "#57106e", 
  "Western Med"     = "#882289", 
  "Alboran Sea"     = "#b73779", 
  "Saharan Upw"     = "#eb5e7c", 
  "S Euro-Atl Shelf"= "#fe9f6d"
)

```

```{r circlize-plot, fig.width=9, fig.height=9}

png("outputs/Circular_plot.png", height = 18, width = 18, units = "cm", res = 300,bg='white')
# Configurar parámetros
circos.par(start.degree = 170, clock.wise = TRUE,
           gap.after = c(rep(1, length(unique(for_plot[[1]])) - 1), 7,
                         rep(1, length(unique(for_plot[[2]])) - 1), 7))

# Crear chord diagram
chordDiagram(for_plot, reduce = -1, order = new_order, grid.col = grid.col, 
             link.sort = T, link.decreasing = T, link.zindex = rank(for_plot[[3]]),
             directional = 1, direction.type = c("diffHeight", "arrows"), 
             link.arr.type = "big.arrow", 
             annotationTrack = c("grid"), annotationTrackHeight = c(0.05), 
             preAllocateTracks = list(track.height = 0.5))

# 1. Grupo WesternMost
highlight.sector(sector.index = c("Ibiza", "Mallorca"), track.index = 1,
                 col = NA, text = "WesternMost", text.vjust = -0.25, text.col = "black", cex = 0.9)

# 2. Grupo Central-West
highlight.sector(sector.index = c("Minorca", "France", "Sardinia"), track.index = 1,
                 col = NA, text = "Central-West", text.vjust = -0.25, text.col = "black", cex = 0.9)

# 3. Grupo Central-East
highlight.sector(sector.index = c("Malta", "Croatia"), track.index = 1,
                 col = NA, text = "Central-East", text.vjust = -0.25, text.col = "black", cex = 0.9)

# 4. Grupo EasternMost
highlight.sector(sector.index = "Greece", track.index = 1,
                 col = NA, text = "EasternMost", text.vjust = -0.25, text.col = "black", cex = 0.9)

# Añadir labels
circos.track(track.index = 1, panel.fun = function(x, y) {
  sector.name = CELL_META$sector.index
  v_adj = 0.1
  if(sector.name == "Aegean Sea") v_adj = 0.3
  if(sector.name == "Levantine Sea") v_adj = -0.5

  circos.text(CELL_META$xcenter, 
              CELL_META$ylim[1] + mm_y(2), # Elevamos un poco el texto del grid
              sector.name,
              facing = "clockwise", 
              niceFacing = TRUE, 
              adj = c(0, v_adj), 
              cex = 0.7)
}, bg.border = NA)

# Limpiar parámetros de circos
circos.clear()
dev.off()
```

---

# Session Information

```{r session-info}
sessionInfo()
```
