---
title: "Migratory Connectivity Analysis"
author: "Mario Gómez Álvarez"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
    theme: minimal
    number_sections: true
    fig_width: 12
    fig_height: 8
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  warning   = FALSE,
  message   = FALSE,
  fig.align = 'center',
  dpi       = 300
)
Sys.setenv(TZ = 'GMT')
```

## Introduction

This document estimates the strength of **migratory connectivity** (MC) for Mediterranean
shearwaters at three hierarchical levels:

1. **Taxon level** — Balearic (PUFMAU) vs. Yelkouan (PUFYEL) shearwaters.
2. **Biogeographical Population level** — Four biogeographic groups (WesternMost, Central-West,
   Central-East, EasternMost).
3. **Population level** — Individual breeding populations within each group.

---

# Setup

## Packages

```{r packages}
packages <- c(
  "tidyverse", "sf", "gdistance",
  "MigConnectivity",
  "rnaturalearth", "rnaturalearthdata",
  "circlize", "knitr", "kableExtra","RANN"
)

invisible(lapply(packages, library, character.only = TRUE))
```

## Coordinate Reference Systems

```{r projections}
CRS_GEO <- "EPSG:4326"
CRS_AEQD <- "+proj=aeqd +lat_0=38 +lon_0=15 +datum=WGS84 +units=m +no_defs"
CRS_MOLL <- "+proj=moll +lon_0=0 +datum=WGS84 +units=m +no_defs"
```

# Ocean Raster Creation

This section creates or loads the ocean raster for cost-distance calculations. The raster
defines ocean (value = 1) and land (value = 0) for the Mediterranean region.

```{r ocean-raster-creation}
# Define Mediterranean extent
xmin <- -20
xmax <- 50
ymin <- 20
ymax <- 55
resolution <- 0.05

# Create empty raster template
ocean.r <- raster(
  xmn = xmin, xmx = xmax,
  ymn = ymin, ymx = ymax,
  res = resolution,
  crs = CRS(CRS_GEO)
)

# Get land polygons from rnaturalearth
land_sf <- ne_countries(scale = "medium", returnclass = "sf") %>%
  st_union()

# Convert sf to sp for rasterization 
land_sp <- as_Spatial(land_sf)

# Rasterize land polygons: land areas will have a value, NA elsewhere
ocean.r <- rasterize(land_sp, ocean.r, field = 1)

# Convert: land = 0, ocean = 1
ocean.r[!is.na(ocean.r)] <- 0  # Land cells set to 0
ocean.r[is.na(ocean.r)] <- 1    # Ocean cells set to 1

cat("Ocean raster created:\n")
```

---

# Helper Functions 

Three reusable functions encapsulate the data-preparation steps that recur at
every analysis level.

```{r fn-relocate}
assign_centroid_from_zones <- function(tracking_df, zones_sf, ocean_raster) {

  pts_sf <- st_as_sf(tracking_df,
                     coords = c("Lon", "Lat"),
                     crs = CRS_GEO,
                     remove = FALSE)

  # Keep ONLY ocean points
  ocean_flag <- raster::extract(ocean_raster, st_coordinates(pts_sf)) == 1
  pts_sf <- pts_sf[ocean_flag, ]

  # Spatial join only ocean-valid points
  joined <- st_join(pts_sf, zones_sf, join = st_within)

  # Keep valid site assignments
  joined <- joined %>%
    filter(!is.na(site_id))

  counts <- joined %>%
    st_drop_geometry() %>%
    group_by(geo_trip, site_id) %>%
    summarise(n = n(), .groups = "drop")

  dominant_zone <- counts %>%
    group_by(geo_trip) %>%
    slice_max(n, n = 1, with_ties = FALSE) %>%
    ungroup()

  cat("\n--- Zone assignment diagnostics ---\n")
  print(table(dominant_zone$site_id, useNA = "ifany"))
  cat("----------------------------------\n\n")

  zone_centroids <- st_centroid(zones_sf)

  zone_centroids_df <- zone_centroids %>%
    st_drop_geometry() %>%
    bind_cols(st_coordinates(zone_centroids) %>% as.data.frame())

  result <- dominant_zone %>%
    left_join(zone_centroids_df, by = "site_id") %>%
    rename(Lon = X, Lat = Y)

  return(result)
}
```

## `prepare_analysis_level()`

```{r fn-prepare}
#' Prepare individual-level origin and target data for `estMC()`

prepare_analysis_level <- function(tracking_df, pops_include = NULL, zones_sf) {

  df <- tracking_df 
  
  if(!is.null(pops_include)){
  df <- df %>% 
    filter(Population %in% pops_include)}

  if (nrow(df) == 0L)
    stop("No tracking records match the requested populations.")

  # Assign points to targetSites
  win_pts <- assign_centroid_from_zones(df, zones_sf, ocean.r)

  df <- df%>%
    rename(Lon_win_og = Lon, Lat_win_og = Lat) %>%
    left_join(win_pts, by = "geo_trip")
  
  # Reasses NAs using the individual centroid
  df_ok  <- df %>% filter(!is.na(Lon), !is.na(Lat))%>%
    dplyr::select(-Lon_win_og,-Lat_win_og)
  
  df_na  <- df %>% filter(is.na(Lon) | is.na(Lat))%>%
    group_by(geo_trip)%>%
    mutate(
      Lon_win_og = mean(Lon_win_og, na.rm = TRUE),
      Lat_win_og = mean(Lat_win_og, na.rm = TRUE)
      )%>%
    ungroup
  
  na_sf <- st_as_sf(df_na, coords = c("Lon_win_og", "Lat_win_og"), crs = CRS_GEO)
  
  # Assign to the nearest zone
  nearest_zone <- st_nearest_feature(na_sf, targetSites)
  df_na$site_id <- targetSites$site_id[nearest_zone]
  df_na <- df_na %>%
    mutate(Lon = Lon_win_og, Lat = Lat_win_og)%>%
    dplyr::select(-Lon_win_og,-Lat_win_og)
  
  df_fixed <- bind_rows(df_ok, df_na)%>%
    distinct()
  
  # originPoints: breeding colony location for each individual
  originPoints <- st_as_sf(df_fixed, coords = c("Lon_col", "Lat_col"), crs = CRS_GEO)

  # targetPoints: individual wintering centroid for each individual
  targetPoints <- st_as_sf(df_fixed, coords = c("Lon", "Lat"), crs = CRS_GEO)
  
  list(
    originPoints = originPoints,
    targetPoints = targetPoints,
    n            = nrow(df_fixed)
    )
}
```
#' Calculate marine cost-distance matrix from origin or target sites
#'
#' Uses a raster-based cost-distance approach with gdistance package.
#' Requires an ocean raster (land = 0, ocean = 1).

``` {r cal_dist}
calculate_distance_matrix <- function(sites_sf, ocean_raster) {
  
  # Centroids of the polygons
  sites_centroids <- st_centroid(sites_sf)
  sites_centroids <- st_transform(sites_centroids, CRS_GEO)
  
  sites <- as.data.frame(st_coordinates(sites_centroids))
  colnames(sites) <- c("X", "Y")
  
  # Detect those points in land
  sites.to.relocate <- raster::extract(ocean_raster, sites) == 0
  
  if (any(sites.to.relocate)) {
    
    cat(sprintf("\n.....................................................................\n"))
    cat(sprintf("%d site(s) to be relocated from land to ocean.\n", sum(sites.to.relocate)))
    cat(sprintf(".....................................................................\n\n"))
    
    # Obtain points of the ocean raster
    ocean_cells <- rasterToPoints(ocean_raster, fun = function(x) x == 1)
    ocean_coords <- ocean_cells[, 1:2]  # X, Y
    
    land_sites <- sites[sites.to.relocate, ]
    
    # KNN → closest neighbor
    nn <- nn2(data = ocean_coords, query = land_sites, k = 1)
    
    sites[sites.to.relocate, ] <- ocean_coords[nn$nn.idx, ]
  }
  
  # Cost surface
  cost_surface <- ocean_raster
  projection(cost_surface) <- CRS(CRS_GEO)
  
  raster_tr <- transition(cost_surface, mean, directions = 8)
  raster_tr_corrected <- geoCorrection(raster_tr, type = "c", multpl = FALSE)
  
  # Cost distance
  raster_cosDist <- costDistance(
    raster_tr_corrected,
    as.matrix(sites),
    as.matrix(sites)
  )
  
  return(as.matrix(raster_cosDist))
}
```

---

# Data Loading {#data}

```{r load-data}

# Load phenological data; retain migrants with known departure dates

phen_all <- read.csv("Phen_data.csv") %>% filter(complete != "Uncomplete")%>%
  mutate(
    BA_dep_date = as.POSIXct(BA_dep_date, format = "%Y-%m-%dT%H:%M:%SZ", tz = "UTC"),
    BA_arr_date = as.POSIXct(BA_arr_date, format = "%Y-%m-%dT%H:%M:%SZ", tz = "UTC")
  ) %>%
  dplyr::select(geo_trip, BA_dep_date, BA_arr_date)

# True colony coordinates
colonies <- read.csv("colonies_coords.csv") 

# Geolocator tracking data — individual wintering centroids
# Computed as the mean position during the core non-breeding months (August-September)
# Restricted to migrant individuals identified via phen_data
tracking <- read.csv("all_trips.csv") %>%
  filter(geo_trip %in% unique(phen_all$geo_trip)) %>%
  filter(month %in% c(8,9)) %>% 
  left_join(
    colonies %>% dplyr::select(Colony, Lon_col = Lon, Lat_col = Lat),
    by = c("Colony")
  )%>%
  dplyr::select(-Date_Time,-month)


cat(sprintf("Tracking records: %d migrants across %d colonies\n",
            n_distinct(tracking$geo_trip), n_distinct(tracking$Colony)))
```
---

# Site Definitions {#sites}

`estMC()` requires geographic polygons that partition the breeding range into
**origin sites** and the non-breeding range into **target sites**. Assignments
of individuals to sites — and the resulting psi matrix — are computed internally
by `estMC()`.

## Origin Sites

Each breeding colony is represented by a circular buffer of radius 50 km,
giving one origin polygon per population. Group- and taxon-level polygons
are formed by dissolving individual colony buffers.

```{r origin-sites-dist}
# Convert colony table to sf and project for buffering
colonies_sf <- st_as_sf(colonies, coords = c("Lon", "Lat"), crs = CRS_GEO)
colonies_proj <- st_transform(colonies_sf, CRS_AEQD)

# Individual colony buffers (2 km radius)
originSites_col <- colonies_proj %>%
  mutate(geometry = st_buffer(geometry, dist = 2000)) %>%
  st_transform(CRS_GEO) %>%
  dplyr::select(Colony,Pop,Group,Taxon)

# Separate by group
originSites_WM <- originSites_col %>%
  filter(Group == "WesternMost")
originSites_CW <- originSites_col %>%
  filter(Group == "Central-West")
originSites_CE <- originSites_col %>%
  filter(Group == "Central-East")
originSites_EM <- originSites_col %>%
  filter(Group == "EasternMost")

# Group-level origin polygons (union of colony buffers within each group)
originSites_group <- originSites_col %>%
  group_by(Group) %>%
  summarise(geometry = st_union(geometry), .groups = "drop") %>%
  rename(site_id = Group)

# Group-level distance
originDist_group <- calculate_distance_matrix(originSites_group, ocean.r)
colnames(originDist_group) <- rownames(originDist_group) <-  originSites_group$site_id

cat("Group origin distance matrix:\n")
print(round(originDist_group, 1))

# Taxon-level origin polygons (union of colony buffers within each taxon)
originSites_taxon <- originSites_col %>%
  group_by(Taxon) %>%
  summarise(geometry = st_union(geometry), .groups = "drop") %>%
  rename(site_id = Taxon)

# Taxon-level distance
originDist_taxon <- calculate_distance_matrix(originSites_taxon, ocean.r)
colnames(originDist_taxon) <- rownames(originDist_taxon) <- originSites_taxon$site_id

cat("Taxon origin distance matrix:\n")
print(round(originDist_taxon, 1))

cat("Origin sites defined:\n")
cat(" Colony level:", nrow(originSites_col),    "points\n")
cat(" Group level: ", nrow(originSites_group), "points\n")
cat(" Taxon level: ", nrow(originSites_taxon), "points\n")
```

## Target Sites

Non-breeding (wintering) target regions correspond directly to the **WA** (wintering
area) categories recorded in `phen_data` for the original study. The eight named
regions are: Alboran Sea, Western Mediterranean, Central Mediterranean, Adriatic Sea,
Aegean Sea, Black Sea, Saharan Upwelling, and South European Atlantic Shelf.

```{r target-sites-dist}

targetSites <- readRDS("targetSites_MC.rds")

cat("Target sites defined:", nrow(targetSites), "wintering zones\n")
print(targetSites$site_id)

targetDist <- calculate_distance_matrix(targetSites, ocean.r)
colnames(targetDist) <- rownames(targetDist) <- targetSites$site_id

cat("Target distance matrix:\n")
print(round(targetDist, 1))
```
---
# Relative abundances of the origin Sites 

```{r rel-ab}
# === TAXON LEVEL ===
pop_sizes_taxa <- matrix(
  c(originSites_taxon$site_id,c(3000*2, 27610*2)),
  nrow=2,ncol=2
  )
originRelAbund_taxa <- c(
  as.numeric(pop_sizes_taxa[1,2])/sum(as.numeric(pop_sizes_taxa[,2])),
  as.numeric(pop_sizes_taxa[2,2])/sum(as.numeric(pop_sizes_taxa[,2]))
  )

# === POPULATION/GROUP LEVEL ===
pop_sizes_group <- matrix(c(
  originSites_group$site_id,
  c(1685*2+400*2,885*2+4500*2,8500*2,3000*2+300*2)), # CE,CW,EM,WM
  nrow=4,ncol=2)

### Relative abundance ####

originRelAbund_group <- c(
  as.numeric(pop_sizes_group[1,2])/sum(as.numeric(pop_sizes_group[,2])),
  as.numeric(pop_sizes_group[2,2])/sum(as.numeric(pop_sizes_group[,2])),
  as.numeric(pop_sizes_group[3,2])/sum(as.numeric(pop_sizes_group[,2])),
  as.numeric(pop_sizes_group[4,2])/sum(as.numeric(pop_sizes_group[,2]))
  )


# === SUB-POPULATION: WesternMost ===
pop_sizes_WM <- matrix(c(
  originSites_WM$Colony,c(300*2, 530*2, 900*2)),#La Mola, Sa Conillera, Sa Cella
  nrow=3,ncol=2)

originRelAbund_WM <- c(
  as.numeric(pop_sizes_WM[1, 2]) / sum(as.numeric(pop_sizes_WM[, 2])),
  as.numeric(pop_sizes_WM[2, 2]) / sum(as.numeric(pop_sizes_WM[, 2]))
)

# === SUB-POPULATION: Central-West ===
pop_sizes_CW <- matrix(c(
  originSites_CW$Colony,c(4500*2,93*2,189*2)),#Colata,La Mola,Porquerolles,Port-Cros
  nrow=3,ncol=2)

originRelAbund_CW <- c(
  as.numeric(pop_sizes_CW[1, 2]) / sum(as.numeric(pop_sizes_CW[, 2])),
  as.numeric(pop_sizes_CW[2, 2]) / sum(as.numeric(pop_sizes_CW[, 2])),
  as.numeric(pop_sizes_CW[3, 2]) / sum(as.numeric(pop_sizes_CW[, 2])),
  as.numeric(pop_sizes_CW[4, 2]) / sum(as.numeric(pop_sizes_CW[, 2]))
)

# === SUB-POPULATION: Central-East ===
pop_sizes_CE <- matrix(c(
  originSites_CE$Colony,c(200*2,300*2,207*2,34*2)),#Majjistral,Rdum,Zaklopatica,Veli Maslovnjak
  nrow=4,ncol=2)

originRelAbund_CE<- c(
  as.numeric(pop_sizes_CE[1, 2]) / sum(as.numeric(pop_sizes_CE[, 2])),
  as.numeric(pop_sizes_CE[2, 2]) / sum(as.numeric(pop_sizes_CE[, 2])),
  as.numeric(pop_sizes_CE[3, 2]) / sum(as.numeric(pop_sizes_CE[, 2])),
  as.numeric(pop_sizes_CE[4, 2]) / sum(as.numeric(pop_sizes_CE[, 2]))
)

# === SUB-POPULATION: EasternMost  ===
pop_sizes_EM <- matrix(c(
  originSites_EM$Colony,c(1000*2, 1730*2)),#Gyaros, Kappari
  nrow=2,ncol=2)

originRelAbund_EM <- c(
  as.numeric(pop_sizes_EM[1, 2]) / sum(as.numeric(pop_sizes_EM[, 2])),
  as.numeric(pop_sizes_EM[2, 2]) / sum(as.numeric(pop_sizes_EM[, 2]))
)
cat("Relative abundance vectors calculated for all levels.\n")

```
---

# Taxon-Level Analysis {#taxon}

We first estimate MC at the broadest taxonomic level, contrasting the two
Mediterranean shearwater taxa Origin sites are the merged colony polygons
for each taxon; target sites are the wintering zones defined above.

## Prepare Data

```{r taxon-data}
dat_taxon <- prepare_analysis_level(
  tracking_df  = tracking,
  zones_sf = targetSites
)

cat(sprintf("Taxon-level analysis: %d individuals total\n", dat_taxon$n))
cat("Sample sizes by taxon:\n")
print(table(dat_taxon$targetPoints$Taxon))

```

```{r convert-meters-AEQD}
# Convert spatial objects to AEQD (meters)

# originPoints and targetPoints
dat_taxon$originPoints <- st_transform(dat_taxon$originPoints, CRS_AEQD)
dat_taxon$targetPoints <- st_transform(dat_taxon$targetPoints, CRS_AEQD)

# originSites and targetSites
originSites_taxon <- st_transform(originSites_taxon, CRS_AEQD)
targetSites <- st_transform(targetSites, CRS_AEQD)
```

## Estimate MC — Taxon Level

```{r taxon-estMC}
mc_taxon <- estMC(
  isGL         = FALSE, # as we use modeled positions there is no need to add model error
  originDist   = originDist_taxon,
  targetDist   = targetDist,
  originPoints = dat_taxon$originPoints,
  targetPoints = dat_taxon$targetPoints,
  originSites  = originSites_taxon,
  targetSites  = targetSites,
  originNames  = originSites_taxon$site_id,
  targetNames  = targetSites$site_id,
  originRelAbund = originRelAbund_taxa,
  nSim         = 1000,
  nSamples     = 1000,
  verbose      = FALSE
)

cat("=== Taxon-level MC estimate ===\n")
print(mc_taxon)
```
---

# Population-Level Analysis {#populations}

We next estimate MC across the four biogeographic groups, using group-level
dissolved origin polygons.

## Prepare Data

```{r pop-data}
dat_pop <- prepare_analysis_level(
  tracking_df  = tracking,
  pops_include = pop_order,
  zones_sf = targetSites
)

cat(sprintf("Taxon-level analysis: %d individuals total\n", dat_pop$n))
cat("Sample sizes by taxon:\n")
print(table(dat_taxon$targetPoints$Group))

```

```{r convert-meters-AEQD}
# Convert spatial objects to AEQD (meters)

# originPoints and targetPoints
dat_pop$originPoints <- st_transform(dat_pop$originPoints, CRS_AEQD)
dat_pop$targetPoints <- st_transform(dat_pop$targetPoints, CRS_AEQD)

# originSites
originSites_group <- st_transform(originSites_group, CRS_AEQD)
```

## Estimate MC — Population Level

```{r pop-estMC}
mc_pop <- estMC(
  isGL         = FALSE,
  originDist   = originDist_group,
  targetDist   = targetDist,
  originPoints = dat_pop$originPoints,
  targetPoints = dat_pop$targetPoints,
  originSites  = originSites_group,
  targetSites  = targetSites,
  originNames  = originSites_group$site_id,
  targetNames  = targetSites$site_id,
  originRelAbund = originRelAbund_group,
  nSim         = 1000,
  nSamples     = 1000,
  verbose      = FALSE
)

cat("=== Population-level MC estimate ===\n")
print(mc_pop)
```
---

# Sub-Population Analyses {#subpopulations}

Finally, we estimate MC *within* each biogeographic group, treating individual
breeding colonies as origin nodes. This reveals whether connectivity is driven
by among-group or within-group structure.

## WesternMost (Sa Conillera & Sa Cella)

```{r wm-prep}
dat_WM <- prepare_analysis_level(
  tracking_df  = tracking,
  pops_include = group_membership$WesternMost,
  zones_sf = targetSites
)

cat(sprintf("WesternMost: %d individuals\n", dat_WM$n))

# Origin distance
originDist_WM <- calculate_distance_matrix(originSites_WM, ocean.r)
colnames(originDist_WM) <- rownames(originDist_WM) <-  originSites_WM$Colony

cat("Group origin distance matrix:\n")
print(round(originDist_WM, 1))
```

```{r convert-meters-AEQD}
# Convert spatial objects to AEQD (meters)

# originPoints and targetPoints
dat_WM$originPoints <- st_transform(dat_WM$originPoints, CRS_AEQD)
dat_WM$targetPoints <- st_transform(dat_WM$targetPoints, CRS_AEQD)

# originSites
originSites_WM <- st_transform(originSites_WM, CRS_AEQD)
```

```{r wm-estMC}
mc_WM <- estMC(
  isGL         = FALSE,
  originDist = originDist_WM,
  targetDist = targetDist,
  originPoints = dat_WM$originPoints,
  targetPoints = dat_WM$targetPoints,
  originSites  = originSites_WM,
  targetSites  = targetSites,
  originNames = originSites_WM$Colony,
  targetNames = targetSites$site_id,
  originRelAbund = originRelAbund_WM,
  nSim        = 1000,
  nSamples    = 1000,
  verbose      = FALSE
)

cat("=== WesternMost sub-population MC ===\n")
print(mc_WM)
```

## Central-West (MINORCA, FRANCE & SARDINIA) {#centralwest}

```{r cw-prep}
dat_CW <- prepare_analysis_level(
  tracking_df  = tracking,
  pops_include = group_membership$CentralWest,
  zones_sf = targetSites
)

# Origin distance
originDist_CW <- calculate_distance_matrix(originSites_CW, ocean.r)
colnames(originDist_CW) <- rownames(originDist_CW) <-  originSites_CW$Colony

cat("Group origin distance matrix:\n")
print(round(originDist_CW, 1))
```
```{r convert-meters-AEQD}
# Convert spatial objects to AEQD (meters)

# originPoints and targetPoints
dat_CW$originPoints <- st_transform(dat_CW$originPoints, CRS_AEQD)
dat_CW$targetPoints <- st_transform(dat_CW$targetPoints, CRS_AEQD)

# originSites 
originSites_CW <- st_transform(originSites_CW, CRS_AEQD)
```

```{r cw-estMC}
mc_CW <- estMC(
  isGL         = FALSE,
  originDist = originDist_CW,
  targetDist = targetDist,
  originPoints = dat_CW$originPoints,
  targetPoints = dat_CW$targetPoints,
  originSites  = originSites_CW,
  targetSites  = targetSites,
  originNames = originSites_CW$Colony,
  targetNames = targetSites$site_id,
  originRelAbund = originRelAbund_CW,
  nSim         = 1000,
  nSamples     = 1000,
  verbose      = FALSE
)

cat("=== Central-West sub-population MC ===\n")
print(mc_CW)
```

## Central-East (MALTA & CROATIA) {#centraleast}

```{r ce-prep}
dat_CE <- prepare_analysis_level(
  tracking_df  = tracking,
  pops_include = group_membership$CentralEast,
  zones_sf = targetSites
)

cat(sprintf("Central-East: %d individuals\n", dat_CE$n))

# Origin distance
originDist_CE <- calculate_distance_matrix(originSites_CE, ocean.r)
colnames(originDist_CE) <- rownames(originDist_CE) <-  originSites_CE$Colony

cat("Group origin distance matrix:\n")
print(round(originDist_CE, 1))
```
```{r convert-meters-AEQD}
# Convert spatial objects to AEQD (meters)

# originPoints and targetPoints
dat_CE$originPoints <- st_transform(dat_CE$originPoints, CRS_AEQD)
dat_CE$targetPoints <- st_transform(dat_CE$targetPoints, CRS_AEQD)

# originSites 
originSites_CE <- st_transform(originSites_CE, CRS_AEQD)

```

```{r ce-estMC}
mc_CE <- estMC(
  isGL         = FALSE,
  originDist = originDist_CE,
  targetDist = targetDist,
  originPoints = dat_CE$originPoints,
  targetPoints = dat_CE$targetPoints,
  originSites  = originSites_CE,
  targetSites  = targetSites,
  originNames = originSites_CE$Colony,
  targetNames = targetSites$site_id,
  originRelAbund = originRelAbund_CE,
  nSim         = 1000,
  nSamples     = 1000,
  verbose      = FALSE
)

cat("=== Central-East sub-population MC ===\n")
print(mc_CE)
```

## EasternMost (GREECE)

```{r em-prep}
dat_EM <- prepare_analysis_level(
  tracking_df  = tracking,
  pops_include = group_membership$EasternMost,
  zones_sf = targetSites
)

originSites_EM <- originSites_col %>%
  filter(toupper(Pop) %in% group_membership$EasternMost)

cat(sprintf("EasternMost: %d individuals\n", dat_EM$n))

# Origin distance
originDist_EM <- calculate_distance_matrix(originSites_EM, ocean.r)
colnames(originDist_EM) <- rownames(originDist_EM) <-  originSites_EM$Colony

cat("Group origin distance matrix:\n")
print(round(originDist_EM, 1))

```
```{r convert-meters-AEQD}
# Convert spatial objects to AEQD (meters)

# originPoints and targetPoints
dat_EM$originPoints <- st_transform(dat_EM$originPoints, CRS_AEQD)
dat_EM$targetPoints <- st_transform(dat_EM$targetPoints, CRS_AEQD)

# originSites 
originSites_EM <- st_transform(originSites_EM, CRS_AEQD)

```

```{r em-estMC}
mc_EM <- estMC(
  isGL         = FALSE,
  originDist = originDist_EM,
  targetDist = targetDist,
  originPoints = dat_EM$originPoints,
  targetPoints = dat_EM$targetPoints,
  originSites  = originSites_EM,
  targetSites  = targetSites,
  originNames = originSites_EM$Colony,
  targetNames = targetSites$site_id,
  originRelAbund = originRelAbund_EM,
  resampleProjection = CRS_GEO,
  nSim         = 1000,
  nSamples     = 1000,
  verbose      = FALSE
)

cat("=== EasternMost sub-population MC ===\n")
print(mc_EM)
```
---

# Visualization: Circlize Chord Diagram {#circlize}

A chord diagram shows the proportion of tracked individuals from each
breeding population that winter in each target zone. Chord width is
proportional to the number of individuals.

```{r circlize-data}

assignments <- dat_taxon$targetPoints

assignments$n <- 1

pop_sizes <- as.data.frame(matrix(
  c(
    c('IBIZA','MALLORCA','MINORCA','FRANCE',
      'SARDINIA','MALTA','CROATIA','GREECE'),
    c(530*2,900*2, 300*2, 885*2, 4500*2, 1685*2, 400*2,8500*2)),
  nrow=8,ncol=2))

assignments <- assignments %>%
  left_join(.,pop_sizes,by=c('Population'='V1'))%>%
  mutate(pop_size = as.numeric(V2))%>%
  dplyr::select(-V2)

for_circle <- assignments %>% 
  group_by(Group, Population, site_id, pop_size, n_col = n) %>%
  dplyr::summarise(n_col_win = n()) %>% 
  mutate(ind_weight = pop_size/n_col) %>% 
  ungroup()


for_circle_pop <- for_circle %>% 
  dplyr::select(Group, Population, site_id, n_col_win, ind_weight) %>% 
  mutate(circle_arch_value = n_col_win*ind_weight)

for_plot <- for_circle_pop %>% 
  mutate(Population = str_to_title(Population),
         WA = case_when(
           site_id == "South European Atlantic Shelf" ~ "S Euro-Atl Shelf",
           site_id == "Saharan Upwelling" ~ "Saharan Upw",
           site_id == "Western Mediterranean" ~ "Western Med",
           site_id == "Central Mediterranean" ~ "Central Med",
           TRUE ~ site_id
         ))%>%
  dplyr::select(Population, WA, circle_arch_value)

union(for_plot[[1]], for_plot[[2]])

new_order <- c("Ibiza","Mallorca",'Minorca','France','Sardinia','Malta','Croatia','Greece',
               'Black Sea' , 'Levantine Sea','Aegean Sea' ,'Adriatic Sea','Central Med','Western Med', 
               'Alboran Sea','Saharan Upw', 'S Euro-Atl Shelf')

grid.col <-  c("Ibiza" = "#fde725",'Mallorca'='#dde318','Minorca' = "#bade28",'France' = "#95d840",'Sardinia' = "#75d054"
                            ,'Malta' = "#56c667",'Croatia'= "#3dbc74",'Greece'= "#29af7f",
                            'Black Sea' = "#21918c", 'Levantine Sea' = "#2b758e",'Aegean Sea' = "#33638d",'Adriatic Sea' = '#39558c','Central Med' = '#404688','Western Med' = "#453781", 
                            'Alboran Sea' = '#482576','Saharan Upw' = "#481467", 'S Euro-Atl Shelf' = "#440154")

```

```{r circlize-plot, fig.width=9, fig.height=9}

# Configurar parámetros
circos.par(start.degree = 170, clock.wise = TRUE,
           gap.after = c(rep(1, length(unique(for_plot[[1]])) - 1), 7,
                         rep(1, length(unique(for_plot[[2]])) - 1), 7))

# Crear chord diagram
chordDiagram(for_plot, reduce = -1, order = new_order, grid.col = grid.col, 
             link.sort = T, link.decreasing = T, link.zindex = rank(for_plot[[3]]),
             directional = 1, direction.type = c("diffHeight", "arrows"), 
             link.arr.type = "big.arrow", 
             annotationTrack = c("grid"), annotationTrackHeight = c(0.05), 
             preAllocateTracks = list(track.height = 0.5))

# 1. Grupo Ibiza-Mallorca
highlight.sector(sector.index = c("Ibiza", "Mallorca"), track.index = 1,
                 col = NA, border = "black", lwd = 2,
                 padding = c(0.2, 0, -1, 0))

# 2. Grupo Minorca-France-Sardinia
highlight.sector(sector.index = c("Minorca", "France", "Sardinia"), track.index = 1,
                 col = NA, border = "black", lwd = 2,
                 padding = c(0.2, 0, -1, 0))

# 3. Grupo Malta-Croatia
highlight.sector(sector.index = c("Malta", "Croatia"), track.index = 1,
                 col = NA, border = "black", lwd = 2,
                 padding = c(0.2, 0, -1, 0))

# 4. Grupo Greece
highlight.sector(sector.index = "Greece", track.index = 1,
                 col = NA, border = "black", lwd = 2,
                 padding = c(0.2, 0, -1, 0))

# Añadir labels
circos.track(track.index = 1, panel.fun = function(x, y) {
  circos.text(CELL_META$xcenter, CELL_META$ylim[1], CELL_META$sector.index,
              facing = "clockwise", niceFacing = TRUE, adj = c(0, 0.1), 
              cex = 0.8)
}, bg.border = NA)

# Limpiar parámetros de circos
circos.clear()
```

---

# Session Information

```{r session-info}
sessionInfo()
```
