---
title: "Environmental Niche Models: Presence Data Preparation"
subtitle: "Hierarchical weighted resampling from KDE-simulated distributions"
author: "Mario Gómez Álvarez"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    theme: cosmo
    highlight: tango
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 8
)
```

# Overview

This analysis prepares presence datasets for Environmental Niche Modeling (ENM) by:

1. Creating spatial masks for breeding and non-breeding distributions
2. Generating KDE-weighted pseudo-absence/resampled points from trip observations
3. Applying hierarchical weighting at multiple levels:
   - **Population level** (`w_pop`): Proportional to population size within species
   - **Group level** (`w_group`): Proportional to population size within geographic groups
   - **Species level** (`w_sp`): Proportional to species abundance

The resulting datasets are thinned to reflect these weights, ensuring balanced representation across taxonomic and geographic scales.

---

# Setup and Data Loading

## Library Dependencies

```{r libraries}
library(tidyverse)
library(spatstat)
library(raster)
library(sf)
library(rnaturalearth)
library(spatstat.geom)
library(adehabitatHR)
```

## Load Raw Trip Data

```{r load-data}
setwd('C:/Users/mario/OneDrive - Universitat de Barcelona/R_code')

trips <- read_csv(
  "C:/Users/mario/OneDrive - Universitat de Barcelona/Escritorio/github puffinus/all_trips_classified.csv"
)

head(trips)
```

---

# Spatial Masks

## Purpose

Spatial masks confine KDE-based resampling to ecologically relevant regions:
- **`final_mask`**: Breeding season (Mediterranean + Atlantic)
- **`r_mask_gen`**: Non-breeding season (West Africa pelagic zone)
- **`r_mask_pyel`**: Levantine basin (PUFYEL-specific region)

## Create Base Raster and World Boundaries

```{r create-raster}
# Global raster template (0.25° resolution)
world <- ne_countries(scale = 50, returnclass = "sf")
rast <- raster(xmn = -15, xmx = 42, ymn = 28, ymx = 51, res = 0.25)
r <- rasterize(world, rast)

# CRS definition for all masks
crs_string <- '+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs +towgs84=0,0,0'
crs_object <- CRS(crs_string)
```

## Define Mask Polygons

```{r create-mask-polygons}
# Left boundary (Western Atlantic cutoff)
mask_polygon_left <- st_polygon(
  list(cbind(
    c(-180, -5.6, -5.6, -180, -180),
    c(-90, -90, 90, 90, -90)
  ))
)

# Top boundary (Northern cutoff)
mask_polygon_top <- st_polygon(
  list(cbind(
    c(-180, 2, 2, -180, -180),
    c(43, 43, 90, 90, 43)
  ))
)

# Bottom boundary (Southern cutoff, excluding sub-Saharan Africa)
mask_polygon_down <- st_polygon(
  list(cbind(
    c(32, 180, 180, 32, 32),
    c(-90, -90, 31, 31, -90)
  ))
)

# Levantine basin (Eastern Mediterranean)
mask_polygon_levantine <- st_polygon(
  list(cbind(
    c(27, 180, 180, 27, 27),
    c(-90, -90, 40, 40, -90)
  ))
)

# Convert to sf objects with CRS
convert_mask_to_sf <- function(poly) {
  poly_sf <- st_sfc(poly)
  st_crs(poly_sf) <- 4326
  st_as_sf(poly_sf)
}

mask_polygon_left <- convert_mask_to_sf(mask_polygon_left)
mask_polygon_top <- convert_mask_to_sf(mask_polygon_top)
mask_polygon_down <- convert_mask_to_sf(mask_polygon_down)
mask_polygon_levantine <- convert_mask_to_sf(mask_polygon_levantine)
```

## Rasterize and Combine Masks

```{r rasterize-masks}
# Rasterize individual mask polygons
r1 <- rasterize(mask_polygon_left, rast)
r2 <- rasterize(mask_polygon_top, rast)
r3 <- rasterize(mask_polygon_down, rast)
r4 <- rasterize(mask_polygon_levantine, rast)

# Breeding mask: exclude all three boundaries
final_mask <- overlay(r, r1, r2, r3, fun = function(x, y, z, t) {
  ifelse(x != 0 | y != 0 | z != 0 | t != 0, 0, 1)
})
proj4string(final_mask) <- crs_object

# Non-breeding mask for Western populations (West Africa pelagic)
r_mask_gen <- overlay(r, r3, fun = function(x, y) {
  ifelse(x != 0 | y != 0, 0, 1)
})
proj4string(r_mask_gen) <- crs_object

# Levantine mask (Central/Eastern populations)
s <- stack(r, r1, r2, r3, r4)
r_mask_pyel <- calc(s, fun = function(x) {
  ifelse(any(x != 0), 0, 1)
})
proj4string(r_mask_pyel) <- crs_object

# Visualize masks
par(mfrow = c(2, 2), mar = c(3, 3, 2, 1))
plot(final_mask, main = "Breeding Mask", cex.main = 1.2)
plot(r_mask_gen, main = "Non-breeding Mask (Western)", cex.main = 1.2)
plot(r_mask_pyel, main = "Levantine Mask", cex.main = 1.2)
par(mfrow = c(1, 1))
```

## Create Inverted Masks for Sampling

```{r invert-masks}
# Invert masks: 0 → NA, 1 → 1 (for sampling compatibility)
mask_inv_1 <- final_mask
mask_inv_1[final_mask == 0] <- NA
mask_inv_1[is.na(final_mask)] <- 0

mask_inv_2 <- r_mask_gen
mask_inv_2[r_mask_gen == 0] <- NA
mask_inv_2[is.na(r_mask_gen)] <- 0
```

---

# Population and Group Sizes

## Define Population-Level Data

```{r pop-sizes}
pop_sizes <- as.data.frame(matrix(
  c(
    c('Ibiza', 'Mallorca', 'Minorca', 'France', 
      'Sardinia', 'Malta', 'Croatia', 'Greece'),
    c(530*2, 900*2, 300*2, 585*2, 
      11050*2, 1685*2, 425*2, 13000*2),
    c('PUFMAU', 'PUFMAU', 'PUFYEL', 'PUFYEL', 
      'PUFYEL', 'PUFYEL', 'PUFYEL', 'PUFYEL')
  ),
  nrow = 8, ncol = 3
))

pop_sizes <- pop_sizes %>%
  rename(Population = V1, pop_size = V2, Sp = V3) %>%
  mutate(pop_size = as.numeric(pop_size)) %>%
  group_by(Sp) %>%
  mutate(w_pop = pop_size / sum(pop_size)) %>%
  ungroup()

knitr::kable(pop_sizes, caption = "Population sizes and weights (w_pop) by species")
```

## Define Geographic Group Data

```{r group-sizes}
group_sizes <- as.data.frame(matrix(
  c(
    c('Ibiza', 'Mallorca', 'Minorca', 'France', 
      'Sardinia', 'Malta', 'Croatia', 'Greece'),
    c(530*2, 900*2, 300*2, 585*2, 
      11050*2, 1685*2, 425*2, 13000*2),
    c('WesternMost', 'WesternMost', 'WesternMost', 'Central-West',
      'Central-West', 'Central-East', 'Central-East', 'EasternMost')
  ),
  nrow = 8, ncol = 3
))

group_sizes <- group_sizes %>%
  rename(Population = V1, pop_size = V2, Group = V3) %>%
  mutate(pop_size = as.numeric(pop_size)) %>%
  group_by(Group) %>%
  mutate(w_group = pop_size / sum(pop_size)) %>%
  ungroup()

knitr::kable(group_sizes, caption = "Population sizes and weights (w_group) by geographic group")
```

## Define Species-Level Data

```{r sp-sizes}
sp_sizes <- as.data.frame(matrix(
  c(
    c('PUFMAU', 'PUFYEL'),
    c(3000*2, 27310*2),
    c('Supra-taxa', 'Supra-taxa')
  ),
  nrow = 2, ncol = 3
))

sp_sizes <- sp_sizes %>%
  rename(Sp = V1, sp_size = V2) %>%
  mutate(
    sp_size = as.numeric(sp_size),
    w_sp = sp_size / sum(sp_size)
  )

knitr::kable(sp_sizes, caption = "Species sizes and weights (w_sp)")
```

---

# Mask Selection Function

## Dynamic Mask Selection

This function selects the appropriate spatial mask based on breeding/non-breeding period and population group:

```{r get-mask-function}
get_mask <- function(gr, breeding = TRUE) {
  # Breeding: use breeding mask
  if (breeding) return(mask_inv_1)
  
  # Non-breeding: Western populations use West Africa pelagic mask
  if (gr == "WesternMost" & !breeding) return(mask_inv_2)
  
  # Non-breeding: Central/Eastern populations use Mediterranean breeding mask
  if (gr != "WesternMost" & !breeding) return(mask_inv_1)
}
```

---

# KDE-Weighted Sampling Functions

## Kernel Density Estimation Sampling

This function generates weighted pseudo-absence points by:

1. Converting coordinates to UTM
2. Computing 2D kernel density estimate from trip observations
3. Applying spatial mask
4. Sampling points weighted by KDE probability
5. Converting back to geographic coordinates

```{r kde-function}
sample_kde_weighted <- function(df_coords, n = 60, mask_raster = NULL,
                                utm_zone = 31, crs_longlat = 4326) {
  
  # --------------------------------------------------
  # 1. Coordenadas → sf → UTM
  # --------------------------------------------------
  crs_utm <- paste0(
    "+proj=utm +zone=", utm_zone, " +datum=WGS84 +units=m +no_defs"
  )
  
  pts_sf <- st_as_sf(df_coords, coords = c("Lon", "Lat"), crs = crs_longlat) |> 
    st_transform(crs_utm)
  
  sp_pts <- as(pts_sf, "Spatial")
  sp_pts$id <- factor(1)
  
  # --------------------------------------------------
  # 2. KDE (kernelUD returns estUDm)
  # --------------------------------------------------
  ud <- try(kernelUD(sp_pts[, "id"], h = "href", grid = 200), silent = TRUE)
  if (inherits(ud, "try-error")) return(NULL)
  
  ud1 <- ud[[1]]  # extract single UD
  r_ud <- raster(ud1)  # KDE raster in UTM
  
  # --------------------------------------------------
  # 3. Project and adjust mask
  # --------------------------------------------------
  if (!is.null(mask_raster)) {
    mask_utm <- projectRaster(mask_raster, r_ud, method = "ngb")
    mask_utm <- resample(mask_utm, r_ud, method = "ngb")
    r_ud <- mask(r_ud, mask_utm, maskvalue = NA)
  }
  
  # --------------------------------------------------
  # 4. Raster → im (spatstat)
  # --------------------------------------------------
  m <- as.matrix(r_ud)
  m <- m[nrow(m):1, ]  # flip rows
  e <- extent(r_ud)
  
  im_ud <- im(
    m,
    xcol = seq(e@xmin, e@xmax, length.out = ncol(m)),
    yrow = seq(e@ymin, e@ymax, length.out = nrow(m)),
    unitname = "m"
  )
  
  # --------------------------------------------------
  # 5. Weighted point sampling
  # --------------------------------------------------
  sim_ppp <- try(rpoint(n, im_ud), silent = TRUE)
  if (inherits(sim_ppp, "try-error")) return(NULL)
  
  sim_df <- as.data.frame(sim_ppp)
  
  # --------------------------------------------------
  # 6. Convert back to lon/lat
  # --------------------------------------------------
  sim_sf <- st_as_sf(sim_df, coords = c("x", "y"), crs = crs_utm) |>
    st_transform(crs_longlat)
  
  data.frame(
    Longitude = st_coordinates(sim_sf)[, 1],
    Latitude = st_coordinates(sim_sf)[, 2]
  )
}
```

## Thinning Function

```{r thinning-function}
thinning_weighted <- function(df, target_n) {
  if (nrow(df) <= target_n) return(df)
  df |> slice_sample(n = target_n)
}
```

---

# Generate Presence Data

## Base Presence Points from KDE Sampling

```{r generate-presence-points}
setwd('C:/Users/mario/OneDrive - Universitat de Barcelona/Escritorio')

desired_n <- 60
pres_list <- list()

for (p in c("breeding", "non-breeding")) {
  
  # Filter by period
  df_month <- trips |> filter(period == p)
  
  # Iterate over unique trips within period
  for (trip in unique(df_month$geo_trip)) {
    print(glue::glue("Processing: {p} - {trip}"))
    
    trip_df <- df_month |> filter(geo_trip == trip)
    
    # Skip trips with insufficient observations
    if (nrow(trip_df) < desired_n) next
    
    # Get group and select appropriate mask
    gr <- unique(trip_df$Group)
    mask_use <- get_mask(gr, p == "breeding")
    
    # Sample points from KDE
    pts <- sample_kde_weighted(
      trip_df |> dplyr::select(Lon, Lat),
      n = desired_n,
      mask_raster = mask_use
    )
    
    # Add metadata
    pres_list[[length(pres_list) + 1]] <-
      pts |> mutate(
        geo_trip = trip,
        Population = unique(trip_df$Population),
        Group = gr,
        Sp = unique(trip_df$Sp),
        period = p
      )
  }
}

pres_df <- bind_rows(pres_list)

cat("Total presence points generated:", nrow(pres_df), "\n")
cat("Trips represented:", n_distinct(pres_df$geo_trip), "\n")
```

---

# Hierarchical Weighting and Thinning

## Level 1: Population-Level Weighting (`w_pop`)

Ensure representation of each population is proportional to its size **within each species**:

```{r thinning-population}
taxa_df <- pres_df %>%
  # Join population-level weights
  left_join(
    pop_sizes %>% dplyr::select(Population, w_pop),
    by = "Population"
  ) %>%
  # Group by period, species, and population to apply weights uniquely
  group_by(period, Sp, Population) %>%
  group_modify(~ {
    # Calculate target sample size based on population weight
    target <- ceiling(nrow(.x) * unique(.x$w_pop))
    thinning_weighted(.x, target)
  }) %>%
  ungroup()

cat("After population-level thinning:", nrow(taxa_df), "points\n")
```

## Level 2: Geographic Group Weighting (`w_group`)

Ensure representation of each geographic group is proportional to its population size:

```{r thinning-group}
group_df <- pres_df %>%
  # Join group-level weights
  left_join(
    group_sizes %>% dplyr::select(Population, Group, w_group),
    by = c("Population", "Group")
  ) %>%
  # Group by period, geographic group, and population
  group_by(period, Group, Population) %>%
  group_modify(~ {
    # Calculate target sample size based on group weight
    target <- ceiling(nrow(.x) * unique(.x$w_group))
    thinning_weighted(.x, target)
  }) %>%
  ungroup()

cat("After group-level thinning:", nrow(group_df), "points\n")
```

## Level 3: Species-Level Weighting (`w_sp`)

Balance representation between the two puffin species proportional to their abundance:

```{r thinning-species}
suprataxa_df <- taxa_df %>%
  # Join species-level weights
  left_join(
    sp_sizes %>% dplyr::select(Sp, w_sp),
    by = "Sp"
  ) %>%
  # Group by period and species
  group_by(period, Sp) %>%
  group_modify(~ {
    # Calculate target sample size based on species weight
    target <- ceiling(nrow(.x) * unique(.x$w_sp))
    thinning_weighted(.x, target)
  }) %>%
  ungroup()

cat("After species-level thinning:", nrow(suprataxa_df), "points\n")
```

---

# Summary Statistics

```{r summary-stats}
cat("=== FINAL DATASET SUMMARIES ===\n\n")

cat("1. POPULATION LEVEL (taxa_df):\n")
print(taxa_df %>%
  group_by(period, Sp) %>%
  summarize(n_points = n(), .groups = "drop"))

cat("\n2. GROUP LEVEL (group_df):\n")
print(group_df %>%
  group_by(period, Group) %>%
  summarize(n_points = n(), .groups = "drop"))

cat("\n3. SPECIES LEVEL (suprataxa_df):\n")
print(suprataxa_df %>%
  group_by(period, Sp) %>%
  summarize(n_points = n(), .groups = "drop"))
```

---

# Export Results

## Prepare and Save Datasets

```{r export-data}
setwd("C:/Users/mario/OneDrive - Universitat de Barcelona/Escritorio/Habitat_modelling/month_positions/presences")

# Population-level presence data
pres_df_export <- pres_df %>%
  dplyr::select(Longitude, Latitude, geo_trip, Population, period)

saveRDS(pres_df_export,
        file.path("pres_by_population_thinned.RDS"))

cat("✓ Saved: pres_by_population_thinned.RDS\n")

# Geographic group-level data
group_df_export <- group_df %>%
  dplyr::select(Longitude, Latitude, geo_trip, Group, period)

saveRDS(group_df_export,
        file.path("pres_by_suprapopulation_thinned.RDS"))

cat("✓ Saved: pres_by_suprapopulation_thinned.RDS\n")

# Species-level data
taxa_df_export <- taxa_df %>%
  dplyr::select(Longitude, Latitude, geo_trip, Sp, period)

saveRDS(taxa_df_export,
        file.path("pres_by_sp_thinned.RDS"))

cat("✓ Saved: pres_by_sp_thinned.RDS\n")

# Supra-taxa (final) data
suprataxa_df_export <- suprataxa_df %>%
  dplyr::select(Longitude, Latitude, geo_trip, Sp, period)

saveRDS(suprataxa_df_export,
        file.path("pres_by_suprataxa_thinned.RDS"))

cat("✓ Saved: pres_by_suprataxa_thinned.RDS\n")

cat("\n=== ALL EXPORTS COMPLETE ===\n")
```

---

# Session Information

```{r session-info}
sessionInfo()
```
