---
title: "Environmental Climatologies Preparation for ENM"
subtitle: "Monthly climatology processing from daily grids (2001-2024)"
author: "Charlotte Lambert & Mario Gómez Álvarez"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    theme: cosmo
    highlight: tango
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 8,
  cache = FALSE
)
```

# Overview

This analysis prepares environmental climatology layers as predictors for Environmental Niche Models (ENM) by:

1. **Processing daily NetCDF grids** (2001-2024) from oceanographic and meteorological data
2. **Extracting period-specific climatologies** aligned with our tracking data
3. **Computing spatial and temporal statistics** (mean and standard deviation)
4. **Generating gradient layers** for selected variables (spatial heterogeneity)
5. **Averaging individual-level layers into single predictors to train ENMs


**Key variables processed:**
- Chlorophyll-a (CHLa), gradient (CHLag)
- Net Primary Productivity (NPP), gradient (NPPg)
- Sea Surface Temperature (SST), gradient (SSTg)
- Sea Surface Salinity (SSS), gradient (SSSg)
- Mixed Layer Depth (MLD)
- Sea Surface Height (SSH)
- Wind Speed
- Wind stress curl (gIUV)

---

# Setup and Configuration

## Library Dependencies

```{r libraries}
library(tidyverse)      # Data manipulation
library(terra)          # Raster processing
library(sf)             # Spatial objects
library(ncdf4)          # NetCDF file handling
library(tidyterra)      # Terra + tidyverse integration
library(cli)            # Command-line interface for messaging
library(grec)           # Gradient computation
```

## System Configuration

```{r system-config}
# Set locale to English for consistent date handling
Sys.setlocale("LC_TIME", "English")

# Set working directory
wd <- 'C:/Users/mario/'
setwd(wd)

# Define utility function
'%!in%' <- function(x, y) !('%in%'(x, y))
```

---

# Data Loading

## NetCDF Files and Tracking Data

```{r load-files}
# Load all NetCDF files matching pattern
files <- list.files(path = wd, pattern = ".nc")
cat("Found", length(files), "NetCDF files:\n")
print(files)

# Define target extent (Mediterranean and adjacent areas)
target_extent <- ext(-15, 42, 28, 51)

# Create template raster for standardization
template_raster <- rast(
  extent = target_extent,
  resolution = 0.25,
  crs = "EPSG:4326"
)
values(template_raster) <- NA

cat("\nTemplate raster resolution:", res(template_raster)[1], "degrees\n")
```

## Trip Classification Data

Load trip data and create period-specific date ranges:

```{r load-trips}
# Load classified trip data
trips_all <- read_csv(
  "OneDrive - Universitat de Barcelona/Escritorio/github puffinus/all_trips_classified.csv"
)

# Separate by breeding/non-breeding
trips_b <- trips_all %>%
  filter(period == "breeding")

trips_nb <- trips_all %>%
  filter(period == "non-breeding")

cat("Breeding trips:", n_distinct(trips_b$geo_trip), "\n")
cat("Non-breeding trips:", n_distinct(trips_nb$geo_trip), "\n")
```

## Define Temporal Periods per Trip

```{r define-periods}
# Extract date ranges for each trip
periods <- bind_rows(
  trips_nb %>%
    group_by(geo_trip) %>%
    summarise(
      first_date = min(as.Date(paste0(year_dep, "-", date_month))),
      last_date = max(as.Date(paste0(year_dep, "-", date_month))),
      period_type = "non-breeding",
      .groups = "drop"
    ),
  trips_b %>%
    group_by(geo_trip) %>%
    summarise(
      first_date = min(as.Date(paste0(year_dep, "-", date_month))),
      last_date = max(as.Date(paste0(year_dep, "-", date_month))),
      period_type = "breeding",
      .groups = "drop"
    )
)

knitr::kable(
  head(periods, 10),
  caption = "Sample of temporal periods extracted from tracking data"
)
```

---

# Variable Configuration

## Variable Dictionary

Create a configuration table mapping variable names, NetCDF short names, and processing options:

```{r var-config}
vars_list <- data.frame(
  var = c(
    "CHLa", "CHLag", "NPP", "NPPg", "SST", "SSTg",
    "SSS", "SSSg", "MLD", "SSH", "Wind", "gIUV"
  ),
  var_short = c(
    "chl", "chl", "nppv", "nppv", "thetao", "thetao",
    "so", "so", "mlotst", "zos", "wind_speed", "uo"
  ),
  archivo_idx = c(1, 1, 1, 1, 3, 3, 3, 3, 2, 2, 5, 4),
  gradient = c(
    FALSE, TRUE, FALSE, TRUE, FALSE, TRUE,
    FALSE, TRUE, FALSE, FALSE, FALSE, FALSE
  ),
  description = c(
    "Chlorophyll-a concentration",
    "Chlorophyll-a spatial gradient",
    "Net Primary Productivity",
    "NPP spatial gradient",
    "Sea Surface Temperature",
    "SST spatial gradient",
    "Sea Surface Salinity",
    "SSS spatial gradient",
    "Mixed Layer Depth",
    "Sea Surface Height",
    "Wind Speed",
    "Wind stress curl (magnitude U,V)"
  )
)

knitr::kable(vars_list, caption = "Environmental variables configuration")
```

---

# Core Processing Function

## Habitat Variable Processing

This function handles the complete processing pipeline for each environmental variable:

```{r process-function}
process_habitat_vars <- function(v_idx, v_short, v_long, is_gradient,
                                track_period, template) {
  
  file_path <- files[v_idx]
  
  # ---------------------------------------------------------------
  # 1. TEMPORAL FILTERING
  # ---------------------------------------------------------------
  if (v_long == "gIUV") {
    # Wind stress curl: compute magnitude from U and V components
    u_raw <- rast(file_path, subds = "uo")
    v_raw <- rast(file_path, subds = "vo")
    
    # Extract time indices within track period
    idx <- which(
      time(u_raw) >= track_period$first_date &
      time(u_raw) <= track_period$last_date
    )
    
    if (length(idx) == 0) return(NULL)
    
    # Compute magnitude: sqrt(U^2 + V^2)
    r_period <- sqrt(u_raw[[idx]]^2 + v_raw[[idx]]^2)
  } else {
    # Standard variable extraction
    r_raw <- rast(file_path, subds = v_short)
    
    idx <- which(
      time(r_raw) >= track_period$first_date &
      time(r_raw) <= track_period$last_date
    )
    
    if (length(idx) == 0) return(NULL)
    
    r_period <- r_raw[[idx]]
  }
  
  # ---------------------------------------------------------------
  # 2. SPATIAL ADJUSTMENT
  # ---------------------------------------------------------------
  # Aggregate to template resolution if necessary
  if (res(r_period)[1] < res(template)[1]) {
    fact <- round(res(template)[1] / res(r_period)[1])
    r_period <- aggregate(r_period, fact = fact, fun = mean, na.rm = TRUE)
  }
  
  # Crop and resample to template extent
  r_period <- crop(r_period, ext(template), snap = "out")
  r_period <- resample(r_period, template, method = "near")
  
  # ---------------------------------------------------------------
  # 3. DEPTH COLLAPSE (for CHLa and NPP)
  # ---------------------------------------------------------------
  # These variables have multiple depth layers; average them
  if (v_idx == 1) {
    n_depths <- 12  # Number of depth levels
    n_days <- nlyr(r_period) / n_depths
    r_period <- tapp(
      r_period,
      index = rep(1:n_days, each = n_depths),
      fun = mean,
      na.rm = TRUE
    )
  }
  
  # ---------------------------------------------------------------
  # 4. GRADIENT COMPUTATION
  # ---------------------------------------------------------------
  # For gradient variables, compute spatial heterogeneity
  if (is_gradient) {
    r_list <- lapply(1:nlyr(r_period), function(i) {
      layer_i <- r_period[[i]]
      
      # Check if layer has data
      if (global(layer_i, "notNA")[1, 1] > 0) {
        grec::getGradients(layer_i)
      } else {
        # Return empty layer if no data
        setValues(template, NA)
      }
    })
    
    r_period <- rast(r_list)
  }
  
  # ---------------------------------------------------------------
  # 5. MASKING AND STATISTICS
  # ---------------------------------------------------------------
  # Apply template mask (ocean only)
  r_period <- mask(r_period, template, inverse = TRUE)
  
  # Compute mean and standard deviation across time
  m_i <- mean(r_period, na.rm = TRUE)
  s_i <- stdev(r_period, na.rm = TRUE)
  
  return(c(m_i, s_i))
}
```

---

# Main Processing Loop

## Execute Variable Processing

Process all environmental variables for all tracked trips:

```{r process-loop}
# Output directory
path_out <- "C:/Users/mario/OneDrive - Universitat de Barcelona/Escritorio/Habitat_modelling/vars/breeding_climatology"

if (!dir.exists(path_out)) {
  dir.create(path_out, recursive = TRUE)
}

# Process each variable
for (i in 1:nrow(vars_list)) {
  
  v <- vars_list[i, ]
  cli::cli_h1("PROCESSING VARIABLE: {v$var}")
  cli::cli_alert("Description: {v$description}")
  
  # List to store individual results
  all_indiv_layers <- list()
  
  # Process each tracked trip
  for (id in periods$geo_trip) {
    cli::cli_alert_info("Computing for trip: {id}")
    
    # Get period for current trip
    track <- periods %>% filter(geo_trip == id)
    
    # Process variable for this individual
    res_indiv <- process_habitat_vars(
      v_idx = v$archivo_idx,
      v_short = v$var_short,
      v_long = v$var,
      is_gradient = v$gradient,
      track_period = track,
      template = template_raster
    )
    
    if (!is.null(res_indiv)) {
      all_indiv_layers[[id]] <- res_indiv
    }
    
    gc()  # Memory cleanup
  }
  
  # ---------------------------------------------------------------
  # FINAL AGGREGATION
  # ---------------------------------------------------------------
  if (length(all_indiv_layers) > 0) {
    # Stack all individual results
    stack_final <- rast(all_indiv_layers)
    
    # Separate mean and sd layers
    idx_means <- seq(1, nlyr(stack_final), by = 2)
    idx_sds <- seq(2, nlyr(stack_final), by = 2)
    
    # 1. Global mean climatology across all trips
    final_mean <- mean(stack_final[[idx_means]], na.rm = TRUE)
    
    # 2. Mean temporal variability
    # (indicates how dynamic conditions are at each pixel)
    final_sd_temporal <- mean(stack_final[[idx_sds]], na.rm = TRUE)
    
    # Prepare output raster
    res_raster <- c(final_mean, final_sd_temporal)
    names(res_raster) <- c("mean_global", "sd_temporal_mean")
    
    # Save output
    output_file <- file.path(path_out, paste0(v$var, "_b_complete.tif"))
    writeRaster(res_raster, output_file, overwrite = TRUE)
    
    cli::cli_alert_success("Saved: {basename(output_file)}")
  }
}

cli::cli_h2("PROCESSING COMPLETE!")
```

---

# Output Validation and Assembly

## Verify Output Files

```{r validate-output}
# List generated climatology files
output_files <- list.files(
  "C:/Users/mario/OneDrive - Universitat de Barcelona/Escritorio/Habitat_modelling/vars/breeding_climatology",
  full.names = TRUE
)

cat("Generated", length(output_files), "climatology files:\n")
print(basename(output_files))
```

## Stack All Variables

Create a final stack containing all environmental predictors:

```{r stack-variables}
# Load all climatology files
climatology_path <- "C:/Users/mario/OneDrive - Universitat de Barcelona/Escritorio/Habitat_modelling/vars/breeding_climatology"
vars_climatology <- list.files(climatology_path, pattern = ".tif", full.names = TRUE)

# Load additional external variables if they exist
external_vars <- list.files(
  "C:/Users/mario/OneDrive - Universitat de Barcelona/Escritorio/Habitat_modelling/vars",
  pattern = ".tif",
  full.names = TRUE
)

# Combine all variables
all_var_paths <- c(vars_climatology, external_vars)

# Create final stack
stack_final <- rast(all_var_paths)

cat("Final predictor stack:\n")
cat("Layers:", nlyr(stack_final), "\n")
cat("Extent:", paste(ext(stack_final)), "\n")
cat("Resolution:", res(stack_final)[1], "degrees\n")

# Verify geometric compatibility
tryCatch(
  compareGeom(stack_final),
  error = function(e) {
    cli::cli_alert_warning("Geometry mismatch detected:")
    print(e)
  }
)
```

## Visualize Predictor Layers

```{r visualize-stack}
# Plot climatology layers
plot(rast(vars_climatology), maxcell = 100000)

# Summary statistics
cat("\nPredictor Stack Summary:\n")
print(summary(stack_final))
```

---

## Output Files

All climatology layers are saved in:
```
C:/Users/mario/OneDrive - Universitat de Barcelona/Escritorio/Habitat_modelling/vars/breeding_climatology/
```

**File format:** GeoTIFF (2 bands per variable: mean + temporal SD)

---

# Session Information

```{r session-info}
sessionInfo()
```
