---
title: "Environmental Climatologies Preparation for ENM"
subtitle: "Monthly climatology processing from daily grids - Breeding vs Non-Breeding"
author: "Charlotte Lambert & Mario Gómez Álvarez"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    theme: cosmo
    highlight: tango
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 12,
  fig.height = 8,
  cache = FALSE
)
```

# Overview

This analysis prepares environmental climatology layers as predictors for Environmental Niche Models (ENM) by:

1. **Processing daily NetCDF grids** (2001-2024) from oceanographic and meteorological data
2. **Separating period-specific climatologies** for breeding and non-breeding seasons
3. **Extracting climatologies aligned with individual bird tracking data** for each period
4. **Computing spatial and temporal statistics** (mean and standard deviation)
5. **Generating gradient layers** for selected variables (spatial heterogeneity)
6. **Creating individual-level predictor matrices** for ENM training (separate by season)

**Key variables processed:**
- Chlorophyll-a (CHLa), gradient (CHLag). https://doi.org/10.48670/moi-00019 
- Net Primary Productivity (NPP), gradient (NPPg). https://doi.org/10.48670/moi-00019 
- Sea Surface Temperature (SST), gradient (SSTg). https://doi.org/10.48670/moi-00024
- Sea Surface Salinity (SSS), gradient (SSSg). https://doi.org/10.48670/moi-00024
- Mixed Layer Depth (MLD). https://doi.org/10.48670/moi-00021
- Sea Surface Height (SSH). https://doi.org/10.48670/moi-00021
- Wind Speed. https://doi.org/10.48670/moi-00181
- Wind stress curl (gIUV). https://doi.org/10.48670/moi-00021

---

# Setup and Configuration

## Library Dependencies

```{r libraries}
library(tidyverse)      
library(terra)         
library(sf)             
library(ncdf4)          
library(tidyterra)      
library(cli)           
library(grec)          
```

## System Configuration

```{r system-config}
# Set locale to English for consistent date handling
Sys.setlocale("LC_TIME", "English")

# Set working directory with raw files
wd <- here::here()
setwd(wd)

# Define utility function
'%!in%' <- function(x, y) !('%in%'(x, y))
```

---

# Data Loading

## NetCDF Files and Tracking Data

```{r load-files}
# Load all NetCDF files matching pattern
files <- list.files(path = wd, pattern = ".nc")
cat("Found", length(files), "NetCDF files:\n")
print(files)

# Define target extent (Mediterranean and adjacent areas)
target_extent <- ext(-15, 42, 28, 51)

# Create template raster for standardization
template_raster <- rast(
  extent = target_extent,
  resolution = 0.25,
  crs = "EPSG:4326"
)
values(template_raster) <- NA

cat("\nTemplate raster resolution:", res(template_raster)[1], "degrees\n")
```

## Trip Classification Data

Load trip data and create period-specific date ranges:

```{r load-trips}
# Load classified trip data
trips_all <- read_csv("all_trips_classified.csv")

# Separate by breeding/non-breeding
trips_b <- trips_all %>%
  filter(period == "breeding")

trips_nb <- trips_all %>%
  filter(period == "non-breeding")

cat("Breeding trips:", n_distinct(trips_b$geo_trip), "\n")
cat("Non-breeding trips:", n_distinct(trips_nb$geo_trip), "\n")
```

## Define Temporal Periods per Trip (Separate by Season)

```{r define-periods}
# Extract date ranges for breeding trips
periods_breeding <- trips_b %>%
  group_by(geo_trip) %>%
  summarise(
    first_date = min(as.Date(paste0(year_dep, "-", date_month))),
    last_date = max(as.Date(paste0(year_dep, "-", date_month))),
    period_type = "breeding",
    .groups = "drop"
  )

# Extract date ranges for non-breeding trips
periods_nonbreeding <- trips_nb %>%
  group_by(geo_trip) %>%
  summarise(
    first_date = min(as.Date(paste0(year_dep, "-", date_month))),
    last_date = max(as.Date(paste0(year_dep, "-", date_month))),
    period_type = "non-breeding",
    .groups = "drop"
  )

# Combine for reference
periods <- bind_rows(periods_breeding, periods_nonbreeding)

cat("\n=== BREEDING PERIODS ===\n")
knitr::kable(
  head(periods_breeding, 10),
  caption = "Sample of breeding temporal periods"
)

cat("\n=== NON-BREEDING PERIODS ===\n")
knitr::kable(
  head(periods_nonbreeding, 10),
  caption = "Sample of non-breeding temporal periods"
)
```

---

# Variable Configuration

## Variable Dictionary

Create a configuration table mapping variable names, NetCDF short names, and processing options:

```{r var-config}
vars_list <- data.frame(
  var = c(
    "CHLa", "CHLag", "NPP", "NPPg", "SST", "SSTg",
    "SSS", "SSSg", "MLD", "SSH", "Wind", "gIUV"
  ),
  var_short = c(
    "chl", "chl", "nppv", "nppv", "thetao", "thetao",
    "so", "so", "mlotst", "zos", "wind_speed", "uo"
  ),
  archivo_idx = c(1, 1, 1, 1, 3, 3, 3, 3, 2, 2, 5, 4),
  gradient = c(
    FALSE, TRUE, FALSE, TRUE, FALSE, TRUE,
    FALSE, TRUE, FALSE, FALSE, FALSE, FALSE
  ),
  description = c(
    "Chlorophyll-a concentration",
    "Chlorophyll-a spatial gradient",
    "Net Primary Productivity",
    "NPP spatial gradient",
    "Sea Surface Temperature",
    "SST spatial gradient",
    "Sea Surface Salinity",
    "SSS spatial gradient",
    "Mixed Layer Depth",
    "Sea Surface Height",
    "Wind Speed",
    "Wind stress curl (magnitude U,V)"
  )
)

knitr::kable(vars_list, caption = "Environmental variables configuration")
```

---

# Core Processing Function

## Habitat Variable Processing

This function handles the complete processing pipeline for each environmental variable:

```{r process-function}
process_habitat_vars <- function(v_idx, v_short, v_long, is_gradient,
                                track_period, template) {
  
  file_path <- files[v_idx]
  
  # ---------------------------------------------------------------
  # 1. TEMPORAL FILTERING
  # ---------------------------------------------------------------
  if (v_long == "gIUV") {
    # Wind stress curl: compute magnitude from U and V components
    u_raw <- rast(file_path, subds = "uo")
    v_raw <- rast(file_path, subds = "vo")
    
    # Extract time indices within track period
    idx <- which(
      time(u_raw) >= track_period$first_date &
      time(u_raw) <= track_period$last_date
    )
    
    if (length(idx) == 0) return(NULL)
    
    # Compute magnitude: sqrt(U^2 + V^2)
    r_period <- sqrt(u_raw[[idx]]^2 + v_raw[[idx]]^2)
  } else {
    # Standard variable extraction
    r_raw <- rast(file_path, subds = v_short)
    
    idx <- which(
      time(r_raw) >= track_period$first_date &
      time(r_raw) <= track_period$last_date
    )
    
    if (length(idx) == 0) return(NULL)
    
    r_period <- r_raw[[idx]]
  }
  
  # ---------------------------------------------------------------
  # 2. SPATIAL ADJUSTMENT
  # ---------------------------------------------------------------
  # Aggregate to template resolution if necessary
  if (res(r_period)[1] < res(template)[1]) {
    fact <- round(res(template)[1] / res(r_period)[1])
    r_period <- aggregate(r_period, fact = fact, fun = mean, na.rm = TRUE)
  }
  
  # Crop and resample to template extent
  r_period <- crop(r_period, ext(template), snap = "out")
  r_period <- resample(r_period, template, method = "near")
  
  # ---------------------------------------------------------------
  # 3. DEPTH COLLAPSE (for CHLa and NPP)
  # ---------------------------------------------------------------
  # These variables have multiple depth layers; average them
  if (v_idx == 1) {
    n_depths <- 12  # Number of depth levels
    n_days <- nlyr(r_period) / n_depths
    r_period <- tapp(
      r_period,
      index = rep(1:n_days, each = n_depths),
      fun = mean,
      na.rm = TRUE
    )
  }
  
  # ---------------------------------------------------------------
  # 4. GRADIENT COMPUTATION
  # ---------------------------------------------------------------
  # For gradient variables, compute spatial heterogeneity
  if (is_gradient) {
    r_list <- lapply(1:nlyr(r_period), function(i) {
      layer_i <- r_period[[i]]
      
      # Check if layer has data
      if (global(layer_i, "notNA")[1, 1] > 0) {
        grec::getGradients(layer_i)
      } else {
        # Return empty layer if no data
        setValues(template, NA)
      }
    })
    
    r_period <- rast(r_list)
  }
  
  # ---------------------------------------------------------------
  # 5. MASKING AND STATISTICS
  # ---------------------------------------------------------------
  # Apply template mask (ocean only)
  r_period <- mask(r_period, template, inverse = TRUE)
  
  # Compute mean and standard deviation across time
  m_i <- mean(r_period, na.rm = TRUE)
  s_i <- stdev(r_period, na.rm = TRUE)
  
  return(c(m_i, s_i))
}
```

---

# Main Processing Loop: Breeding Climatologies

## Execute Variable Processing for Breeding Season

Process all environmental variables for breeding trips:

```{r process-loop-breeding}
# Output directory for breeding
path_out_breeding <- "breeding_climatology"

if (!dir.exists(path_out_breeding)) {
  dir.create(path_out_breeding, recursive = TRUE)
}

cli::cli_h1("PROCESSING BREEDING SEASON CLIMATOLOGIES")

# Process each variable
for (i in 1:nrow(vars_list)) {
  
  v <- vars_list[i, ]
  cli::cli_h2("VARIABLE: {v$var}")
  cli::cli_alert("Description: {v$description}")
  
  # List to store individual results
  all_indiv_layers <- list()
  
  # Process each breeding trip
  for (id in periods_breeding$geo_trip) {
    cli::cli_alert_info("Computing for breeding trip: {id}")
    
    # Get period for current trip
    track <- periods_breeding %>% filter(geo_trip == id)
    
    # Process variable for this individual
    res_indiv <- process_habitat_vars(
      v_idx = v$archivo_idx,
      v_short = v$var_short,
      v_long = v$var,
      is_gradient = v$gradient,
      track_period = track,
      template = template_raster
    )
    
    if (!is.null(res_indiv)) {
      all_indiv_layers[[id]] <- res_indiv
    }
    
    gc()  # Memory cleanup
  }
  
  # ---------------------------------------------------------------
  # FINAL AGGREGATION FOR BREEDING
  # ---------------------------------------------------------------
  if (length(all_indiv_layers) > 0) {
    # Stack all individual results
    stack_final <- rast(all_indiv_layers)
    
    # Separate mean and sd layers
    idx_means <- seq(1, nlyr(stack_final), by = 2)
    idx_sds <- seq(2, nlyr(stack_final), by = 2)
    
    # 1. Global mean climatology across all trips
    final_mean <- mean(stack_final[[idx_means]], na.rm = TRUE)
    
    # 2. Mean temporal variability
    # (indicates how dynamic conditions are at each pixel)
    final_sd_temporal <- mean(stack_final[[idx_sds]], na.rm = TRUE)
    
    # Prepare output raster
    res_raster <- c(final_mean, final_sd_temporal)
    names(res_raster) <- c("mean_global", "sd_temporal_mean")
    
    # Save output
    output_file <- file.path(path_out_breeding, 
                             paste0(v$var,"_breeding_complete.tif"))
    writeRaster(res_raster, output_file, overwrite = TRUE)
    
    cli::cli_alert_success("Saved: {basename(output_file)}")
  } else {
    cli::cli_alert_warning("No valid data for {v$var} in breeding period")
  }
}

cli::cli_h2("BREEDING CLIMATOLOGIES PROCESSING COMPLETE!")
```

---

# Main Processing Loop: Non-Breeding Climatologies

## Execute Variable Processing for Non-Breeding Season

Process all environmental variables for non-breeding trips:

```{r process-loop-nonbreeding}
# Output directory for non-breeding
path_out_nonbreeding <- "non_breeding_climatology"

if (!dir.exists(path_out_nonbreeding)) {
  dir.create(path_out_nonbreeding, recursive = TRUE)
}

cli::cli_h1("PROCESSING NON-BREEDING SEASON CLIMATOLOGIES")

# Process each variable
for (i in 1:nrow(vars_list)) {
  
  v <- vars_list[i, ]
  cli::cli_h2("VARIABLE: {v$var}")
  cli::cli_alert("Description: {v$description}")
  
  # List to store individual results
  all_indiv_layers <- list()
  
  # Process each non-breeding trip
  for (id in periods_nonbreeding$geo_trip) {
    cli::cli_alert_info("Computing for non-breeding trip: {id}")
    
    # Get period for current trip
    track <- periods_nonbreeding %>% filter(geo_trip == id)
    
    # Process variable for this individual
    res_indiv <- process_habitat_vars(
      v_idx = v$archivo_idx,
      v_short = v$var_short,
      v_long = v$var,
      is_gradient = v$gradient,
      track_period = track,
      template = template_raster
    )
    
    if (!is.null(res_indiv)) {
      all_indiv_layers[[id]] <- res_indiv
    }
    
    gc()  # Memory cleanup
  }
  
  # ---------------------------------------------------------------
  # FINAL AGGREGATION FOR NON-BREEDING
  # ---------------------------------------------------------------
  if (length(all_indiv_layers) > 0) {
    # Stack all individual results
    stack_final <- rast(all_indiv_layers)
    
    # Separate mean and sd layers
    idx_means <- seq(1, nlyr(stack_final), by = 2)
    idx_sds <- seq(2, nlyr(stack_final), by = 2)
    
    # 1. Global mean climatology across all trips
    final_mean <- mean(stack_final[[idx_means]], na.rm = TRUE)
    
    # 2. Mean temporal variability
    # (indicates how dynamic conditions are at each pixel)
    final_sd_temporal <- mean(stack_final[[idx_sds]], na.rm = TRUE)
    
    # Prepare output raster
    res_raster <- c(final_mean, final_sd_temporal)
    names(res_raster) <- c("mean_global", "sd_temporal_mean")
    
    # Save output
    output_file <- file.path(path_out_nonbreeding, 
                             paste0(v$var, "_non_breeding_complete.tif"))
    writeRaster(res_raster, output_file, overwrite = TRUE)
    
    cli::cli_alert_success("Saved: {basename(output_file)}")
  } else {
    cli::cli_alert_warning("No valid data for {v$var} in non-breeding period")
  }
}

cli::cli_h2("NON-BREEDING CLIMATOLOGIES PROCESSING COMPLETE!")
```

---

# Output Validation and Assembly

## Verify Output Files

```{r validate-output}
# List generated breeding climatology files
breeding_files <- list.files(
  "breeding_climatology",
  full.names = TRUE
)

# List generated non-breeding climatology files
nonbreeding_files <- list.files(
  "non_breeding_climatology",
  full.names = TRUE
)

cat("=== BREEDING CLIMATOLOGIES ===\n")
cat("Generated", length(breeding_files), "files:\n")
print(basename(breeding_files))

cat("\n=== NON-BREEDING CLIMATOLOGIES ===\n")
cat("Generated", length(nonbreeding_files), "files:\n")
print(basename(nonbreeding_files))
```

## Stack Variables by Season

Create separate stacks for breeding and non-breeding predictors:

```{r stack-variables-by-season}
# ===== BREEDING STACK =====
stack_breeding <- rast(breeding_files)

cat("\n=== BREEDING PREDICTOR STACK ===\n")
cat("Layers:", nlyr(stack_breeding), "\n")
cat("Extent:", paste(ext(stack_breeding)), "\n")
cat("Resolution:", res(stack_breeding)[1], "degrees\n")

# ===== NON-BREEDING STACK =====
stack_nonbreeding <- rast(nonbreeding_files)

cat("\n=== NON-BREEDING PREDICTOR STACK ===\n")
cat("Layers:", nlyr(stack_nonbreeding), "\n")
cat("Extent:", paste(ext(stack_nonbreeding)), "\n")
cat("Resolution:", res(stack_nonbreeding)[1], "degrees\n")

# Verify geometric compatibility within each season
cli::cli_h2("Verifying geometric compatibility...")

tryCatch(
  compareGeom(stack_breeding),
  error = function(e) {
    cli::cli_alert_warning("Breeding geometry mismatch detected:")
    print(e)
  },
  success = function() {
    cli::cli_alert_success("Breeding stack geometry OK")
  }
)

tryCatch(
  compareGeom(stack_nonbreeding),
  error = function(e) {
    cli::cli_alert_warning("Non-breeding geometry mismatch detected:")
    print(e)
  },
  success = function() {
    cli::cli_alert_success("Non-breeding stack geometry OK")
  }
)
```

## Visualize Predictor Layers by Season

```{r visualize-stacks}
# Plot breeding climatology layers
cli::cli_h2("BREEDING CLIMATOLOGIES VISUALIZATION")
plot(stack_breeding, maxcell = 100000, main = "Breeding Season")

# Plot non-breeding climatology layers
cli::cli_h2("NON-BREEDING CLIMATOLOGIES VISUALIZATION")
plot(stack_nonbreeding, maxcell = 100000, main = "Non-Breeding Season")

# Summary statistics
cat("\n=== BREEDING STACK SUMMARY ===\n")
print(summary(stack_breeding))

cat("\n=== NON-BREEDING STACK SUMMARY ===\n")
print(summary(stack_nonbreeding))
```

# Session Information

```{r session-info}
sessionInfo()
```
