---
title: "Environmental Niche Models (ENMs)"
author: "Mario Gómez Álvarez"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---

# ENM Workflow

This script implements ENMs using BART (Bayesian Additive Regression Trees) at multiple hierarchical levels: Population, Biogeographic unit, Taxa, and Supra-taxa (Mediterranean Puffinus shearwaters).

---

## 1. Setup and Landmask Creation

### 1.1 Load Libraries

```{r setup, message=FALSE, warning=FALSE}
packages <- c(
  "rnaturalearth", "rworldmap", "rworldmap",
  "grid", "terra", "adehabitatHR", "sf", 
  "sp", "geosphere", "wordcloud", "gdistance",
  "rasterVis", "RColorBrewer",
  "purrr", "Hmisc", "embarcadero"
  )

invisible(lapply(packages, library, character.only = TRUE))
```

### 1.2 Create Landmask

```{r landmask}
# 1. Create world land vector
world <- ne_countries(scale = 50, returnclass = "sf")
world_vect <- vect(world)

# 2. Template raster
template <- rast(
  xmin = -15, xmax = 42,
  ymin = 28, ymax = 51,
  resolution = 0.25,
  crs = "EPSG:4326"
)

# 3. Effective mask design:
# Rasterize land with value 1
r_land <- rasterize(world_vect, template, field = 1)

# Create sea mask: 
# Where there is NO land (is.na), we put 1. Where there IS land, we put NA.
r_mask <- is.na(r_land) 
r_mask[r_mask == 0] <- NA # Convert 0s (land) to real NA
plot(r_mask, main = "Sea Mask (1 = Sea, NA = Land)")
```

---

## 2. Load and Process Covariates

### 2.1 Set Working Directory and Load Data

```{r load_data}
# Load breeding season covariates
predictors_original_breeding <- rast(list.files(
  'breeding_climatology',
  pattern = ".tif", 
  full.names = TRUE
))

# Assign layer names
names(predictors_original_breeding) <- c(
  "BATH", "CHLa_mean", "CHLa_sd", "CHLag_mean",
  "CHLag_sd", "gIUV_mean", "gIUV_sd",
  "MLD_mean", "MLD_sd", "NPP_mean",
  "NPP_sd", "NPPg_mean", "NPPg_sd", "SLOPE",
  "SSH_mean", "SSH_sd", "SSS_mean", "SSS_sd",
  "SSSg_mean", "SSSg_sd", "SST_mean", "SST_sd",
  "SSTg_mean", "SSTg_sd", "Wind_mean", "Wind_sd"
)

# Load non-breeding season covariates
predictors_original_nonbreeding <- rast(list.files(
  'nonbreeding_climatology',
  pattern = ".tif", 
  full.names = TRUE
))

# Assign layer names
names(predictors_original_nonbreeding) <- c(
  "BATH", "CHLa_mean", "CHLa_sd", "CHLag_mean",
  "CHLag_sd", "gIUV_mean", "gIUV_sd",
  "MLD_mean", "MLD_sd", "NPP_mean",
  "NPP_sd", "NPPg_mean", "NPPg_sd", "SLOPE",
  "SSH_mean", "SSH_sd", "SSS_mean", "SSS_sd",
  "SSSg_mean", "SSSg_sd", "SST_mean", "SST_sd",
  "SSTg_mean", "SSTg_sd", "Wind_mean", "Wind_sd"
)
```

### 2.2 Multicollinearity Analysis - Breeding Season

```{r multicollinearity_breeding}
# Create correlation matrix using Spearman similarity
v <- as.formula(paste("~", paste(names(predictors_original_breeding), collapse = "+")))
varclus_plot <- varclus(v, similarity = "spearman", data = predictors_original_breeding)
plot(varclus_plot, cex = 0.8, main = "Breeding Season - Variable Clustering")

# Extract similarity matrix
sim_matrix <- varclus_plot$sim

# Find redundant pairs (correlation > 0.8)
redundantes <- which(sim_matrix > 0.8 & sim_matrix < 1.0, arr.ind = TRUE)
print("Redundant variable pairs (breeding):")
print(redundantes)
```

### 2.3 Multicollinearity Analysis - Non-Breeding Season

```{r multicollinearity_nonbreeding}
# Create correlation matrix using Spearman similarity
v <- as.formula(paste("~", paste(names(predictors_original_nonbreeding), collapse = "+")))
varclus_plot <- varclus(v, similarity = "spearman", data = predictors_original_nonbreeding)
plot(varclus_plot, cex = 0.8, main = "Non-Breeding Season - Variable Clustering")

# Extract similarity matrix
sim_matrix <- varclus_plot$sim

# Find redundant pairs (correlation > 0.8)
redundant <- which(sim_matrix > 0.8 & sim_matrix < 1.0, arr.ind = TRUE)
print("Redundant variable pairs (non-breeding):")
print(redundant)
```

### 2.4 Select Final Variables and Scale Covariates

```{r select_and_scale_covariates}
# Define final selected variables
final_vars <- c(
  "NPP_mean", "NPPg_mean", "SST_mean", "SST_sd", "SSTg_mean", 
  "SSS_mean", "SSSg_mean", "MLD_mean", "SSH_mean", "SSH_sd", 
  "gIUV_mean", "Wind_mean", "Wind_sd", "BATH", "SLOPE"
)

# Create pattern for filtering
patern <- paste(final_vars, collapse = "|")

# Filter and prepare breeding covariates
predictors_original_breeding <- predictors_original_breeding[[grepl(patern, names(predictors_original_breeding))]]
predictors_original_breeding <- terra::mask(predictors_original_breeding, r_mask)
predictors_scaled_breeding <- scale(predictors_original_breeding)

# Filter and prepare non-breeding covariates
predictors_original_nonbreeding <- predictors_original_nonbreeding[[final_vars]]
predictors_original_nonbreeding <- terra::mask(predictors_original_nonbreeding, r_mask)
predictors_scaled_nonbreeding <- scale(predictors_original_nonbreeding)

cat("Final variables selected:\n")
print(final_vars)
cat("\nBreeding predictor stack:\n")
print(predictors_scaled_breeding)
cat("\nNon-breeding predictor stack:\n")
print(predictors_scaled_nonbreeding)
```

### 2.5 Load Presence Data

```{r load_presence_data}
# Load presence data at different hierarchical levels
Population.pr <- readRDS('outputs/pres_by_population_thinned.RDS')
Unit.pr <- readRDS('outputs/pres_by_unit_thinned.RDS')
Taxa.pr <- readRDS('outputs/pres_by_sp_thinned.RDS')
Suprataxa.pr <- readRDS('outputs/pres_by_suprataxa_thinned.RDS')

cat("Presence data loaded successfully\n")
```

---

## 3. Population Level Models

### 3.1 Prepare Dataset - Population Level

```{r prep_population_data}

all.cov_list <- list()

for (i in unique(Population.pr$period)) {
  
  Population.pr_period <- Population.pr[Population.pr$period == i, ]
  
  if (i == "breeding") {
    covs <- predictors_scaled_breeding
  } else {
    covs <- predictors_scaled_nonbreeding
  }
  
  all.cov_list[[i]] <- list()
  
  for (j in unique(Population.pr_period$Population)) {
    
    Population.pr_sub <- Population.pr_period %>% 
      filter(Population == j)
    
    # Convert to spatial points
    pts <- st_as_sf(Population.pr_sub,
                    coords = c("Longitude", "Latitude"),
                    crs = 4326)
    
    # Get cell IDs
    pts$cell_id <- cellFromXY(covs$BATH, st_coordinates(pts))
    
    # Remove duplicates within cells
    pts_unique <- pts %>%
      filter(!is.na(cell_id)) %>%
      group_by(cell_id) %>%
      slice(1) %>%
      ungroup()
    
    # Extract presence covariates
    pres.cov <- terra::extract(covs, vect(pts_unique))
    pres.cov$pres <- 1
    
    # Generate random absences
    masked_bath <- terra::mask(covs$BATH, r_mask)
    
    absences <- spatSample(
      masked_bath,
      size = nrow(pres.cov),
      method = "random",
      na.rm = TRUE,
      xy = TRUE
    )
    
    # Extract absence covariates
    abs.cov <- terra::extract(covs, absences[, 1:2])
    abs.cov$pres <- 0
    
    # Combine presence and absence data
    all.cov <- rbind(pres.cov, abs.cov)
    all.cov2 <- all.cov %>% drop_na()
    
    all.cov_list[[i]][[j]] <- all.cov2
  }
}

all.cov_list <- compact(all.cov_list)
cat("Population-level dataset prepared\n")
```

### 3.2 Run BART Models - Population Level

```{r run_population_models, cache=TRUE}

path.out.pop <- "outputs/Pop_ENMs"

if (!dir.exists(path.out.pop)) {
  dir.create(path.out.pop, recursive = TRUE)
}

Pop_models <- list()
Pop_stack <- stack()
pop.results <- data.frame(Model = character(), 
                          AUC = numeric(), 
                          stringsAsFactors = FALSE)

for (i in seq_along(all.cov_list)) {
  
  all_cov_period <- all.cov_list[[i]]
  
  for (j in seq_along(all_cov_period)) {
    
    pop <- names(all_cov_period)[j]
    if (i == 1) p <- "breeding" else p <- "non-breeding"
    
    if (p == "breeding") {
      covs <- predictors_scaled_breeding
    } else {
      covs <- predictors_scaled_nonbreeding
    }
    
    cat(paste("Running model:", pop, p, "\n"))
    
    all_cov_sub <- as.data.frame(all_cov_period[[j]])
    xvars <- names(covs)
    
    # Run BART model with automatic variable selection
    pop.sdm <- bart.step(
      x.data = all_cov_sub[, xvars],
      y.data = all_cov_sub[, 'pres'],
      full = TRUE,
      quiet = TRUE
    )
    
    # Calculate AUC
    fitobj <- pop.sdm$fit
    true.vector <- fitobj$data@y 
    pred <- prediction(colMeans(pnorm(pop.sdm$yhat.train)), true.vector)
    auc <- performance(pred, "auc")@y.values[[1]]
    
    pop.results <- rbind(pop.results, 
                         data.frame(Model = paste0(pop, '_', p), 
                                    AUC = auc))
    
    # Store model
    Pop_models[paste0(pop, "_", p)] <- list(pop.sdm)
    
    # Make predictions
    Pop_prediction <- predict(
      object = pop.sdm,
      x.layers = stack(covs),
      quantiles = c(0.5),
      splitby = 20,
      quiet = TRUE
    )
    
    # Stack predictions
    Pop_stack <- stack(Pop_stack, Pop_prediction[[1]])
  }
  
  # Save predictions
  writeRaster(Pop_stack,
              filename = paste0(path.out.pop, "/Population_pred_", p, ".tif"), 
              overwrite = TRUE)
  # Save models
  saveRDS(Pop_models, 
          file = paste0(path.out.pop, "/Population_models_", p, ".RDS"))
  
  Pop_stack <- stack()
  Pop_models <- list()
}

# Save results summary
write.csv(pop.results, 
          paste0(path.out.pop, "/Population_summary.csv"), 
          row.names = FALSE)

cat("Population models completed\n")
rm(Pop_models, pop.results, pop.sdm, Pop_stack)
```

---

## 4. Biogeographic Group Level Models

### 4.1 Prepare Dataset - Group Level

```{r prep_group_data}
all.cov_list <- list()

for (i in unique(Unit.pr$period)) {
  
Unit.pr_period <- Unit.pr[Unit.pr$period == i, ]
  
  if (i == "breeding") {
    covs <- predictors_scaled_breeding
  } else {
    covs <- predictors_scaled_nonbreeding
  }
  
  all.cov_list[[i]] <- list()
  
  for (j in unique(Unit.pr_period$Unit)) {
        
    Unit.pr_sub <- Unit.pr_period %>% 
      filter(Unit == j)
    
    # Convert to spatial points
    pts <- st_as_sf(Unit.pr_sub,
                    coords = c("Longitude", "Latitude"),
                    crs = 4326)
    
    # Get cell IDs
    pts$cell_id <- cellFromXY(covs$BATH, st_coordinates(pts))
    
    # Remove duplicates within cells
    pts_unique <- pts %>%
      filter(!is.na(cell_id)) %>%
      group_by(cell_id) %>%
      slice(1) %>%
      ungroup()
    
    # Extract presence covariates
    pres.cov <- terra::extract(covs, vect(pts_unique))
    pres.cov$pres <- 1
    
    # Generate random absences
    masked_bath <- terra::mask(covs$BATH, r_mask)
    
    absences <- spatSample(
      masked_bath,
      size = nrow(pres.cov),
      method = "random",
      na.rm = TRUE,
      xy = TRUE
    )
    
    # Extract absence covariates
    abs.cov <- terra::extract(covs, absences[, 1:2])
    abs.cov$pres <- 0
    
    # Combine presence and absence data
    all.cov <- rbind(pres.cov, abs.cov)
    all.cov2 <- all.cov %>% drop_na()
    
    all.cov_list[[i]][[j]] <- all.cov2
  }
}

all.cov_list <- compact(all.cov_list)
cat("Group-level dataset prepared\n")
```

### 4.2 Run BART Models - Unit Level

```{r run_group_models, cache=TRUE}

path.out.unit <- "outputs/Unit_ENMs"

if (!dir.exists(path.out.unit)) {
  dir.create(path.out.unit, recursive = TRUE)
}

Unit_stack <- stack()
Unit_models <- list()
un.results <- data.frame(Model = character(), 
                         AUC = numeric(), 
                         stringsAsFactors = FALSE)

for (i in seq_along(all.cov_list)) {
  
  all_cov_period <- all.cov_list[[i]]
  
  for (j in seq_along(all_cov_period)) {
    
    un <- names(all_cov_period)[j]
    if (i == 1) p <- "breeding" else p <- "non-breeding"
    
    if (p == "breeding") {
      covs <- predictors_scaled_breeding
    } else {
      covs <- predictors_scaled_nonbreeding
    }
    
    cat(paste("Running model:", un, p, "\n"))
    
    all_cov_sub <- as.data.frame(all_cov_period[[j]])
    xvars <- names(covs)
    
    # Run BART model
    un.sdm <- bart.step(
      x.data = all_cov_sub[, xvars],
      y.data = all_cov_sub[, 'pres'],
      full = TRUE,
      quiet = TRUE
    )
    
    # Calculate AUC
    fitobj <- un.sdm$fit
    true.vector <- fitobj$data@y 
    pred <- prediction(colMeans(pnorm(un.sdm$yhat.train)), true.vector)
    auc <- performance(pred, "auc")@y.values[[1]]
    
    un.results <- rbind(un.results, data.frame(Model = paste0(un, '_', p), AUC = auc))
    
    # Store model
    Unit_models[paste0(un, "_", p)] <- list(un.sdm)
    
    # Make predictions
    Unit_prediction <- predict(
      object = un.sdm,
      x.layers = stack(covs),
      quantiles = c(0.5),
      splitby = 20,
      quiet = TRUE
    )
    
    # Stack predictions
    Unit_stack <- stack(Unit_stack, Unit_prediction[[1]])
  }
  
  # Save predictions
  writeRaster(Unit_stack, 
              filename = paste0(path.out.unit, "/Unit_pred_", p, ".tif"), 
              overwrite = TRUE)
  
  # Save models
  saveRDS(Unit_models, 
          file = paste0(path.out.unit, "/Unit_pred_models_", p, ".RDS"))
  
  Unit_stack <- stack()
  Unit_models <- list()
}

# Save results summary
write.csv(un.results, 
          paste0(path.out.unit, "/Unit_summary.csv"), 
          row.names = FALSE)

cat("Group models completed\n")
rm(Unit_models, un.results, un.sdm, Unit_stack)
```

---

## 5. Taxa (Species) Level Models

### 5.1 Prepare Dataset - Taxa Level

```{r prep_taxa_data}
all.cov_list <- list()

for (i in unique(Taxa.pr$period)) {
  
  Taxa.pr_period <- Taxa.pr[Taxa.pr$period == i, ]
  
  if (i == "breeding") {
    covs <- predictors_scaled_breeding
  } else {
    covs <- predictors_scaled_nonbreeding
  }
  
  all.cov_list[[i]] <- list()
  
  for (j in unique(Taxa.pr_period$Sp)) {
    
    Taxa.pr_sub <- Taxa.pr_period %>% 
      filter(Sp == j)
    
    # Convert to spatial points
    pts <- st_as_sf(Taxa.pr_sub,
                    coords = c("Longitude", "Latitude"),
                    crs = 4326)
    
    # Get cell IDs
    pts$cell_id <- cellFromXY(covs$BATH, st_coordinates(pts))
    
    # Remove duplicates within cells
    pts_unique <- pts %>%
      filter(!is.na(cell_id)) %>%
      group_by(cell_id) %>%
      slice(1) %>%
      ungroup()
    
    # Extract presence covariates
    pres.cov <- terra::extract(covs, vect(pts_unique))
    pres.cov$pres <- 1
    
    # Generate random absences
    masked_bath <- terra::mask(covs$BATH, r_mask)
    
    absences <- spatSample(
      masked_bath,
      size = nrow(pres.cov),
      method = "random",
      na.rm = TRUE,
      xy = TRUE
    )
    
    # Extract absence covariates
    abs.cov <- terra::extract(covs, absences[, 1:2])
    abs.cov$pres <- 0
    
    # Combine presence and absence data
    all.cov <- rbind(pres.cov, abs.cov)
    all.cov2 <- all.cov %>% drop_na()
    
    all.cov_list[[i]][[j]] <- all.cov2
  }
}

all.cov_list <- compact(all.cov_list)
cat("Taxa-level dataset prepared\n")
```

### 5.2 Run BART Models - Taxa Level

```{r run_taxa_models, cache=TRUE}

path.out.tx <- "outputs/Taxa_ENMs"

if (!dir.exists(path.out.tx)) {
  dir.create(path.out.tx, recursive = TRUE)
}

Tx_stack <- stack()
Tx_models <- list()
tx.results <- data.frame(Model = character(), 
                         AUC = numeric(), 
                         stringsAsFactors = FALSE)

for (i in seq_along(all.cov_list)) {
  
  all_cov_period <- all.cov_list[[i]]
  
  for (j in seq_along(all_cov_period)) {
    
    tx <- names(all_cov_period)[j]
    if (i == 1) p <- "breeding" else p <- "non-breeding"
    
    if (p == "breeding") {
      covs <- predictors_scaled_breeding
    } else {
      covs <- predictors_scaled_nonbreeding
    }
    
    cat(paste("Running model:", tx, p, "\n"))
    
    all_cov_sub <- as.data.frame(all_cov_period[[j]])
    xvars <- names(covs)
    
    # Run BART model
    tx.sdm <- bart.step(
      x.data = all_cov_sub[, xvars],
      y.data = all_cov_sub[, 'pres'],
      full = TRUE,
      quiet = TRUE
    )
    
    # Calculate AUC
    fitobj <- tx.sdm$fit
    true.vector <- fitobj$data@y 
    pred <- prediction(colMeans(pnorm(tx.sdm$yhat.train)), true.vector)
    auc <- performance(pred, "auc")@y.values[[1]]
    
    tx.results <- rbind(tx.results, data.frame(Model = paste0(tx, '_', p), AUC = auc))
    
    # Store model
    Tx_models[paste0(tx, "_", p)] <- list(tx.sdm)
    
    # Make predictions
    Sp_prediction <- predict(
      object = tx.sdm,
      x.layers = stack(covs),
      quantiles = c(0.5),
      splitby = 20,
      quiet = TRUE
    )
    
    # Stack predictions
    Tx_stack <- stack(Tx_stack, Sp_prediction[[1]])
  }
  
  # Save predictions
  writeRaster(Tx_stack, 
              filename = paste0(path.out.tx, "Taxa_pred_", p, ".tif"), 
              overwrite = TRUE)
  
  # Save models
  saveRDS(Tx_models, 
          file = paste0(path.out.tx, "Taxa_pred_models_", p, ".RDS"))
  
  Tx_stack <- stack()
  Tx_models <- list()
}

# Save results summary
write.csv(tx.results, 
          paste0(path.out.tx, "Taxa_summary.csv"), 
          row.names = FALSE)

cat("Taxa models completed\n")
rm(Tx_models, tx.results, tx.sdm, Tx_stack)
```

---

## 6. Supra-Taxa Level Models

### 6.1 Prepare Dataset - Supra-Taxa Level

```{r prep_suprataxa_data}
all.cov_list <- list()

for (i in unique(Suprataxa.pr$period)) {
  
  Suprataxa.pr_period <- Suprataxa.pr[Suprataxa.pr$period == i, ]
  
  if (i == "breeding") {
    covs <- predictors_scaled_breeding
  } else {
    covs <- predictors_scaled_nonbreeding
  }
  
  # Convert to spatial points
  pts <- st_as_sf(Suprataxa.pr_period,
                  coords = c("Longitude", "Latitude"),
                  crs = 4326)
  
  # Get cell IDs
  pts$cell_id <- cellFromXY(covs$BATH, st_coordinates(pts))
  
  # Remove duplicates within cells
  pts_unique <- pts %>%
    filter(!is.na(cell_id)) %>%
    group_by(cell_id) %>%
    slice(1) %>%
    ungroup()
  
  # Extract presence covariates
  pres.cov <- terra::extract(covs, vect(pts_unique))
  pres.cov$pres <- 1
  
  # Generate random absences
  masked_bath <- terra::mask(covs$BATH, r_mask)
  
  absences <- spatSample(
    masked_bath,
    size = nrow(pres.cov),
    method = "random",
    na.rm = TRUE,
    xy = TRUE
  )
  
  # Extract absence covariates
  abs.cov <- terra::extract(covs, absences[, 1:2])
  abs.cov$pres <- 0
  
  # Combine presence and absence data
  all.cov <- rbind(pres.cov, abs.cov)
  all.cov2 <- all.cov %>% drop_na()
  
  all.cov_list[[i]] <- all.cov2
}

cat("Supra-taxa dataset prepared\n")
```

### 6.2 Run BART Models - Supra-Taxa Level

```{r run_suprataxa_models, cache=TRUE}

path.out.st <- "outputs/Supra-taxa_ENMs"

if (!dir.exists(path.out.st)) {
  dir.create(path.out.st, recursive = TRUE)
}

Supra_models <- list()
supra.results <- data.frame(Model = character(), 
                            AUC = numeric(), 
                            stringsAsFactors = FALSE)

for (i in seq_along(all.cov_list)) {
  
  all_cov_period <- as.data.frame(all.cov_list[[i]])
  
  if (i == 1) p <- "breeding" else p <- "non-breeding"
  
  if (p == "breeding") {
    covs <- predictors_scaled_breeding
  } else {
    covs <- predictors_scaled_nonbreeding
  }
  
  sp <- "Supra-taxa"
  cat(paste("Running model:", sp, p, "\n"))
  
  xvars <- names(covs)
  
  # Run BART model
  sp.sdm <- bart.step(
    x.data = all_cov_period[, xvars],
    y.data = all_cov_period[, 'pres'],
    full = TRUE,
    quiet = FALSE
  )
  
  # Calculate AUC
  fitobj <- sp.sdm$fit
  true.vector <- fitobj$data@y 
  pred <- prediction(colMeans(pnorm(sp.sdm$yhat.train)), true.vector)
  auc <- performance(pred, "auc")@y.values[[1]]
  
  supra.results <- rbind(supra.results, data.frame(Model = paste0(sp, '_', p), AUC = auc))
  
  # Store model
  Supra_models[paste0(sp, "_", p)] <- list(sp.sdm)
}

# Save models
saveRDS(Supra_models, 
        file = paste0(path.out.st, "/Suprataxa_models.RDS"))

# Save results summary
write.csv(supra.results, 
          paste0(path.out.st, "/Suprataxa_summary.csv"), 
          row.names = FALSE)

cat("Supra-taxa models completed\n")
```

# Session Information

```{r session-info}
sessionInfo()
```
