---
title: "Partial Response Curves from ENM Models"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---

# Partial Response Curves from Ensemble Niche Models

This script extracts and visualizes partial response curves from the fitted ENM models, showing how species responses vary with environmental variables across biogeographic units and seasons.

---

## 1. Setup and Libraries

### 1.1 Load Required Libraries

```{r setup, message=FALSE, warning=FALSE}
library(terra)
library(embarcadero)
library(dplyr)
library(stringr)
library(tidyr)
library(ggplot2)
library(viridisLite)

```

---

## 2. Load Models and Covariates

### 2.1 Load Fitted Unit Models

```{r load_models}
# List Unit model files
Unit_models_list <- list.files(
  "outputs/Unit_ENMs", 
  pattern = ".RDS", 
  full.names = TRUE
)

# Load and flatten models
Unit_models <- lapply(Unit_models_list, readRDS)
Unit_models <- do.call(c, Unit_models)

cat("Unit models loaded:", length(Unit_models), "\n")
```

### 2.2 Load Breeding Covariates for Scaling Reference

```{r load_covariates}
# Load breeding season covariates for unscaling predictions
predictors_original_breeding <- rast(
  list.files("breeding_climatology",
    pattern = ".tif", 
    full.names = TRUE
  )
)

names(predictors_original_breeding) <- c(
  "BATH", "CHLa_mean", "CHLa_sd", "CHLag_mean",
  "CHLag_sd", "gIUV_mean", "gIUV_sd",
  "MLD_mean", "MLD_sd", "NPP_mean",
  "NPP_sd", "NPPg_mean", "NPPg_sd", "SLOPE",
  "SSH_mean", "SSH_sd", "SSS_mean", "SSS_sd",
  "SSSg_mean", "SSSg_sd", "SST_mean", "SST_sd",
  "SSTg_mean", "SSTg_sd", "Wind_mean", "Wind_sd"
)

cat("Breeding covariates loaded for scaling reference\n")
```

---

## 3. Select and Organize Models

### 3.1 Define Variables by Season

```{r define_variables}
# Define variables to extract partial effects for each season
vars <- list(
  breeding = c("NPP_mean", "SST_mean", "SST_sd", "SSS_mean", "SSH_mean", "SSH_sd", "BATH"),
  nonbreeding = c("SST_mean", "SST_sd", "SSS_mean", "SSH_mean", "SSH_sd")
)

cat("Variables defined for partial effects analysis\n")
```

### 3.2 Extract Specific Models for Each Biogeographic Unit and Season

```{r select_models}
# Extract models for each biogeographic unit and season
wmb <- Filter(Negate(is.null), Unit_models[grepl("WesternMost_breeding", names(Unit_models))][1])
wmnb <- Filter(Negate(is.null), Unit_models[grepl("WesternMost_non-breeding", names(Unit_models))][1])
cwb <- Filter(Negate(is.null), Unit_models[grepl("Central-West_breeding", names(Unit_models))][1])
cwnb <- Filter(Negate(is.null), Unit_models[grepl("Central-West_non-breeding", names(Unit_models))][1])
ceb <- Filter(Negate(is.null), Unit_models[grepl("Central-East_breeding", names(Unit_models))][1])
cenb <- Filter(Negate(is.null), Unit_models[grepl("Central-East_non-breeding", names(Unit_models))][1])
emb <- Filter(Negate(is.null), Unit_models[grepl("EasternMost_breeding", names(Unit_models))][1])
emnb <- Filter(Negate(is.null), Unit_models[grepl("EasternMost_non-breeding", names(Unit_models))][1])

# Organize models in list
unit_models <- list(
  WesternMost_breeding = wmb,
  WesternMost_nonbreeding = wmnb,
  CentralWest_breeding = cwb,
  CentralWest_nonbreeding = cwnb,
  CentralEast_breeding = ceb,
  CentralEast_nonbreeding = cenb,
  EasternMost_breeding = emb,
  EasternMost_nonbreeding = emnb
)

cat("Models selected for all biogeographic units and seasons\n")
```

---

## 4. Calculate Partial Effects

### 4.1 Extract Partial Response Curves

```{r calculate_partials, message=TRUE}
# Calculate partial effects for all models and variables
partials_all <- list()

for (unit_name in names(unit_models)) {
  
  message("Running partials for: ", unit_name)
  
  mod <- unit_models[[unit_name]][[1]]
  
  # Detect period
  period <- ifelse(grepl("nonbreeding", unit_name), "nonbreeding", "breeding")
  vars_here <- vars[[period]]
  
  df_unit <- data.frame()
  
  # Get variables actually present in the model
  vars_in_model <- row.names(varimp(mod))
  
  for (v in vars_here) {
    
    # Skip variables not in model
    if (!(v %in% vars_in_model)) {
      message("  Skipping ", v, " (not in model)")
      next
    }
    
    message("  Variable: ", v)
    
    # Calculate partial effect
    part <- partial(
      mod,
      x.vars = v,
      trace = FALSE,
      ci = TRUE,
      equal = TRUE,
      smooth = 10,
      panels = FALSE
    )
    
    # Extract data
    df_v <- part[[1]]$data %>%
      dplyr::mutate(
        variable = v,
        unit = unit_name,
        period = period
      )
    
    df_unit <- dplyr::bind_rows(df_unit, df_v)
  }
  
  partials_all[[unit_name]] <- df_unit
}

# Combine all partial effects
partials_df <- dplyr::bind_rows(partials_all)

cat("\nPartial effects calculated\n")
cat("Variables in partial effects:", unique(partials_df$variable), "\n")
```

### 4.2 Unscale Partial Effects Back to Original Scale

```{r unscale_partials}
# Calculate scaling attributes for each variable
scaled_npp <- scale(predictors_original_breeding$NPP_mean[])
scaled_ssh_sd <- scale(predictors_original_breeding$SSH_sd[])
scaled_ssh <- scale(predictors_original_breeding$SSH_mean[])
scaled_sss <- scale(predictors_original_breeding$SSS_mean[])
scaled_sst <- scale(predictors_original_breeding$SST_mean[])
scaled_sst_sd <- scale(predictors_original_breeding$SST_sd[])
scaled_bath <- scale(predictors_original_breeding$BATH[])

# Unscale partial effects back to original units
partials_df <- partials_df %>% 
  mutate(
    x2 = case_when(
      variable == "NPP_mean" ~ x * attr(scaled_npp, 'scaled:scale') + attr(scaled_npp, 'scaled:center'),
      variable == "SSH_mean" ~ x * attr(scaled_ssh, 'scaled:scale') + attr(scaled_ssh, 'scaled:center'),
      variable == "SSH_sd" ~ x * attr(scaled_ssh_sd, 'scaled:scale') + attr(scaled_ssh_sd, 'scaled:center'),
      variable == "SSS_mean" ~ x * attr(scaled_sss, 'scaled:scale') + attr(scaled_sss, 'scaled:center'),
      variable == "SST_mean" ~ x * attr(scaled_sst, 'scaled:scale') + attr(scaled_sst, 'scaled:center'),
      variable == "SST_sd" ~ x * attr(scaled_sst_sd, 'scaled:scale') + attr(scaled_sst_sd, 'scaled:center'),
      variable == "BATH" ~ x * attr(scaled_bath, 'scaled:scale') + attr(scaled_bath, 'scaled:center'),
      TRUE ~ NA_real_
    ),
    unit = factor(
      sub("_breeding$|_nonbreeding$", "", unit),
      levels = c("WesternMost", "CentralWest", "CentralEast", "EasternMost")
    ),
    period = as.factor(ifelse(period == "breeding", "Breeding", "Non-breeding"))
  ) %>%
  filter(variable != "BATH")

# Save partial effects to file
write.csv(partials_df, 
          file = "outputs/partials.csv",
          row.names = FALSE)

cat("Partial effects unscaled and saved\n")
```

---

## 5. Visualization

### 5.1 Create Partial Response Curves Plot

```{r plot_partials, fig.width=12, fig.height=8}
# Define colors for biogeographic units
colors <- mako(12)[c(1, 5, 7, 10)]
colors <- setNames(colors, levels(partials_df$unit))

# Create partial response plot
p_partial <- ggplot(partials_df, aes(x = x2, color = unit, Unit = unit)) + 
  geom_ribbon(aes(ymin = q05, ymax = q95, fill = unit), color = NA, alpha = 0.2) + 
  geom_line(aes(y = med), size = 0.8) + 
  theme_bw() + 
  labs(
    x = "Value", 
    y = "Response",
    color = "Biogeographical unit",
    fill = "Biogeographical unit"
  ) +
  scale_colour_manual(values = colors) +
  scale_fill_manual(values = colors) +
  facet_grid(period ~ variable, scales = "free_x") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom",
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 11),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    strip.text = element_text(size = 12, face = "bold")
  )

# Display plot
print(p_partial)

# Save to file
ggsave("outputs/partial_effects.png", 
  p_partial, 
  width = 12, height = 8, dpi = 600
)

cat("Partial response curves plot saved")
```

---

## Summary

This workflow extracts and visualizes partial response curves from the fitted ENM models at the biogeographic Unit level:

### Key Features:

- **Partial Response Curves**: Show how species suitability responds to each environmental variable
- **Confidence Intervals**: 95% confidence intervals (shaded areas) indicate uncertainty
- **Biogeographic Comparison**: Four biogeographic units shown in different colors
- **Seasonal Separation**: Breeding and non-breeding seasons analyzed separately
- **Variable-Specific Predictors**: Different variables are relevant for each season

### Output:

- `partials.csv` - Raw partial effect data (scaled values)
- `partial_effects.png` - Visualization of partial response curves

### Variables Analyzed:

**Breeding Season:** NPP_mean, SST_mean, SST_sd, SSS_mean, SSH_mean, SSH_sd

**Non-Breeding Season:** SST_mean, SST_sd, SSS_mean, SSH_mean, SSH_sd

---

*Last updated: `r Sys.Date()`*
