---
title: "Earth Mover's Distance (EMD) and Geographic Correlations"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---

# EMD and Geographic Distance Correlations

This script calculates Earth Mover's Distance (EMD) between population habitat projections and correlates it with geographic distance across biogeographic units and seasons.

---

## 1. Setup and Libraries

### 1.1 Load Required Libraries

```{r setup, message=FALSE, warning=FALSE}
library(raster)
library(sp)
library(sf)
library(terra)
library(ggplot2)
library(readxl)
library(embarcadero)
library(emdist)
library(gdistance)
library(tidyr)
library(dplyr)
library(ggrepel)
library(viridisLite)
library(broom)
library(lme4)

```

---

## 2. Load and Prepare Data

### 2.1 Load Population Predictions

```{r load_predictions}
# Load breeding season predictions
predictions_b <- stack("outputs/Pop_ENMs/Population_pred_breeding.tif")
names(predictions_b) <- c("France", "Mallorca", "Sardinia", "Minorca",
                          "Ibiza", "Malta", "Greece", "Croatia")

# Aggregate to lower resolution for faster computation
predictions_b_low <- aggregate(predictions_b, fact = 2, fun = mean)

# Load non-breeding season predictions
predictions_nb <- stack("outputs/Pop_ENMs/Population_pred_non-breeding.tif")
names(predictions_nb) <- names(predictions_b)
predictions_nb_low <- aggregate(predictions_nb, fact = 2, fun = mean)

cat("Predictions loaded and aggregated\n")
```

### 2.2 Load and Prepare Colony Data

```{r load_colonies}
# Load colony data
colonies <- read.csv('data/colonies_coords.csv')

# Define population groups
cols <- c("Ibiza", "Mallorca", "Minorca", "France", "Sardinia", "Malta", "Croatia", "Greece")
PUFMAU_cols <- c("Ibiza", "Mallorca")
PUFYEL_cols <- c("Minorca", "France", "Sardinia", "Malta", "Croatia", "Greece")
g1_cols <- c("Ibiza", "Mallorca", "Minorca")
g2_cols <- c("France", "Sardinia")
g3_cols <- c("Malta", "Croatia")
g4_cols <- c("Greece")

cat("Colony data prepared\n")
```

### 2.3 Create All Pairwise Comparisons

```{r create_pairs}
# Create all pairs of populations
col_pairs <- combn(seq_along(cols), 2)
n_pairs <- ncol(col_pairs)  # 28 pairs

cat("Total population pairs:", n_pairs, "\n")
```

---

## 3. Calculate Earth Mover's Distance (EMD)

### 3.1 Compute EMD for All Population Pairs

```{r calculate_emd, message=TRUE}
# Initialize results data frame
emd_monthly_pop <- data.frame(
  period = character(),
  pop1 = character(),
  pop2 = character(),
  emd = numeric()
)

# Calculate EMD for each season and population pair
for (i in c("Breeding", "Non-breeding")) {
  
  if (i == "Breeding") {
    rast_list <- predictions_b_low[[cols]]
  } else {
    rast_list <- predictions_nb_low[[cols]]
  }
  
  for (k in 1:n_pairs) {
    
    j1 <- col_pairs[1, k]
    j2 <- col_pairs[2, k]
    
    p1 <- rast_list[[j1]]
    p2 <- rast_list[[j2]]
    
    m1 <- as.matrix(p1)
    m2 <- as.matrix(p2)
    
    # Keep only cells where both have data
    valid_mask <- !is.na(m1) & !is.na(m2)
    m1[!valid_mask] <- 0
    m2[!valid_mask] <- 0
    
    # Normalize to probability distributions
    m1 <- m1 / sum(m1)
    m2 <- m2 / sum(m2)
    
    # Calculate EMD
    emd_val <- emd2d(m1, m2)
    
    emd_monthly_pop <- rbind(
      emd_monthly_pop,
      data.frame(
        period = i,
        pop1 = cols[j1],
        pop2 = cols[j2],
        emd = emd_val
      )
    )
    
    gc()
  }
  
  cat(i, "done\n")
}

cat("EMD calculations completed\n")
cat("Total EMD values:", nrow(emd_monthly_pop), "\n")
```

### 3.2 Classify Population Pairs by Biogeographic Unit

```{r classify_pairs}
# Classify pairs as within or between biogeographic units
emd_monthly_pop2 <- emd_monthly_pop %>%
  mutate(
    group = case_when(
      pop1 %in% g1_cols & pop2 %in% g1_cols ~ "WesternMost",
      pop1 %in% g1_cols & !pop2 %in% g1_cols ~ "Between-unit",
      !pop1 %in% g1_cols & pop2 %in% g1_cols ~ "Between-unit",
      pop1 %in% g2_cols & pop2 %in% g2_cols ~ "Central-West",
      pop1 %in% g2_cols & !pop2 %in% g2_cols ~ "Between-unit",
      !pop1 %in% g2_cols & pop2 %in% g2_cols ~ "Between-unit",
      pop1 %in% g3_cols & pop2 %in% g3_cols ~ "Central-East",
      pop1 %in% g3_cols & !pop2 %in% g3_cols ~ "Between-unit",
      !pop1 %in% g3_cols & pop2 %in% g3_cols ~ "Between-unit",
      pop1 %in% g4_cols & pop2 %in% g4_cols ~ "EasternMost",
      pop1 %in% g4_cols & !pop2 %in% g4_cols ~ "Between-unit",
      !pop1 %in% g4_cols & pop2 %in% g4_cols ~ "Between-unit",
      TRUE ~ NA_character_
    )
  )

cat("Population pairs classified by biogeographic unit\n")
```

---

## 4. Calculate Geographic Distances

### 4.1 Prepare Ocean Raster for Cost-Distance Calculation

```{r prepare_ocean_raster}
# Define Mediterranean extent
xmin <- -20
xmax <- 50
ymin <- 20
ymax <- 55
resolution <- 0.05

# Create empty raster template
ocean.r <- raster(
  xmn = xmin, xmx = xmax,
  ymn = ymin, ymx = ymax,
  res = resolution,
  crs = CRS("EPSG:4326")
)

# Get land polygons from rnaturalearth
land_sf <- ne_countries(scale = "medium", returnclass = "sf") %>%
  st_union()

# Convert sf to sp for rasterization 
land_sp <- as_Spatial(land_sf)

# Rasterize land polygons: land areas will have a value, NA elsewhere
ocean.r <- rasterize(land_sp, ocean.r, field = 1)

# Convert: land = 0, ocean = 1
ocean.r[!is.na(ocean.r)] <- 0  # Land cells set to 0
ocean.r[is.na(ocean.r)] <- 1    # Ocean cells set to 1

cat("Ocean raster created:\n")
```

### 4.2 Relocate Colonies to Ocean if Necessary

```{r relocate_colonies}

# Convert to spatial
sites_sf <- st_as_sf(colonies, coords = c("Lon", "Lat"), crs = 4326)
sites <- as.data.frame(st_coordinates(sites_sf))

# Check which sites are on land
sites.to.relocate <- raster::extract(ocean.r, sites) == 0
sites.to.relocate.xy <- sites[sites.to.relocate, ]

# Relocate land-based sites to nearest ocean cell
if (nrow(sites.to.relocate.xy) > 0) {
  
  cat("\n.....................................................................\n\n")
  
  if (nrow(sites.to.relocate.xy) == 1) {
    cat(" 1 site to be relocated.\n")
  } 
  if (nrow(sites.to.relocate.xy) > 1) {
    cat(" ", nrow(sites.to.relocate.xy), "sites to be relocated.\n")
  }
  
  ocean.r.sites <- as.data.frame(ocean.r, xy = TRUE)[, 1:2]
  ocean.r.sites <- cbind(ocean.r.sites, raster::extract(ocean.r, ocean.r.sites))
  ocean.r.sites <- ocean.r.sites[ocean.r.sites[, 3] == 1, 1:2]
  
  for (i in 1:nrow(sites.to.relocate.xy)) {
    near.cells <- as.data.frame(
      sort(sp::spDistsN1(as.matrix(ocean.r.sites), 
                         as.matrix(sites.to.relocate.xy[i, 1:2]), 
                         longlat = TRUE), 
           decreasing = FALSE, 
           index.return = TRUE)
    )
    sites[which(sites.to.relocate)[i], ] <- ocean.r.sites[near.cells[1, 2], ]
  }
}

colonies$Lon <- sites$X
colonies$Lat <- sites$Y

cat("Colonies relocated to ocean\n")
```

### 4.3 Calculate Minimum Marine Distance Between Colonies

```{r calculate_marine_distance}
# Create cost surface (ocean = movement cost 1, land = no movement)
cost.surface <- ocean.r
projection(cost.surface) <- CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0")

# Calculate transition matrix
raster_tr <- gdistance::transition(cost.surface, mean, directions = 8)
raster_tr_corrected <- gdistance::geoCorrection(raster_tr, type = "c", multpl = FALSE)

# Calculate cost distances between all colony pairs
raster_cosDist <- gdistance::costDistance(
  raster_tr_corrected,
  as.matrix(colonies[, c("Lon", "Lat")]),
  as.matrix(colonies[, c("Lon", "Lat")])
)

# Convert to kilometers
distance.marine <- as.matrix(raster_cosDist) / 1000

cat("Marine distances calculated\n")
```

### 4.4 Merge EMD and Geographic Distance Data

```{r merge_emd_distance}
# Convert distance matrix to long format
dist_df <- as.data.frame(distance.marine)
dist_df$col1 <- colonies$pop

dist_long <- dist_df %>%
  pivot_longer(
    cols = -col1,
    names_to = "col2_idx",
    values_to = "distance"
  ) %>%
  mutate(
    col2_idx = as.integer(gsub("\\D", "", col2_idx)),
    col2 = colonies$pop[col2_idx]
  ) %>%
  dplyr::select(col1, col2, distance)

# Merge with EMD data
emd_monthly_pop_final <- emd_monthly_pop2 %>%
  mutate(period = as.factor(period)) %>%
  left_join(
    dist_long, 
    by = c("pop1" = "col1", "pop2" = "col2")
  )

cat("EMD and geographic distance data merged\n")
cat("Final dataset:", nrow(emd_monthly_pop_final), "pairwise comparisons\n")
```

---

## 5. Visualization

### 5.1 Prepare Data for Plotting

```{r prepare_plot_data}
# Expand data to long format for plotting
emd_plot <- emd_monthly_pop_final %>%
  mutate(id = row_number()) %>%
  tidyr::pivot_longer(
    cols = c(pop1, pop2),
    names_to = "which",
    values_to = "pop"
  )

emd_plot$pop <- factor(emd_plot$pop, levels = unique(c(emd_monthly_pop_final$pop1, emd_monthly_pop_final$pop2)))
emd_plot$group <- factor(emd_plot$group, levels = c("WesternMost", "Central-West", "Central-East", "EasternMost", "Between-unit"))

# Prepare unique pair data
emd_plot_unique <- emd_plot %>%
  dplyr::select(id, period, emd, distance, group, which, pop) %>%
  pivot_wider(names_from = which, values_from = pop) %>%
  mutate(label = paste(pop1, pop2, sep = "–"))

# Add taxa classification
emd_plot_unique <- emd_plot_unique %>%
  mutate(
    Taxa = case_when(
      pop1 %in% PUFMAU_cols & pop2 %in% PUFMAU_cols ~ "PUFMAU",
      pop1 %in% PUFMAU_cols & !pop2 %in% PUFMAU_cols ~ "Between-taxa",
      !pop1 %in% PUFMAU_cols & pop2 %in% PUFMAU_cols ~ "Between-taxa",
      pop1 %in% PUFYEL_cols & pop2 %in% PUFYEL_cols ~ "PUFYEL",
      TRUE ~ NA_character_
    ),
    Taxa = factor(Taxa, levels = c("PUFMAU", "PUFYEL", "Between-taxa"))
  )

cat("Plot data prepared\n")
```

### 5.2 Plot EMD vs Geographic Distance with Population Labels

```{r plot_emd_labeled}
# Define colors for biogeographic units
colors <- c(mako(4))
colors <- setNames(colors, levels(factor(emd_plot_unique$group)))
colors["Between-unit"] <- "gray60"

# Create plot
p2.1 <- ggplot(emd_plot_unique, aes(x = distance, y = emd, colour = group)) + 
  facet_wrap(~ period, labeller = labeller(
    period = c("Breeding" = "Breeding", "Non-breeding" = "Non-breeding")
  )) +
  geom_point(aes(shape = Taxa), alpha = 0.7, size = 3) +
  geom_smooth(aes(group = 1), method = "lm", se = TRUE, colour = "black") +
  ggrepel::geom_text_repel(
    data = subset(emd_plot_unique),
    aes(label = label),
    size = 3,
    colour = "black",
    show.legend = FALSE,
    max.overlaps = 20
  ) +
  scale_colour_manual(values = colors) +
  labs(
    x = "Minimum marine distance (km)",
    y = "Earth Mover's Distance",
    colour = "Biogeographical unit",
    shape = "Taxa"
  ) +
  theme_bw() +
  theme(
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 11),
    legend.position = "top",
    legend.box = "vertical",
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 11),
    strip.text = element_text(size = 12, face = "bold")
  ) +
  xlim(c(-100, 2400))

print(p2.1)

ggsave("outputs/plot_emd_dist.png", 
       p2.1, width = 10, height = 6, dpi = 600)

cat("Labeled EMD plot saved\n")
```
---

## 6. Statistical Analysis

### 6.1 Classify Comparisons as Within or Between Units

```{r classify_comparisons}
# Create function to get group membership
group_list <- list(
  g1 = g1_cols,
  g2 = g2_cols,
  g3 = g3_cols,
  g4 = g4_cols
)

get_group <- function(colony) {
  for (g in names(group_list)) {
    if (colony %in% group_list[[g]]) return(g)
  }
  return(NA)
}

# Add group assignments and comparison type
emd_monthly_pop_final <- emd_monthly_pop_final %>%
  mutate(
    group1 = sapply(pop1, get_group),
    group2 = sapply(pop2, get_group),
    comparison_type = ifelse(group1 == group2, "within", "between")
  )

cat("Comparisons classified\n")
```

### 6.2 Linear Models by Season

```{r lm_breeding}
# Linear regression for breeding season
model_breeding <- lm(emd ~ distance,
                     data = emd_monthly_pop_final %>% filter(period == "Breeding"))

cat("\n===== Breeding Season Model =====\n")
print(summary(model_breeding))
```

```{r lm_nonbreeding}
# Linear regression for non-breeding season
model_nonbreeding <- lm(emd ~ distance,
                        data = emd_monthly_pop_final %>% filter(period == "Non-breeding"))

cat("\n===== Non-Breeding Season Model =====\n")
print(summary(model_nonbreeding))
```


---

## Summary

This workflow analyzes the relationship between Earth Mover's Distance (EMD) and geographic distance:

### Key Steps:

1. **EMD Calculation** - Compares habitat suitability distributions between population pairs
2. **Geographic Distance** - Calculates minimum marine distance between colonies
3. **Correlation Analysis** - Examines EMD-distance relationships by biogeographic unit and season
4. **Statistical Models** - Linear models quantify significant relationships

### Key Findings:

- **Breeding vs Non-breeding**: Different EMD-distance relationships by season
- **Biogeographic Units**: Within-unit comparisons vs. between-unit comparisons
- **Taxa Differentiation**: PUFMAU vs. PUFYEL population divergence patterns

### Output Files:

- `plot_emd_dist.png` - Labeled population pairs with EMD-distance relationships

---

*Last updated: `r Sys.Date()`*

# Session Information

```{r session-info}
sessionInfo()
```
